### 2013-10-21 paper.tex
ENH:  More edits.

%% This is file `elsarticle-template-1-num.tex',
%%
%% Copyright 2009 Elsevier Ltd
%%
%% This file is part of the 'Elsarticle Bundle'.
%% ---------------------------------------------
%%
%% It may be distributed under the conditions of the LaTeX Project Public
%% License, either version 1.2 of this license or (at your option) any
%% later version.  The latest version of this license is in
%%    http://www.latex-project.org/lppl.txt
%% and version 1.2 or later is part of all distributions of LaTeX
%% version 1999/12/01 or later.
%%
%% The list of all files belonging to the 'Elsarticle Bundle' is
%% given in the file `manifest.txt'.
%%
%% Template article for Elsevier's document class `elsarticle'
%% with numbered style bibliographic references
%%
%% $Id: elsarticle-template-1-num.tex 149 2009-10-08 05:01:15Z rishi $
%% $URL: http://lenova.river-valley.com/svn/elsbst/trunk/elsarticle-template-1-num.tex $
%%

%\documentclass[preprint,authoryear,review,12pt]{elsarticle}
\documentclass[final,5p,times,twocolumn]{elsarticle}

%% Use the option review to obtain double line spacing
%% \documentclass[preprint,review,12pt]{elsarticle}

%% Use the options 1p,two column; 3p; 3p,twocolumn; 5p; or 5p,twocolumn
%% for a journal layout:
%% \documentclass[final,1p,times]{elsarticle}
%% \documentclass[final,1p,times,twocolumn]{elsarticle}
%% \documentclass[final,3p,times]{elsarticle}
%% \documentclass[final,3p,times,twocolumn]{elsarticle}
%% \documentclass[final,5p,times]{elsarticle}
%% \documentclass[final,5p,times,twocolumn]{elsarticle}

\usepackage{color}
\usepackage{multirow,booktabs,ctable,array}
\usepackage{lscape}
\usepackage{amsmath}
\usepackage{lineno}
\usepackage{ulem}
\usepackage{setspace}
\usepackage{listings}
\usepackage{float}
\usepackage{listings}
\usepackage{color,colortbl}
\usepackage{rccol}
%\usepackage[table]{xcolor}

    \definecolor{listcomment}{rgb}{0.0,0.5,0.0}
    \definecolor{listkeyword}{rgb}{0.0,0.0,0.5}
    \definecolor{listnumbers}{gray}{0.65}
    \definecolor{listlightgray}{gray}{0.955}
    \definecolor{listwhite}{gray}{1.0}
    \definecolor{lightcyan}{rgb}{0.88,1,1}

\newcommand{\lstsetcpplong}
{
\lstset{frame = tb,
        framerule = 0.25pt,
        float,
        fontadjust,
        backgroundcolor={\color{listlightgray}},
        basicstyle = {\ttfamily\scriptsize},
        keywordstyle = {\ttfamily\color{listkeyword}\textbf},
        identifierstyle = {\ttfamily},
        commentstyle = {\ttfamily\color{listcomment}\textit},
        stringstyle = {\ttfamily},
        showstringspaces = false,
        showtabs = false,
        numbers = none,
        numbersep = 6pt,
        numberstyle={\ttfamily\color{listnumbers}},
        tabsize = 2,
        language=,
        floatplacement=!h,
        caption={\small \baselineskip 12pt DiReCT long command line menu which is invoked using the `{\ttfamily {-}{-}help}' option.  The short command line menu is obtained by typing `{\ttfamily {-}h}'},
        captionpos=b,
        label=listing:long
        }
}

\floatstyle{plain}
\newfloat{command}{thp}{lop}
\floatname{command}{Command}

%\usepackage[nomarkers,notablist]{endfloat}

%% if you use PostScript figures in your article
%% use the graphics package for simple commands
%% \usepackage{graphics}
%% or use the graphicx package for more complicated commands
%% \usepackage{graphicx}
%% or use the epsfig package if you prefer to use the old commands
%% \usepackage{epsfig}

%% The amssymb package provides various useful mathematical symbols
\usepackage{amssymb}
%% The amsthm package provides extended theorem environments
% \usepackage{amsthm}

 \usepackage{makecell}

%% The lineno packages adds line numbers. Start line numbering with
%% \begin{linenumbers}, end it with \end{linenumbers}. Or switch it on
%% for the whole article with \linenumbers after \end{frontmatter}.
%% \usepackage{lineno}

%% natbib.sty is loaded by default. However, natbib options can be
%% provided with \biboptions{...} command. Following options are
%% valid:

%%   round  -  round parentheses are used (default)
%%   square -  square brackets are used   [option]
%%   curly  -  curly braces are used      {option}
%%   angle  -  angle brackets are used    <option>
%%   semicolon  -  multiple citations separated by semi-colon
%%   colon  - same as semicolon, an earlier confusion
%%   comma  -  separated by comma
%%   numbers-  selects numerical citations
%%   super  -  numerical citations as superscripts
%%   sort   -  sorts multiple citations according to order in ref. list
%%   sort&compress   -  like sort, but also compresses numerical citations
%%   compress - compresses without sorting
%%
%% \biboptions{comma,round}

% \biboptions{}

\providecommand{\OO}[1]{\operatorname{O}\bigl(#1\bigr)}

\graphicspath{
             {./Figures/}
             }

\long\def\symbolfootnote[#1]#2{\begingroup%
\def\thefootnote{\fnsymbol{footnote}}\footnote[#1]{#2}\endgroup}

\journal{NeuroImage}

\begin{document}

\begin{frontmatter}

\title{Multivariate Neuroanalysis with ANTsR: Application to Supervised Brain Tumor Segmentation using Random Forests}

\author[label1]{Nicholas J.~Tustison
  \fnref{label0}}
  \fntext[label0]{\scriptsize Corresponding author:  PO Box 801339, Charlottesville, VA 22908; T:  434-924-7730; email address:  ntustison@virginia.edu.
  }
\author[label2]{K.~L.~Shrinidhi}
\author[label2]{Jeffrey T.~Duda}
\author[label1]{Christopher R.~Durst}
\author[label1]{Max Wintermark}
\author[label1]{James C.~Gee}
\author[label1]{Murray C.~Grossman}
\author[label2]{Brian B.~Avants}
\address[label1]{Department of Radiology and Medical Imaging, University of Virginia, Charlottesville, VA}
\address[label2]{Penn Image Computing and Science Laboratory, University of Pennsylvania,
                Philadelphia, PA}

%\maketitle

\linenumbers

\begin{abstract} 
Improvements in image acquisition and increasingly sophisticated statistical techniques 
for neuroanalysis have spurred the development of advanced computational frameworks for studying the
brain---many of which
have been made publicly available.  One such popular toolkit is the open source Advanced
Normalization Tools (ANTs) which contains a suite of well-vetted processing tools for
core data transformation tasks such as image registration, image segmentation, 
inhomogeneity field correction, and template building.  However, despite the numerous
solutions afforded by such packages, neuroscience (and data analysis in general) requires
robust, statistical tools for making predictive inferences with respect to given hypotheses
and for abstracting quantitative information to various graphical representations.  
In this paper, we describe the
coupling of ANTs with the widely-used R language---an open source environment for 
statistical processing and data visualization---which we denote as ANTsR.  One of the
many benefits from such an integration is the accessibility to advanced statistical 
and machine learning techniques.  To showcase the flexibility and power of ANTsR, we 
apply the combination of one well-known random forest R package  and ANTs processing 
tools to the difficult problem of brain tumor segmentation from multi-modality
MRI.  This algorithmic framework was the top-performing entry in the MICCAI 2013 Multimodal 
Brain Tumor Segmentation (BRATS) challenge and, to our knowledge, is the only entry for
both challenge years (2012, 2013) which has been made publicly available as open source.
\end{abstract}

\begin{keyword}
advanced normalization tools \sep BRATS \sep brain tumor segmentation \sep R project \sep random forests
%% keywords here, in the form: keyword \sep keyword
\end{keyword}

\end{frontmatter}
%
%
\newpage

%% MSC codes here, in the form: \MSC code \sep code
%% or \MSC[2008] code \sep code (2000 is the default)

%%
%% Start line numbering here if you want
%%
% \linenumbers

%% The Appendices part is started with the command \appendix;
%% appendix sections are then done as normal sections
%% \appendix

%% \section{}
%% \label{}

%% References
%%
%% Following citation commands can be used in the body text:
%% Usage of \cite is as follows:
%%   \citep{key}          ==>>  [#]
%%   \cite[chap. 2]{key} ==>>  [#, chap. 2]
%%   \citet{key}         ==>>  Author [#]

%% main text

%\input{intro}
%
%\input{methods}

\section{Introduction}

ANTs (Advanced Normalization Tools) originated with the open source availing
of state-of-the-art registration algorithms for neuroimage analysis
\cite{avants2008a} built upon the mature and well-vetted Insight Toolkit (ITK)
of the National Institutes of Health.  Since then, ANTs has grown to include 
several algorithmic solutions necessary
for robust medical image analysis including bias correction \cite{tustison2010}, 
$n$-tissue multivariate segmentation \cite{avants2011}, template construction \cite{avants2010}, 
and cortical thickness estimation \cite{das2009} (many of which have been
introduced into ITK partially in an attempted leveraging of Linus's Law---``Given enough eyeballs, all bugs are shallow'').
However, in the evolution of the toolkit, it became clear (as neuroimaging
research certainly falls under the popular rubric of ``big data analysis'')
that robust statistical machinery was lacking for proper inferences regarding
the data produced by the various ANTs tools.
ANTsR was developed
specifically to provide an interface between a 
powerful neuroimaging toolkit for producing reliable imaging data 
transformations and the R project%
\footnote{
http://www.r-project.org
}
for statistical computing and visualization thus providing a complete
set of tools for multivariate neuroimage analysis.

In addition to describing ANTsR basics and how it can be generally
used in multivariate neuroimaging studies, we showcase its use in
the particularly difficult application of performing supervised brain
segmentation using random forests.  Although random forests have
been proposed previously in the literature for, among other things, 
supervised brain segmentation (e.g. \cite{geremia2011,zikic2012}), 
we demonstrate that ANTsR significantly facilitates the construction 
of a fully functional, parallelizable workflow for such a task and 
that the performance of our proposed framework was proven extremely 
competitive in the recent MICCAI 2013 Multimodal Brain Tumor Segmentation
Challenge (see Figure \ref{fig:award}).%
\footnote{
http://martinos.org/qtim/miccai2013/
}
\begin{figure}
  \centering
  \includegraphics[width=80mm]{Figures/award.pdf}
  \caption{First place certificate from the MICCAI 2013 BRATS challenge.}
  \label{fig:award}
\end{figure}

It is interesting to note that our participation in the 2013 challenge
was the result of observing the 2012
challenge, wanting to take advantage of the same methods as the winners,
and finding that no teams had availed their software in any form.  As a 
corrective \citep{ince2012}, and very much in line with our philosophy 
regarding open science, we have made this framework available as open 
source (even releasing the code before the actual 2013 contest).%
\footnote{
https://github.com/ntustison/BRATS2013
}
To our knowledge, we are the only competitors out of the 20 or so teams 
that participated in either years' challenges to do this.

\section{Materials and Methods}

\subsection{ANTsR:  An ANTs/R Interface}

\subsubsection{Overview}

The quantitative complexity of neuroimaging research necessitates 
commensurable numerical analysis capabilities.  Similarly, concomitant
with the era of ``big data'' (and neuroimaging certainly qualifies
\citep{vanhorn2013}) are new visualization changes 
\citep{childs2013,kehrer2013}.
In response, various software packages have been developed to
integrate tools specific to neuroimaging research with numerical
and visualization software packages.
The well-known neuroimaging standard SPM%
\footnote{
http://www.fil.ion.ucl.ac.uk/spm/
}
is built as an add-on to the commercial computing and visualization environment
Matlab.  Open source neuroimaging packages, such as NIPY (neuroimaging in Python),%
\footnote{
http://nipy.org
} 
rely on other open source packages for numerical/statistical analysis.  NIPY,
for example, uses other open source python packages such as NumPy.%
\footnote{
http://www.numpy.org
}

Careful consideration of available options
led to the adoption of R to complement ANTs capabilities which resulted in the
ANTsR package.
The R project for statistical computing,%
\footnote{
www.r-project.org
}
 or more compactly
`R', is an environment for statistical computation
and data visualization.  R's open source code base
and add-on packages coupled with its rapidly growing 
community of developers and users has caused wide-scale
adoption within both academia and industry.

\subsubsection{Installation}

The ANTsR package is publicly available on the github project hosting service.%
\footnote{
https://github.com/stnava/ANTsR
}
Prior to installation of ANTsR, several external R packages
need to be installed including: \verb#Rcpp#, \verb#signal#, \verb#timeSeries#, 
\verb#mFilter#, \verb#doParallel#, \verb#robust#, \verb#magic#, \verb#knitr#, \verb#pixmap#, 
\verb#rgl#, \verb#misc3d# which is easily facilitated by the 
\verb#install.packages()# mechanism.  Additionally, in order
to perform the supervised brain segmentation as described 
in later sections, one would need to also install 
\verb#randomForest#, \verb#snowfall#, \verb#rlecuyer#,
and \verb#ggplot2#.

In addition to R and relevant packages, CMake needs to also be
installed.  CMake%
\footnote{
http://www.cmake.org/
}
is an open source tool for the management and building of 
large-scale software projects.  It is used
to coordinate the downloading of external packages,
such as the Insight Toolkit (ITK)%
\footnote{
http://www.itk.org/
}
and ANTs.  Further instructions for download and
installation can be found on the ANTsR github website.

\subsection{Supervised Brain Tumor Segmentation}

Given a set of training data consisting of 
labeled (csf, gray matter, white matter, edema, and tumor),
multimodal brain image data, 
supervised brain segmentation entails the generation of a 
statistical model from such training data which can
then be extended to testing data, i.e. unlabeled 
brain data.  In subsequent sections, we detail the generic
{\it random forest} modeling framework which takes as input
a set of {\it feature images} (also described in a later section)
and labels for each training subject 
and outputs a statistical model.  This model, in turn, can then
be used to provide a probabilistic estimate of the labels in an
unlabeled subject.

One of the core extensions that we provide in this work is
the use of concatenated random forests for improved probabilistic estimation
of the labels.
As we demonstrate in the Results section, the set of feature
images employed work sufficiently well for good performance
on the training data (which exceeds that of what has been
currently published in the literature).  However, we discovered
that further improvements could be gained by using the probabilistic 
label estimates as input to enhance preprocessing before a 
second round of feature image generation including 
modality-specific, prior-based $n$-tissue segmentation and 
random forest model creation.

\subsubsection{Random Forests}

Several machine learning concepts were integrated to create 
the random forests framework first articulated in its entirety by Breiman
et al. \cite{breiman2001} for performing classification/regression.  
Although decision trees had been previously explored in the literature, 
it was the success of ``boosting''-style machine learning 
techniques, such as AdaBoost \cite{schapire1990,freund1997}, which influenced 
the aggregation of such decision trees into ``forests'' 
with randomized node optimization for improved
classification/regression performance \cite{ho1995,amit1997}.
The final element of bootstrap aggregating or ``bagging'' (i.e.
random sampling of the training data) was
introduced by Breiman \cite{breiman1996} to achieve improved
accuracy.

Early adoption \cite{viola2005} and success in the
computer vision community
has led to a recent surge within the medical image analysis
community of using random forests for handling complex 
classification/regression tasks including
normal brain segmentation \cite{yi2009},
MS lesion segmentation \cite{geremia2011}, 
multimodal brain tumor segmentation
\cite{zikic2012,geremia2012}, brain extraction \cite{iglesias2010}, 
anatomy detection in computed tomography \cite{criminisi2013}, and
segmentation of echocardiographic images \cite{verhoek2011}.
A thorough introduction for those interested in delving deeper 
into the more theoretical aspects of random forests can be found
in \cite{criminisi2011}.

One of the principal advantages of R is the extensive community of
developers  who have contributed on the order of thousands of packages 
extending R's capabilities beyond its core functionality.
Most relevant for the work described in this paper
is the \verb#randomForest# package developed from Breiman's original
Fortran code by Liaw and Wiener (described briefly in \cite{liaw2002}).

\subsubsection{Multi-Modal Feature Image Preprocessing}

Although several studies have pointed out the importance of
intensity normalization and bias correction, our experience 
with the training data illustrated a degradation in performance
when one or both steps (using \cite{nyul2000} and N4 \cite{tustison2010},
respectively) were performed due to the presence of the tumor/edema complex.

As a corrective, for the first stage we simply windowed the image intensity
for all images to be between the quantiles $[0.01,0.99]$ and
subsequently rescaled to $[0,1]$.  From these ``corrected'' images,
the first set of feature images were derived.  For the training cohort, these
data were used to create the random forest regression model for the first
stage.  During the second stage, the probabilistic estimates
of the white matter and gray matter labels were used to generate a
``pure tissue weight mask'' to estimate the bias field 
using N4 (although the resulting bias field estimation was used
to correct the image within the entire cerebral mask).  Formally, this 
involved generation of a probabilistic map defined as:
\begin{align}
  P_{pure\,\,tissue}(\mathbf{x}) = \sum_{i=1}^N P_i(\mathbf{x}) \prod_{j=1, j \neq i}^N \left( 1 - P_j(\mathbf{x}) \right)
\end{align}
where $N$ is the set of user-selected tissue labels (in our
case $N=2$ consisting of the gray and white matter probability
maps).

Both rescaling and weighted bias correction were applied to produce
the ``corrected images'' for the second stage resulting in
modified features images for the second stage.  Note that we
perform a similar iterative scenario for normal brain 
segmentation \cite{avants2011} (encapsulated in the ANTs script 
\verb#antsAtroposN4.sh#).

\subsubsection{Multi-Modal Feature Image Generation}

Key to any supervised regression or classification protocol are the 
selected features for training and subsequent testing.  Based on previous
work and our own experience, we selected the following feature images
to showcase the supervised segmentation strategy developed in this work.

\begin{figure*}
  \centering
  \includegraphics[width=180mm]{Figures/featuresImages.pdf}
  \caption{Feature images from the BRATS\_HG0004 data set.}
\end{figure*}

\begin{itemize}
  \item Per modality (FLAIR, T1, T1C, T2)
    \begin{itemize}
      \item First-order neighborhood statistical images:
            mean, variance, skewness, and entropy. 
            Neighborhood radius = 3.
    \item GMM-based posteriors: CSF, gray matter, white matter, edema, and tumor
    \item GMM connected component geometry features:  volume, volume to 
          surface area ratio, eccentricity, elongation
    \item Template-based:  symmetric template difference and contralateral difference.
          Gaussian smoothing ($\sigma = 4$mm).
    \end{itemize}
  \item Not modality specific
    \begin{itemize}
    \item Normalized Euclidean distance
    \item Log Jacobian  
    \item T1C,T1 difference image
    \end{itemize}
\end{itemize}

For each modality, we create four first-order statistical feature images,
five Gaussian mixture model (GMM)-based posterior probability feature images,
four geometry features generated from the GMM posterior probability images
based on connected components, and two difference images using symmetric template
construction for a total of 4 modalities $\times$ (4 + 5 + 4 + 2) feature images per modality $=$ 60 total
feature images.  We employ two additional images consisting of the 
Euclidean distance image \cite{maurer2003} created from the skull-stripped 
binary mask rescaled to the range $[0,1]$ and the log
Jacobian image derived from the spatial normalization of the symmetric multivariate template and individual subject images.  Given the intensity corrected images
the corresponding multivariate template images, and a brain mask for each subject
creation of all feature images is performed using the script \verb#createFeatureImages.sh#.

Prior  cluster centers for specific tissue types learned from training data \cite{reynolds2009} are used in the GMM to create multiple feature images.  
Given $M$ tissue types (e.g. CSF, gray matter,
white matter, edema, and tumor), a GMM formulates the 
probability distribution at each voxel, $\mathbf{x}$, as the
sum of Gaussian components, $\mathcal{N}(\mathbf{x}|\mu,\sigma)$, i.e.
\begin{align}
p\left(\mathbf{x}|\mu_m,\sigma_m,\lambda_m\right) = \sum_{i=1}^M \lambda_m \mathcal{N}(\mathbf{x}|\mu_m,\sigma_m)
\end{align}
where $\sum_{m=1}^M \lambda_m = 1$.  One popular method for 
determining the parameters of the GMM is maximum likelihood 
estimation which can performed using the Atropos segmentation 
tool \cite{avants2011}.  In contrast to previous generative
modeling approaches for multi-modal tumor segmentation 
(e.g. \cite{prastawa2003,zikic2012}), we do not use multivariate 
Gaussians to specify tissue probabilities but rather incorporate each
univariate probability map into the feature vector of the training
data.  As pointed out in \cite{menze2010}, multivariate modeling
might obscure the distinct biological information provided by each 
modality.  Instead, we let the random forest construction 
process determine the optimal combination of such multivariate
information.
Additionally, maximum posterior labeling from the GMM processing
is used to determine the connected components for each label.  
Geometric features (assigned voxel-wise) include the physical volumes 
of each connected component including the volume to surface area ratio, 
the elongation, and eccentricity.

In order to better characterize deviations from normal
multi-modal brain shape and appearance, several features were derived 
using population-specific multivariate template 
construction. A recent neuroimaging reproducibility study
by Landman et al. resulted in an open data cohort of 21
normal individuals, each imaged twice, comprising several
modalities including arterial spin labeling, 
fluid attenuated inversion recovery (FLAIR),
diffusion tensor imaging, functional imaging, T1, and T2 
\cite{landman2011}.

Given $K$ image modality types for $N$ subjects,  
${\mathbf I} = \{I_1,I_2,\ldots, I_K\}$, multivariate 
template construction iterates between optimizing the set 
of diffeomorphic transforms between the subjects and the 
template, 
$\left\{\left(\phi_1,\phi_1^{-1}\right),\ldots,\left(\phi_N,\phi_N^{-1}\right)\right\}$ 
and constructing the 
optimal multivariate template appearance 
$\mathbf{J}=\{J_1,J_2,\ldots, J_K\}$ to minimize the
following cost function:
\begin{align}
  \sum_{n=1}^N 
        \Bigg[ D &\left( \psi(\mathbf{x}),\phi_1^n(\mathbf{x},1)\right) \\ \nonumber 
        +& \sum_{k=1}^K \lambda_k \Pi_k \left(I_k^n\left(\phi_n(\mathbf{x},0.5)\right),J_k\left(\phi^{-1}_n(\mathbf{x},0.5)\right)\right)\Bigg]
\end{align}
where $D$ is the diffeomorphic shape distance,
\begin{align}
D\left( \phi( \mathbf{x},0),\phi( \mathbf{x},1)\right) = \int_0^1 \| \nu(\mathbf{x},t)\|_L dt
\end{align}
dependent on the choice of linear operator, $L$, and $\nu$
is the velocity field
\begin{align}
\nu\left( \phi(\mathbf{x},t) \right) = \frac{d\phi(\mathbf{x},t)}{dt},\,\,\, \phi(\mathbf{x},0) = \mathbf{x}.
\end{align}
Each pairwise registration employing the similarity metric $\Pi_k$ can 
be assigned a relative weighting, $\lambda_k$, to weight a particular
modality's influence in the construction process.  Further theoretical
details can be found in \cite{avants2008,avants2010}.
In terms of implementation, this algorithm is 
encapsulated in the script \verb#antsMultivariateTemplateConstruction.sh#,
available in the ANTs repository, which permits parallel processing on
an individual workstation or on a large computational cluster.

\begin{figure}
  \centering
  \begin{tabular}{cc}
    \includegraphics[width=40mm]{Figures/S_templateFA_140.png} &
    \includegraphics[width=40mm]{Figures/S_templateFLAIR_140.png} \\
    (a) & (b) \\
    \includegraphics[width=40mm]{Figures/S_templateT1_140.png} &
    \includegraphics[width=40mm]{Figures/S_templateT2_140.png} \\
    (c) & (d) 
  \end{tabular}
  \caption{Multivariate symmetric template created from the Kirby 
           21 data described in \cite{landman2011}.  Shown are the
           (a) fractional anisotropy (FA), (b) FLAIR, (c) MPRAGE, 
           and (d) T2 template components.
          }
  \label{fig:symmetrictemplates}
\end{figure}

\subsubsection{\underline{Bra}in \underline{T}umor \underline{S}egmentation Challenge Data}

The Brain tumor segmentation
Associated with the 2012 International Conference on Medical Image Computing and Computer Assisted
Intervention (MICCAI),%
\footnote{
http://www.miccai2012.org
}
the

\subsubsection{MS lesion segmentation challenge Challenge Data}

\section{Results}

\begin{table*}
\caption{Dice scores from the MICCAI 2012 BRATs Study}
\begin{center}
\begin{tabular*}{0.975\textwidth}{@{\extracolsep{\fill} } c c c c c c c c c}
\toprule
{} & \multicolumn{2}{c}{High-grade (real)} & \multicolumn{2}{c}{Low-grade (real)} & \multicolumn{2}{c}{High-grade (simulated)} & \multicolumn{2}{c}{Low-grade (simulated)}\\
{\bf Method} & Edema & Tumor & Edema & Tumor & Edema & Tumor & Edema & Tumor\\
\midrule
\cite{zikic2012} & {$0.70 \pm 0.09$} & {$0.71 \pm 0.24$} & {$0.44 \pm 0.18$} & {$0.62 \pm 0.27$} & {$0.65 \pm 0.27$} & {$0.90 \pm 0.05$} & {$0.55 \pm 0.23$} & {$0.71 \pm 0.20$} \\
\cite{bauer2012} & {$0.61 \pm 0.15$} & {$0.62 \pm 0.27$} & {$0.35 \pm 0.18$} & {$0.49 \pm 0.26$} & {$0.68 \pm 0.26$} & {$0.90 \pm 0.06$} & {$0.57 \pm 0.24$} & {$0.74 \pm 0.10$} \\
ANTsR & {$0.65 \pm 0.15$} & {$0.66 \pm 0.28$} & {$0.49 \pm 0.16$} & {$0.65 \pm 0.21$} & {$0.68 \pm 0.25$} & {$0.91 \pm 0.08$} & {$0.61 \pm 0.25$} & {$0.84 \pm 0.09$} \\
w/ Atropos & {$0.68 \pm 0.15$} & {$0.67 \pm 0.30$} & {$0.50 \pm 0.15$} & {$0.67 \pm 0.23$} & {$0.74 \pm 0.26$} & {$0.92 \pm 0.09$} & {$0.65 \pm 0.26$} & {$0.84\pm 0.08$} \\
\bottomrule
\end{tabular*}
\end{center}
\end{table*}

\section{Discussion and Conclusions}

%% References with bibTeX database:

\section*{Acknowledgments}

\section*{References}

\bibliographystyle{elsarticle-harv}
\bibliography{references}

%% Authors are advised to submit their bibtex database files. They are
%% requested to list a bibtex style file in the manuscript if they do
%% not want to use model1-num-names.bst.

%% References without bibTeX database:

% \begin{thebibliography}{00}

%% \bibitem must have the following form:
%%   \bibitem{key}...
%%

% \bibitem{}

% \end{thebibliography}

\end{document}

%%
%% End of file `elsarticle-template-1-num.tex'.

### 2013-11-03 paper.tex
ENH:  So tired.

@@
 Improvements in image acquisition and increasingly sophisticated statistical techniques 
-for neuroanalysis have spurred the development of advanced computational frameworks for studying the
+for neuroanalysis have motivated the development of advanced computational software 
+packages for studying the
 brain---many of which
@@
 core data transformation tasks such as image registration, image segmentation, 
-inhomogeneity field correction, and template building.  However, despite the numerous
+inhomogeneity field correction, and template building.  Despite the numerous
 solutions afforded by such packages, neuroscience (and data analysis in general) requires
 robust, statistical tools for making predictive inferences with respect to given hypotheses
-and for abstracting quantitative information to various graphical representations.  
+and for rendering quantitative information in clarifying and insightful ways.  
 In this paper, we describe the
@@
 statistical processing and data visualization---which we denote as ANTsR.  One of the
-many benefits from such an integration is the accessibility to advanced statistical 
-and machine learning techniques.  To showcase the flexibility and power of ANTsR, we 
-apply the combination of one well-known random forest R package  and ANTs processing 
-tools to the difficult problem of brain tumor segmentation from multi-modality
-MRI.  This algorithmic framework was the top-performing entry in the MICCAI 2013 Multimodal 
+many benefits from such integration is the accessibility to advanced statistical 
+and machine learning techniques as well as state-of-the-art visualization possibilities.  
+To showcase the flexibility and utility of ANTsR, we 
+apply it to the difficult problem of brain tumor segmentation from multi-modality
+MRI.  The reported algorithmic framework was the top-performing entry in the MICCAI 2013 Multimodal 
 Brain Tumor Segmentation (BRATS) challenge and, to our knowledge, is the only entry for
@@
 ANTs (Advanced Normalization Tools) originated with the open source availing
-of state-of-the-art registration algorithms for neuroimage analysis
-\cite{avants2008a} built upon the mature and well-vetted Insight Toolkit (ITK)
+of high-performance registration algorithms for neuroimage analysis
+\citep{avants2008a} built upon the mature and well-vetted Insight Toolkit (ITK)
 of the National Institutes of Health.  Since then, ANTs has grown to include 
-several algorithmic solutions necessary
-for robust medical image analysis including bias correction \cite{tustison2010}, 
-$n$-tissue multivariate segmentation \cite{avants2011}, template construction \cite{avants2010}, 
-and cortical thickness estimation \cite{das2009} (many of which have been
+several algorithmic solutions for robust medical image analysis including bias 
+correction \citep{tustison2010}, $n$-tissue multivariate segmentation 
+\citep{avants2011}, template construction \citep{avants2010}, and cortical 
+thickness estimation \citep{das2009} (many of which have been
 introduced into ITK partially in an attempted leveraging of Linus's Law---``Given enough eyeballs, all bugs are shallow'').  
-However, in the evolution of the toolkit, it became clear (as neuroimaging
-research certainly falls under the popular rubric of ``big data analysis'')
-that robust statistical machinery was lacking for proper inferences regarding
-the data produced by the various ANTs tools.
-ANTsR was developed
-specifically to provide an interface between a 
+However, in the evolution of the toolkit, it became clear 
+that robust statistical machinery was lacking for making inferences regarding
+the data produced during the course of ANTs processing.  ANTsR was developed
+specifically to provide an interface between ANTs, a 
 powerful neuroimaging toolkit for producing reliable imaging data 
-transformations and the R project%
+transformations, and the R project%
 \footnote{
@@
 
-In addition to describing ANTsR basics and how it can be generally
-used in multivariate neuroimaging studies, we showcase its use in
-the particularly difficult application of performing supervised brain
-segmentation using random forests.  Although random forests have
+In addition to describing ANTsR basics and how it can be
+used generally in multivariate neuroimaging studies, we showcase its use in
+the particularly difficult application of supervised brain tumor
+segmentation \cite{bauer2013} for which we use the widely popular 
+random forests machine learning technique.  Although random forests have
 been proposed previously in the literature for, among other things, 
-supervised brain segmentation (e.g. \cite{geremia2011,zikic2012}), 
+supervised brain segmentation (e.g. \cite{geremia2011}), 
 we demonstrate that ANTsR significantly facilitates the construction 
 of a fully functional, parallelizable workflow for such a task and 
-that the performance of our proposed framework was proven extremely 
-competitive in the recent MICCAI 2013 Multimodal Brain Tumor Segmentation
+that the performance of our proposed framework has proven extremely in 
+actual practice, specifically
+ in the recent MICCAI 2013 Multimodal Brain Tumor Segmentation
 Challenge (see Figure \ref{fig:award}).%
@@
 challenge, wanting to take advantage of the same methods as the winners,
-and finding that no teams had availed their software in any form.  As a 
-corrective \citep{ince2012}, and very much in line with our philosophy 
-regarding open science, we have made this framework available as open 
+and finding that no teams had availed their software in any form.  In contrast, 
+and very much in line with our philosophy 
+regarding open science \citep{tustison2013,ince2012}, we have made this framework available as open 
 source (even releasing the code before the actual 2013 contest).%
@@
 
-The quantitative complexity of neuroimaging research necessitates 
+The complexity of neuroimaging research necessitates 
 commensurable numerical analysis capabilities.  Similarly, concomitant
-with the era of ``big data'' (and neuroimaging certainly qualifies
-\citep{vanhorn2013}) are new visualization changes 
+with the era of ``big data'' (specifically with respect to neuroimaging
+\citep{vanhorn2013}) are new visualization needs and challenges
 \citep{childs2013,kehrer2013}.
 In response, various software packages have been developed to
-integrate tools specific to neuroimaging research with numerical
-and visualization software packages.
-The well-known neuroimaging standard SPM%
+integrate tools specific to neuroimaging research with more general
+numerical and visualization software packages.
+The well-known neuroimaging de facto standard SPM%
 \footnote{
@@
 }
-is built as an add-on to the commercial computing and visualization environment
+is a significant extension of the commercial computing and visualization environment
 Matlab.  Open source neuroimaging packages, such as NIPY (neuroimaging in Python),%
@@
 rely on other open source packages for numerical/statistical analysis.  NIPY,
-for example, uses other open source python packages such as NumPy.%
+for example, uses the more generic packages NumPy and SciPy for numerical analysis and 
+optimization.%
 \footnote{
@@
 Careful consideration of available options
-led to the adoption of R to complement ANTs capabilities which resulted in the
+led to the adoption of R to complement ANTs capabilities resulting in the
 ANTsR package.
@@
 \verb#mFilter#, \verb#doParallel#, \verb#robust#, \verb#magic#, \verb#knitr#, \verb#pixmap#, 
-\verb#rgl#, \verb#misc3d# which is easily facilitated by the 
-\verb#install.packages()# mechanism.  Additionally, in order
+\verb#rgl#, and \verb#misc3d#.  Additionally, in order
 to perform the supervised brain segmentation as described 
-in later sections, one would need to also install 
+in later sections, one needs to also install the packages
 \verb#randomForest#, \verb#snowfall#, \verb#rlecuyer#,
-and \verb#ggplot2#.
+and \verb#ggplot2#.%
+\footnote{
+Packages are easily installed using the {\tt install.packages()} R function.
+}
 
-In addition to R and relevant packages, CMake needs to also be
-installed.  CMake%
+In addition to R and the add-on packages previously mentioned, CMake is also 
+required.  CMake%
 \footnote{
@@
 
-\subsection{Supervised Brain Tumor Segmentation}
+\subsubsection{Usage}
+The utility of ANTsR stems from the ability to couple ANTs core 
+functionality, including IO tools such as \verb#antsImageRead#, 
+with the large number of R statistical and
+visualization packages.  Due to this combination, several
+functions have been easily created for such neuroimaging-specific 
+tasks as fMRI/ASL data manipulation and analysis,
+voxel and ROI-based  analyses,
+%(e.g. \verb#filterMRIforNetworkAnalysis#, \verb#aslPerfusion#)
+and connectivity visualization. % (e.g. \verb#plotBasicNetwork#).
+The user help menu and documentation for the library  and its
+constituent functions are invoked in the similar manner as other
+R libraries.
 
-Given a set of training data consisting of 
-labeled (csf, gray matter, white matter, edema, and tumor),
-multimodal brain image data, 
-supervised brain segmentation entails the generation of a 
-statistical model from such training data which can
-then be extended to testing data, i.e. unlabeled 
-brain data.  In subsequent sections, we detail the generic
-{\it random forest} modeling framework which takes as input
-a set of {\it feature images} (also described in a later section)
+\subsection{Supervised Brain Tumor Segmentation with Random Forests}
+
+Supervised techniques generally consist of a training phase
+for model construction followed by prediction using the previously
+generated model.  For supervised brain tumor segmentation, 
+a set of training data consisting of labeled brain images 
+(e.g. csf, gray matter, white matter, 
+necrosis, edema, enhancing and  non-enhancing tumor) is
+used to construct a predictive model.  Although other 
+classification techniques have been used to segment
+brain tumors (e.g. support vector machines \citep{bauer2011}),
+random forest models have proven particularly successful.
+
+In subsequent sections, we detail the generic
+random forest modeling framework which takes as input
+a set of feature images (also described in a later section)
 and labels for each training subject 
@@
 
-One of the core extensions that we provide in this work is
-the use of concatenated random forests for improved probabilistic estimation
-of the labels.
-As we demonstrate in the Results section, the set of feature
-images employed work sufficiently well for good performance
-on the training data (which exceeds that of what has been
-currently published in the literature).  However, we discovered
-that further improvements could be gained by using the probabilistic 
-label estimates as input to enhance preprocessing before a 
-second round of feature image generation including 
-modality-specific, prior-based $n$-tissue segmentation and 
-random forest model creation.
-
-\subsubsection{Random Forests}
+%One of the core extensions that we provide in this work is
+%the use of concatenated random forests for improved probabilistic estimation
+%of the labels.
+%As we demonstrate in the Results section, the set of feature
+%images employed work sufficiently well for good performance
+%on the training data.  However, we discovered
+%that further improvements could be gained by using the probabilistic 
+%label estimates as input to enhance preprocessing before a 
+%second round of feature image generation including 
+%modality-specific, prior-based $n$-tissue segmentation and 
+%random forest model creation.
 
@@
 multimodal brain tumor segmentation
-\cite{zikic2012,geremia2012}, brain extraction \cite{iglesias2010}, 
+\cite{bauer2012,zikic2012}, brain extraction \cite{iglesias2010}, 
 anatomy detection in computed tomography \cite{criminisi2013}, and
@@
 is the \verb#randomForest# package developed from Breiman's original
-Fortran code by Liaw and Wiener (described briefly in \cite{liaw2002}).
+Fortran code by Liaw and Wiener \citep{liaw2002}.
 
-\subsubsection{Multi-Modal Feature Image Preprocessing}
+\subsubsection{Brain Tumor Data}
 
-Although several studies have pointed out the importance of
-intensity normalization and bias correction, our experience 
-with the training data illustrated a degradation in performance
-when one or both steps (using \cite{nyul2000} and N4 \cite{tustison2010},
-respectively) were performed due to the presence of the tumor/edema complex.
+Brain tumor image data used in this work were obtained from the NCI-MICCAI 2013 
+Challenge on Multimodal Brain Tumor Segmentation%
+\footnote{
+http://martinos.org/qtim/miccai2013/index.html
+}
+organized by K. Farahani, M. Reyes,B. Menze, E. Gerstner, J. Kirby and J. Kalpathy-Cramer. The challenge database contains fully anonymized images from the following institutions: 
+ETH Zurich, University of Bern, University of Debrecen, and University of Utah and 
+publicly available images from the Cancer Imaging Archive (TCIA).  Both training 
+and testing data were made freely available through the Creative Commons Attribution-NonCommercial 3.0 license.
 
-As a corrective, for the first stage we simply windowed the image intensity
-for all images to be between the quantiles $[0.01,0.99]$ and
-subsequently rescaled to $[0,1]$.  From these ``corrected'' images,
-the first set of feature images were derived.  For the training cohort, these
-data were used to create the random forest regression model for the first
-stage.  During the second stage, the probabilistic estimates
-of the white matter and gray matter labels were used to generate a
-``pure tissue weight mask'' to estimate the bias field 
-using N4 (although the resulting bias field estimation was used
-to correct the image within the entire cerebral mask).  Formally, this 
-involved generation of a probabilistic map defined as:
+Training data consisted of multi-modal brain MRI (T1, T2, FLAIR, and 
+post-Gadolinium T1) from 30 glioma patients (both low, $n=10$, and high-grade, $n=20$,
+and with and without resection).  For each subject, the T1, T2, and 
+FLAIR MRI were linearly registered to the post-contrast T1.  Subsequently,
+the brains were skull-stripped and resampled to 1 mm isotropic resolution.
+Testing data was processed similarly and released during the course of the
+challenge in two sets denoted as ``LeaderBoard'' and ``Challenge'' data.  
+The former consisted of 21 and 4 high and low-grade tumor patients, respectively,
+whereas the latter comprised 10 high-grade only patients.
+
+Manual labeling was performed in the axial plane following a detailed
+protocol.%
+\footnote{
+http://martinos.org/qtim/miccai2013/data.html.
+}
+The labeling of pathology was categorized into four regions:
+edema, non-enhancing tumor (including low-grade tumor center), 
+enhancing tumor (excluding necrotic center), and abnormal
+necrotic center or necrocyst in high-grade gliomas.
+Normal brain tissue was not labeled.
+
+%\subsubsection{Multi-Modal Feature Image Preprocessing}
+%Although several studies have pointed out the importance of
+%intensity normalization and bias correction, our experience 
+%with the training data illustrated a degradation in performance
+%when one or both steps (using \cite{nyul2000} and N4 \cite{tustison2010},
+%respectively) were performed due to the presence of the tumor/edema complex. 
+% 
+%As a corrective, for the first stage we simply windowed the image intensity
+%for all images to be between the quantiles $[0.01,0.99]$ and
+%subsequently rescaled to $[0,1]$.  From these ``corrected'' images,
+%the first set of feature images were derived.  For the training cohort, these
+%data were used to create the random forest regression model for the first
+%stage.  During the second stage, the probabilistic estimates
+%of the white matter and gray matter labels were used to generate a
+%``pure tissue weight mask'' to estimate the bias field 
+%using N4 (although the resulting bias field estimation was used
+%to correct the image within the entire cerebral mask).  Formally, this 
+%involved generation of a probabilistic map defined as:
+%\begin{align}
+%  P_{pure\,\,tissue}(\mathbf{x}) = \sum_{i=1}^N P_i(\mathbf{x}) \prod_{j=1, j \neq i}^N \left( 1 - P_j(\mathbf{x}) \right)
+%\end{align}
+%where $N$ is the set of user-selected tissue labels (in our
+%case $N=2$ consisting of the gray and white matter probability
+%maps).
+%
+%Both rescaling and weighted bias correction were applied to produce
+%the ``corrected images'' for the second stage resulting in
+%modified features images for the second stage.  Note that we
+%perform a similar iterative scenario for normal brain 
+%segmentation \cite{avants2011} (encapsulated in the ANTs script 
+%\verb#antsAtroposN4.sh#).
+
+\subsubsection{Symmetric Multivariate Template Construction}
+
+\begin{figure}
+  \centering
+  \begin{tabular}{cc}
+    \includegraphics[width=40mm]{Figures/S_templateFA_140.png} &
+    \includegraphics[width=40mm]{Figures/S_templateFLAIR_140.png} \\
+    (a) & (b) \\
+    \includegraphics[width=40mm]{Figures/S_templateT1_140.png} &
+    \includegraphics[width=40mm]{Figures/S_templateT2_140.png} \\
+    (c) & (d) 
+  \end{tabular}
+  \caption{Multivariate symmetric template created from the Kirby 
+           21 data described in \cite{landman2011}.  Shown are the
+           (a) fractional anisotropy (FA), (b) FLAIR, (c) MPRAGE, 
+           and (d) T2 template components.
+          }
+  \label{fig:symmetrictemplates}
+\end{figure}
+
+In order to better characterize deviations from normal
+multi-modal brain shape and appearance, several features were derived 
+using population-specific multivariate template 
+construction. A recent neuroimaging reproducibility study
+by Landman et al. resulted in an open data cohort of 21
+normal individuals, each imaged twice, comprising several
+modalities including ASL, FLAIR, DTI, fMRI, T1, and T2 
+\cite{landman2011}.  These data (known informally as the
+``Kirby'' data set) were selected for deriving
+a multivariate template due to its public availability and
+inclusion of several modalities (even permitting future 
+incorporation into our segmentation framework of modalities 
+not currently included with the BRATS challenge).
+
+As detailed in \cite{avants2008,avants2010}, 
+given $K$ image modalities for $N$ subjects,  
+${\mathbf I} = \{I_1,I_2,\ldots, I_K\}$, multivariate 
+template construction iterates between optimizing the set 
+of diffeomorphic transforms between the subjects and the 
+template, 
+$\left\{\left(\phi_1,\phi_1^{-1}\right),\ldots,\left(\phi_N,\phi_N^{-1}\right)\right\}$ 
+and constructing the 
+optimal multivariate template appearance 
+$\mathbf{J}=\{J_1,J_2,\ldots, J_K\}$ to minimize the
+following cost function:
 \begin{align}
-  P_{pure\,\,tissue}(\mathbf{x}) = \sum_{i=1}^N P_i(\mathbf{x}) \prod_{j=1, j \neq i}^N \left( 1 - P_j(\mathbf{x}) \right)
+  \sum_{n=1}^N 
+        \Bigg[ D &\left( \psi(\mathbf{x}),\phi_1^n(\mathbf{x},1)\right) \\ \nonumber 
+        +& \sum_{k=1}^K \lambda_k \Pi_k \left(I_k^n\left(\phi_n(\mathbf{x},0.5)\right),J_k\left(\phi^{-1}_n(\mathbf{x},0.5)\right)\right)\Bigg].
 \end{align}
-where $N$ is the set of user-selected tissue labels (in our
-case $N=2$ consisting of the gray and white matter probability
-maps).
+$D$ is the diffeomorphic shape distance,
+\begin{align}
+D\left( \phi( \mathbf{x},0),\phi( \mathbf{x},1)\right) = \int_0^1 \| \nu(\mathbf{x},t)\|_L dt
+\end{align}
+dependent on the choice of linear operator, $L$, and $\nu$
+is the velocity field
+\begin{align}
+\nu\left( \phi(\mathbf{x},t) \right) = \frac{d\phi(\mathbf{x},t)}{dt},\,\,\, \phi(\mathbf{x},0) = \mathbf{x}.
+\end{align}
+Each pairwise registration employing the similarity metric $\Pi_k$ can 
+be assigned a relative weighting, $\lambda_k$, to weight a particular
+modality's influence in the construction process.
 
-Both rescaling and weighted bias correction were applied to produce
-the ``corrected images'' for the second stage resulting in
-modified features images for the second stage.  Note that we
-perform a similar iterative scenario for normal brain 
-segmentation \cite{avants2011} (encapsulated in the ANTs script 
-\verb#antsAtroposN4.sh#).
+In terms of implementation, this algorithm is 
+encapsulated in the script \verb#antsMultivariateTemplateConstruction.sh#,
+available in the ANTs repository, which permits parallel processing on
+an individual workstation or on a large computational cluster.  In
+Figure \ref{fig:symmetrictemplates} we show mid-axial slices from
+the Kirby multivariate symmetrical template consisting of FA, FLAIR,  
+T1, and T2 components.  Although DTI-based images were not included in the
+challenge data, such images have shown discriminative potential 
+\citep{price2003} to warrant investigation in our future work.
 
@@
 
-Key to any supervised regression or classification protocol are the 
-selected features for training and subsequent testing.  Based on previous
-work and our own experience, we selected the following feature images
-to showcase the supervised segmentation strategy developed in this work.
-
 \begin{figure*}
   \centering
-  \includegraphics[width=180mm]{Figures/featuresImages.pdf}
-  \caption{Feature images from the BRATS\_HG0004 data set.}
+  \includegraphics[width=180mm]{Figures/featureImages.pdf}
+  \caption{Representative feature images derived from the BRATS\_HG0301 ``challenge'' data set.
+           Neighborhood statistical images for each modality were generated by calculating a 
+           given statistic within
+           a specified neighborhood radius.  Also calculated for each modality were feature
+           images based on either the GMM or the MAP-MRF segmentation.  For the former, we
+           show the probability maps for each of the seven labels which are used as feature
+           images.  From the resulting hard segmentation, we calculate various geometric 
+           measures per connected component of each of the seven labels.  Similarly, the 
+           registration to the symmetric template produces the modality-specific difference images with the
+           corresponding symmetric template itself and with respect to the contralateral side.   
+           This mapping is also used to produce the log Jacobian image.  Finally, the T1 - TC image
+           is calculated and, from the
+           cerebral mask, we calculate the normalized distance image.  
+           }
 \end{figure*}
 
+The workflow for estimating tumor-based labeling from multi-modal 
+images involves the following steps:
+\begin{enumerate}
+  \item Symmetric multivariate template construction \cite{avants2010}
+ using the data described in \cite{landman2011}.
+  \item Image preprocessing:
+    \begin{itemize}
+      \item Windowing intensities (quantiles $[0.01,0.99]$).
+      \item N4 bias correction \cite{tustison2010}.
+      \item Rescaling intensity range to $[0,1]$.  
+    \end{itemize}
+  \item Stage 1 (GMM) processing:
+  \begin{itemize}
+    \item generation of feature images,
+    \item construction of the Stage 1 RF model and probability images.
+  \end{itemize}
+  \item Stage 2 processing:
+  \begin{itemize}
+    \item generation of single-modality MAP-MRF images using the Stage 1 RF probability images as spatial priors,
+    \item construction of the Stage 2 RF model and labelings.
+  \end{itemize}
+  \item Refinement of Stage 2 labelings using a heuristically-derived binary morphological processing protocol.
+\end{enumerate}
+
+Key to any supervised regression or classification protocol are the 
+selected features for training and subsequent prediction.  Based on previous
+work and our own experience, the following feature images were selected:
 \begin{itemize}
@@
             mean, variance, skewness, and entropy. 
-            Neighborhood radius = 3.
-    \item GMM-based posteriors: CSF, gray matter, white matter, edema, and tumor
-    \item GMM connected component geometry features:  volume, volume to 
-          surface area ratio, eccentricity, elongation
-    \item Template-based:  symmetric template difference and contralateral difference.
-          Gaussian smoothing ($\sigma = 4$mm).
+            Neighborhood radius $\in \{1,3\}$.
+    \item GMM (stage 1) and MAP-MRF (stage 2) posteriors: CSF, gray matter, white 
+          matter, necrosis, edema, non-enhancing tumor and enhancing tumor (or a
+          subset for the simulated data).
+    \item GMM (stage 1) and MAP-MRF (stage 2) connected component geometry 
+          features:  distance to tumor core label, volume, volume to surface area ratio, eccentricity, and elongation
+    \item Template-based:  symmetric template difference and 
+          contralateral difference with Gaussian smoothing ($\sigma = 4$mm).
     \end{itemize}
-  \item Not modality specific
-    \begin{itemize}
-    \item Normalized Euclidean distance
-    \item Log Jacobian  
-    \item T1C,T1 difference image
-    \end{itemize}
+  \item Miscellaneous: normalized Euclidean distance based on cerebral mask,
+    log Jacobian image, and  (T1C - T1) difference image.
 \end{itemize}
@@
 modeling approaches for multi-modal tumor segmentation 
-(e.g. \cite{prastawa2003,zikic2012}), we do not use multivariate 
+(e.g. \cite{prastawa2003}), we do not use multivariate 
 Gaussians to specify tissue probabilities but rather incorporate each
@@
 
-In order to better characterize deviations from normal
-multi-modal brain shape and appearance, several features were derived 
-using population-specific multivariate template 
-construction. A recent neuroimaging reproducibility study
-by Landman et al. resulted in an open data cohort of 21
-normal individuals, each imaged twice, comprising several
-modalities including arterial spin labeling, 
-fluid attenuated inversion recovery (FLAIR),
-diffusion tensor imaging, functional imaging, T1, and T2 
-\cite{landman2011}.
+\subsubsection{Training:  Random Forest Model Creation}
 
-Given $K$ image modality types for $N$ subjects,  
-${\mathbf I} = \{I_1,I_2,\ldots, I_K\}$, multivariate 
-template construction iterates between optimizing the set 
-of diffeomorphic transforms between the subjects and the 
-template, 
-$\left\{\left(\phi_1,\phi_1^{-1}\right),\ldots,\left(\phi_N,\phi_N^{-1}\right)\right\}$ 
-and constructing the 
-optimal multivariate template appearance 
-$\mathbf{J}=\{J_1,J_2,\ldots, J_K\}$ to minimize the
-following cost function:
-\begin{align}
-  \sum_{n=1}^N 
-        \Bigg[ D &\left( \psi(\mathbf{x}),\phi_1^n(\mathbf{x},1)\right) \\ \nonumber 
-        +& \sum_{k=1}^K \lambda_k \Pi_k \left(I_k^n\left(\phi_n(\mathbf{x},0.5)\right),J_k\left(\phi^{-1}_n(\mathbf{x},0.5)\right)\right)\Bigg]
-\end{align}
-where $D$ is the diffeomorphic shape distance,
-\begin{align}
-D\left( \phi( \mathbf{x},0),\phi( \mathbf{x},1)\right) = \int_0^1 \| \nu(\mathbf{x},t)\|_L dt
-\end{align}
-dependent on the choice of linear operator, $L$, and $\nu$
-is the velocity field
-\begin{align}
-\nu\left( \phi(\mathbf{x},t) \right) = \frac{d\phi(\mathbf{x},t)}{dt},\,\,\, \phi(\mathbf{x},0) = \mathbf{x}.
-\end{align}
-Each pairwise registration employing the similarity metric $\Pi_k$ can 
-be assigned a relative weighting, $\lambda_k$, to weight a particular
-modality's influence in the construction process.  Further theoretical
-details can be found in \cite{avants2008,avants2010}.
-In terms of implementation, this algorithm is 
-encapsulated in the script \verb#antsMultivariateTemplateConstruction.sh#,
-available in the ANTs repository, which permits parallel processing on
-an individual workstation or on a large computational cluster.
+\subsubsection{Prediction:  Applying Random Forest Models}
 
@@
   \centering
-  \begin{tabular}{cc}
-    \includegraphics[width=40mm]{Figures/S_templateFA_140.png} &
-    \includegraphics[width=40mm]{Figures/S_templateFLAIR_140.png} \\
-    (a) & (b) \\
-    \includegraphics[width=40mm]{Figures/S_templateT1_140.png} &
-    \includegraphics[width=40mm]{Figures/S_templateT2_140.png} \\
-    (c) & (d) 
+  \begin{tabular}{c}
+    \includegraphics[width=85mm]{Figures/pipeline.pdf}
   \end{tabular}
-  \caption{Multivariate symmetric template created from the Kirby 
-           21 data described in \cite{landman2011}.  Shown are the
-           (a) fractional anisotropy (FA), (b) FLAIR, (c) MPRAGE, 
-           and (d) T2 template components.
+  \caption{Diagrammatic workflow for the proposed RF model prediction.  The feature
+  images are first generated from the input MRI and symmetric multivariate template. 
+  After organization of the feature images in a csv file, the stage 1 RF model is 
+  applied which results in a first stage classification voting output.  This output
+  is used to generate an additional set of feature images for the second classification
+  stage.  The final refinement process entails a series of binary morphological 
+  operations which results in the final labeled brain.  In the diagram we also list
+  the files used to perform each set of functions which are available as open source.
   }
-  \label{fig:symmetrictemplates}
+  \label{fig:pipeline}
 \end{figure}
 
-\subsubsection{\underline{Bra}in \underline{T}umor \underline{S}egmentation Challenge Data}
-
-The Brain tumor segmentation
-Associated with the 2012 International Conference on Medical Image Computing and Computer Assisted
-Intervention (MICCAI),%
-\footnote{
-http://www.miccai2012.org
-}
-the
-
-\subsubsection{MS lesion segmentation challenge Challenge Data}
+\begin{figure*}
+  \centering
+  \begin{tabular}{cc}
+  \includegraphics[width=80mm]{Figures/BRATS_HG_GMM.pdf} &
+  \includegraphics[width=80mm]{Figures/BRATS_HG_MAP_MRF.pdf} \\
+  \includegraphics[width=80mm]{Figures/BRATS_LG_GMM.pdf} &
+  \includegraphics[width=80mm]{Figures/BRATS_LG_MAP_MRF.pdf}
+  \end{tabular}
+  \caption{Feature images from the BRATS\_HG0004 data set.}
+\end{figure*}
 
@@
 
-\section*{Acknowledgments}
+%\section*{Acknowledgments}
 

### 2013-11-03 Scripts/applyModel.R

@@ -1,0 +1,168 @@
+library( ANTsR )
+library( randomForest )
+library( snowfall )
+
+stopQuietly <- function(...)
+  {
+  blankMsg <- sprintf( "\r%s\r", paste( rep(" ", getOption( "width" ) - 1L ), collapse = " ") );
+  stop( simpleError( blankMsg ) );
+  } # stopQuietly()
+
+args <- commandArgs( trailingOnly = TRUE )
+
+###############################################
+#
+# Selected parameters
+#
+###############################################
+
+if( length( args ) < 3 )
+  {
+  cat( "Usage: Rscript applyModel.R dimension inputModel inputCSVFile ",
+       "outputProbabilityImagePrefix <numberOfThreads=4>", sep = "" )
+  stopQuietly()
+  }
+
+dimension <- as.numeric( args[1] )
+inputModelName <- args[2]
+fileList <- read.csv( args[3] )
+probImagePrefix <- args[4]
+
+numberOfThreads <- 1
+if( length( args ) >= 5 )
+  {
+  numberOfThreads <- as.numeric( args[5] )
+  }
+
+###############################################
+#
+# Load model:  contained in the variable "modelForest"
+#
+###############################################
+
+load( inputModelName )
+
+###############################################
+#
+# Put the image data into a data frame (modelData)
+#
+###############################################
+
+maskName <- fileList[1,1]
+featureImages <- fileList[1,2:ncol( fileList )]
+featureNames <- colnames( featureImages )
+
+## Create the model data frame
+
+maskImage <- antsImageRead( as.character( maskName ), dimension = dimension, pixeltype = 'unsigned int' )
+mask <- as.array( maskImage )
+
+maskIndices <- which( mask != 0 )
+
+subjectData <- matrix( NA, nrow = length( maskIndices ), ncol = length( featureNames ) )
+for( j in 1:length( featureNames ) )
+  {
+  cat( "  Reading feature image ", featureNames[j], ".\n", sep = "" )
+  featureImage <- as.array( antsImageRead( as.character( featureImages[1,j] ), dimension = dimension, pixeltype = 'float' ) )
+
+  values <- featureImage[maskIndices]
+  subjectData[, j] <- values
+  }
+
+colnames( subjectData ) <- c( featureNames )
+subjectData <- as.data.frame( subjectData )
+
+# If the subject data has NA's, we need to get rid of them
[…]

### 2013-11-03 Scripts/applyTumorSegmentationModelForCohort.pl

@@ -1,0 +1,349 @@
+#/usr/bin/perl -w
+
+use strict;
+
+use Cwd 'realpath';
+use File::Find;
+use File::Basename;
+use File::Path;
+use File::Spec;
+use FindBin qw($Bin);
+
+my $usage = qq{
+
+  Usage: applyTumorSegmentationModelForCohort.pl <inputDirectory> <outputDirectory>
+
+ };
+
+my ( $inputDirectory, $outputDirectory ) = @ARGV;
+
+
+my $ANTSPATH = $ENV{ 'ANTSPATH' };
+my $UTILPATH = $ENV{ 'UTILPATH' };
+
+my $basedir = '/home/njt4n/share/Data/Tumor/BRATS/';
+
+my $templatedir = '/home/njt4n/share/Data/Public/Kirby/SymmetricTemplate/';
+my $templateLabels = "${templatedir}/S_templateJointLabels_6labels.nii.gz";
+my $templateMask = "${templatedir}/S_templateMask.nii.gz";
+my $clusterdir = "${basedir}/ClusterCenters/";
+
+my $modeldir = "${basedir}/RandomForestModels/";
+
+my @templates = ();
+$templates[0] = "${templatedir}/S_template2_skullStripped_RESCALED.nii.gz";          #FLAIR
+$templates[1] = "${templatedir}/S_template3_skullStripped_RESCALED.nii.gz";          #T1
+$templates[2] = "${templatedir}/S_template3_skullStripped_RESCALED.nii.gz";          #T1
+$templates[3] = "${templatedir}/S_template5_skullStripped_RESCALED.nii.gz";          #T2
+
+my @suffixList = ( ".mha", ".nii.gz" );
+
+###
+##
+## Might need to calculated the cluster centers or list them here for future use.
+##
+###
+
+###
+##
+## Create feature images for each subject (send to cluster)
+##
+###
+
+my @subjectdirs = <${inputDirectory}/*>;
+
+for( my $i = 0; $i < @subjectdirs; $i++ )
+  {
+  print "****************************\n";
+  print "$subjectdirs[$i]            \n";
+  print "****************************\n";
+
+  if( ! -d $subjectdirs[$i] )
+    {
+    next;
+    }
+
+  my @comps = split( '/', $subjectdirs[$i] );
+
+  my $subjectID = ${comps[-1]};
[…]

### 2013-11-03 Scripts/createFeatureImages.sh

@@ -1,0 +1,600 @@
+#!/bin/bash
+
+VERSION="0.0"
+
+if [[ ! -d "$ANTSPATH" ]];
+  then
+    echo We can\'t find the ANTs path -- does not seem to exist.  Please \(re\)define \$ANTSPATH in your environment.
+    exit 1
+  fi
+
+function Usage {
+    cat <<USAGE
+
+`basename $0` creates feature images for tumor segmentation using random decision forests.
+   See <reference goes here>
+
+Usage:
+
+`basename $0` -d imageDimension
+              -a inputImage(s)
+              -t symmetricTemplate(s)
+              -c clusterCenters
+              -n imageNames
+              -r neighgorhoodRadius
+              -s smoothSigma
+              -f differencePair
+              -x maskImage
+              -o outputPrefix
+
+Example:
+
+  bash $0 -d 3 -a t1.nii.gz -t symmetricTemplate.nii.gz -o output
+
+Required arguments:
+
+     -d:  image dimension                       2 or 3 (for 2- or 3-dimensional image)
+     -a:  input image                           Anatomical image(s) comprised of flair, t2, t1, or t1 contrast.
+                                                It is assumed that the following preprocessing steps have
+                                                already been performed:
+                                                   * brain extraction
+                                                   * bias correction
+                                                   * intensity normalized (i.e. histogram matching and intensity truncation)
+     -c:  cluster centers                       Array describing the intensity centers of the intensity normalized images.
+                                                Need one for each input image.  Should be of the form:  e.g. 0.14x0.57x0.37x0.83x0.95
+                                                (for 7 classes: csf, gm, wm, edema, tumor, core, enhancement).  Note that either the cluster
+                                                centers are specified (for testing) or the truth labels (for training) but not both. If these
[…]

### 2013-11-06 paper.tex
ENH:  Getting closer.

@@
 
-%\documentclass[preprint,authoryear,review,12pt]{elsarticle}
+% \documentclass[preprint,authoryear,review,12pt]{elsarticle}
 \documentclass[final,5p,times,twocolumn]{elsarticle}
@@
 \author[label2]{K.~L.~Shrinidhi}
+\author[label1]{Max Wintermark}
+\author[label1]{Christopher R.~Durst}
 \author[label2]{Jeffrey T.~Duda}
-\author[label1]{Christopher R.~Durst}
-\author[label1]{Max Wintermark}
 \author[label1]{James C.~Gee}
-\author[label1]{Murray C.~Grossman}
+\author[label3]{Murray C.~Grossman}
 \author[label2]{Brian B.~Avants}
 \address[label1]{Department of Radiology and Medical Imaging, University of Virginia, Charlottesville, VA}
-\address[label2]{Penn Image Computing and Science Laboratory, University of Pennsylvania,
+\address[label2]{Penn Image Computing and Science Laboratory, 
+                 Department of Radiology, University of Pennsylvania,
+                Philadelphia, PA}
+\address[label3]{Department of Neurology, University of Pennsylvania,
                 Philadelphia, PA}
@@
 %\maketitle
-
-\linenumbers
 
@@
 \end{frontmatter}
+%\linenumbers
+
 %
@@
   \centering
-  \includegraphics[width=80mm]{Figures/award.pdf}
+  \includegraphics[width=90mm]{Figures/award.pdf}
   \caption{First place certificate from the MICCAI 2013 BRATS challenge.}
@@
 \footnote{
-Packages are easily installed using the {\tt install.packages()} R function.
+Packages are easily installed using the {\tt install.packages()} R mechanism.
 }
@@
 
+As mentioned earlier, we have made this entire framework
+available as open source.  In addition to the ANTsR repository
+already on github which houses both ANTs and ANTsR functionality, 
+we created a special github repository specifically for this work
+containing figures, references, and text.%
+\footnote{
+https://github.com/ntustison/ANTsAndArboles
+}
+Also, we posted all
+scripts (R, shell, and perl) used to coordinate the ANTsR processing 
+including (along with a brief description of each script):
+\begin{itemize}
+  \item {\tt applyModel.R}:  applies a random forest model to a new 
+  feature data set from a testing subject resulting in a set of probability
+  images (one for each label).
+  \item {\tt applyTumorSegmentationModel.sh}:  generates the new feature image set 
+  from the testing MRI (by calling {\tt createFeatureImages.sh}).
+  organizes the file names in a csv file (via {\tt createCSVFileFromModel.R}),
+  and applies the random forest model using {\tt applyModel.R}. 
+  \item {\tt applyTumorSegmentationModelForCohort.pl}:  Coordinates tumor 
+  segmentation on the computational cluster for a given cohort.
+  \item {\tt createCSVFileFromModel.R}:  organizes the set of feature image
+  file names in a csv file for input into {\tt applyModel.R}.
+  \item {\tt createFeatureImages.sh};  creates the set of feature images given
+  a set of co-registered input MRI from a single subject. 
+  \item {\tt createModel.R}:  creates random forest model given the input csv 
+  file of the feature image file names for all training data and set of truth
+  label maps.
+  \item {\tt plotVariableImportance.R}:  produces a plot of the importance 
+  (`{\tt MeanDecreaseAccuracy}') of each feature variable used 
+  in constructing the model.
+\end{itemize}
+We also include a fully functional 2-D example which performs
+both testing and training on sample challenge data.
+
 \subsection{Supervised Brain Tumor Segmentation with Random Forests}
@@
 Supervised techniques generally consist of a training phase
-for model construction followed by prediction using the previously
+for model construction followed by prediction using the 
 generated model.  For supervised brain tumor segmentation, 
@@
 random forest models have proven particularly successful.
-
-In subsequent sections, we detail the generic
-random forest modeling framework which takes as input
-a set of feature images (also described in a later section)
-and labels for each training subject 
-and outputs a statistical model.  This model, in turn, can then
-be used to provide a probabilistic estimate of the labels in an
-unlabeled subject.
-
-%One of the core extensions that we provide in this work is
-%the use of concatenated random forests for improved probabilistic estimation
-%of the labels.
-%As we demonstrate in the Results section, the set of feature
-%images employed work sufficiently well for good performance
-%on the training data.  However, we discovered
-%that further improvements could be gained by using the probabilistic 
-%label estimates as input to enhance preprocessing before a 
-%second round of feature image generation including 
-%modality-specific, prior-based $n$-tissue segmentation and 
-%random forest model creation.
 
@@
 introduced by Breiman \cite{breiman1996} to achieve improved
-accuracy.
+accuracy.%
+\footnote{
+One of the principal advantages of R is the extensive community of
+developers  who have contributed on the order of thousands of packages 
+extending R's capabilities beyond its core functionality.
+Most relevant 
+is the {\tt randomForest} package developed from Breiman's original
+Fortran code by Liaw and Wiener \citep{liaw2002}.
+}
 
@@
 
-One of the principal advantages of R is the extensive community of
-developers  who have contributed on the order of thousands of packages 
-extending R's capabilities beyond its core functionality.
-Most relevant for the work described in this paper
-is the \verb#randomForest# package developed from Breiman's original
-Fortran code by Liaw and Wiener \citep{liaw2002}.
+%In subsequent sections, we detail the generic
+%random forest modeling framework which takes as input
+%a set of feature images (also described in a later section)
+%and labels for each training subject 
+%and outputs a statistical model.  This model, in turn, can then
+%be used to provide a probabilistic estimate of the labels in an
+%unlabeled subject.
+
+One of the innovations that we provide in this work is
+the use of concatenated random forests for improved probabilistic estimation
+of the brain segmentation labels.  Briefly, Gaussian mixture
+modeling (GMM) is used to create a subset of feature images which are
+used to train and predict during the first classification stage.  
+The random forest voting output probabilistic maps are then used as 
+spatial probability priors for performing a MAP-MRF segmentation
+and subsequent feature image creation
+thereby replacing the GMM-derived feature images for a second
+classification stage.  This dual random forest model generation
+and application
+improves results and is detailed in later sections.
+
+%As we demonstrate in the Results section, the set of feature
+%images employed work sufficiently well for good performance
+%on the training data.  However, we discovered
+%that further improvements could be gained by using the probabilistic 
+%label estimates as input to enhance preprocessing before a 
+%second round of feature image generation including 
+%modality-specific, prior-based $n$-tissue segmentation and 
+%random forest model creation.
 
@@
 Testing data was processed similarly and released during the course of the
-challenge in two sets denoted as ``LeaderBoard'' and ``Challenge'' data.  
+challenge in two sets denoted as ``Leaderboard'' and ``Challenge'' data.  
 The former consisted of 21 and 4 high and low-grade tumor patients, respectively,
@@
 challenge data, such images have shown discriminative potential 
-\citep{price2003} to warrant investigation in our future work.
+\citep[e.g.][]{price2003} to warrant investigation in our future work.
 
@@
   \centering
+  \makebox[\textwidth][c]{
   \includegraphics[width=180mm]{Figures/featureImages.pdf}
-  \caption{Representative feature images derived from the BRATS\_HG0301 ``challenge'' data set.
-           Neighborhood statistical images for each modality were generated by calculating a 
-           given statistic within
+  }
+  \caption{Representative feature images derived from the 
+           BRATS\_HG0301  ``challenge'' data set.
+           Neighborhood statistical images for each modality were 
+           generated by calculating a given statistic within
            a specified neighborhood radius.  Also calculated for each modality were feature
@@
            measures per connected component of each of the seven labels.  Similarly, the 
-           registration to the symmetric template produces the modality-specific difference images with the
-           corresponding symmetric template itself and with respect to the contralateral side.   
-           This mapping is also used to produce the log Jacobian image.  Finally, the T1 - TC image
-           is calculated and, from the
+           registration to the symmetric template produces the modality-specific 
+           difference images with the
+           corresponding symmetric template itself and with respect to the 
+           contralateral side.   
+           This mapping is also used to produce the log Jacobian image.  Finally, the 
+           (T1 - TC) image is calculated and, from the
            cerebral mask, we calculate the normalized distance image.  
            }
+  \label{fig:featureImages}         
 \end{figure*}
 
-The workflow for estimating tumor-based labeling from multi-modal 
-images involves the following steps:
-\begin{enumerate}
-  \item Symmetric multivariate template construction \cite{avants2010}
- using the data described in \cite{landman2011}.
-  \item Image preprocessing:
-    \begin{itemize}
-      \item Windowing intensities (quantiles $[0.01,0.99]$).
-      \item N4 bias correction \cite{tustison2010}.
-      \item Rescaling intensity range to $[0,1]$.  
-    \end{itemize}
-  \item Stage 1 (GMM) processing:
-  \begin{itemize}
-    \item generation of feature images,
-    \item construction of the Stage 1 RF model and probability images.
-  \end{itemize}
-  \item Stage 2 processing:
-  \begin{itemize}
-    \item generation of single-modality MAP-MRF images using the Stage 1 RF probability images as spatial priors,
-    \item construction of the Stage 2 RF model and labelings.
-  \end{itemize}
-  \item Refinement of Stage 2 labelings using a heuristically-derived binary morphological processing protocol.
-\end{enumerate}
-
+%For both training and prediction protocols, feature images are generated
+%for each subject using the selected input MRI.  
 Key to any supervised regression or classification protocol are the 
 selected features for training and subsequent prediction.  Based on previous
-work and our own experience, the following feature images were selected:
+work and our own experience, the following feature images are generated
+for use in both training and prediction:
 \begin{itemize}
@@
   \item Miscellaneous: normalized Euclidean distance based on cerebral mask,
-    log Jacobian image, and  (T1C - T1) difference image.
+    log Jacobian image, and  (T1 - T1C) difference image.
 \end{itemize}
+The generation of all feature images is performed using the available script 
+{\tt createFeatureImages.sh} which also preprocesses the input images performing N4
+bias correction \citep{tustison2010} and intensity rescaling to $[0,1]$.  Note that 
+it is easily expandable to include 
+additional modalities and/or neighborhood and smoothing parameters.
+In Figure \ref{fig:featureImages}
+we provide sample mid-axial slices from the BRATS\_HG0301
+data set from the Challenge cohort.
 
-For each modality, we create four first-order statistical feature images,
-five Gaussian mixture model (GMM)-based posterior probability feature images,
-four geometry features generated from the GMM posterior probability images
-based on connected components, and two difference images using symmetric template
-construction for a total of 4 modalities $\times$ (4 + 5 + 4 + 2) feature images per modality $=$ 60 total
-feature images.  We employ two additional images consisting of the 
-Euclidean distance image \cite{maurer2003} created from the skull-stripped 
-binary mask rescaled to the range $[0,1]$ and the log
-Jacobian image derived from the spatial normalization of the symmetric multivariate template and individual subject images.  Given the intensity corrected images
-the corresponding multivariate template images, and a brain mask for each subject
-creation of all feature images is performed using the script \verb#createFeatureImages.sh#.
+We previously mentioned that all data was provided already 
+skull-stripped.  However, we note to the reader that an
+available ANTs-based solution to the brain extraction problem
+is found in the script {\tt antsBrainExtraction.sh} which uses
+a hybrid approach based on image registration and binary morphology
+to isolate the brain \cite{avants2010a}.  We recently performed an evaluation of over 1200
+MRI which demonstrated perfect performance \cite{tustison2013a}.
 
-Prior  cluster centers for specific tissue types learned from training data \cite{reynolds2009} are used in the GMM to create multiple feature images.  
-Given $M$ tissue types (e.g. CSF, gray matter,
-white matter, edema, and tumor), a GMM formulates the 
+\paragraph{Stage 1 (GMM)}
+Prior  cluster centers for specific tissue types learned from training data \citep{reynolds2009} are used in the GMM random forest stage to create 
+multiple feature images.  
+Given $M$ tissue types, a GMM formulates the 
 probability distribution at each voxel, $\mathbf{x}$, as the
@@
 estimation which can performed using the Atropos segmentation 
-tool \cite{avants2011}.  In contrast to previous generative
+tool \citep{avants2011} available in ANTs.
+
+\paragraph{Stage 2 (MAP-MRF):}
+
+Prediction during the GMM stage results in a set of random 
+forest voting classification maps giving a probabilistic estimate
+of the label value at each voxel.  These spatial probability
+maps can be used as spatial priors in a MAP-MRF segmentation
+estimation protocol also performed using the Atropos segmentation 
+tool \citep{avants2011} available in ANTs although we couple
+it with iterative bias correction as contained in the script
+{\tt antsAtroposN4.sh}.  The probabilistic output (and corresponding
+feature images) 
+for each modality are used to replace the analogous GMM-based feature images used 
+during the first stage of prediction along with the other feature
+images previously generated (such as the template-based data).
+We discovered that this composite application of random
+forest models improves the resulting labeled output.
+
+For both stages, in contrast to previous generative
 modeling approaches for multi-modal tumor segmentation 
-(e.g. \cite{prastawa2003}), we do not use multivariate 
+\citep[e.g.][]{prastawa2003}, we do not use multivariate 
 Gaussians to specify tissue probabilities but rather incorporate each
@@
 process determine the optimal combination of such multivariate
-information.
-Additionally, maximum posterior labeling from the GMM processing
+information.  Additionally, maximum posterior labeling from both stages
 is used to determine the connected components for each label.  
 Geometric features (assigned voxel-wise) include the physical volumes 
-of each connected component including the volume to surface area ratio, 
+of each connected component, the volume to surface area ratio, 
 the elongation, and eccentricity.
+
+\paragraph{Neighborhood First-Order Statistic Feature Images}
+Similar to other work \cite{bauer2012,zikic2012} we calculated
+feature images using first order statistics such as mean, 
+variance, skewness, and entropy.  Images were created for 
+neighborhood radii of 1 and 3 voxels.
+
+\paragraph{Template-based Feature Images}
+Additional feature images are derived from the multivariate 
+symmetric template described earlier. The T1 images of the 
+subject and symmetric template are co-registered 
+\citep{tustison2013a} using the {\tt antsRegistration} 
+tool.  The motivation is similar in spirit to the method
+using reflection about the mid-sagittal plane described in 
+\cite{geremia2012}.  However, tumor growth often causes 
+distortion of the mid-plane thereby limiting the accuracy of
+this method.  By mapping to the symmetric template using the
+highly accurate SyN algorithm \cite{avants2011a}, we can 
+account for any distortion for comparing contralateral
+sides.
+
+Two images per modality use this derived mapping.  The modality-specific
+contralateral difference feature image is calculated by transforming
+the image to the symmetric template, flipping the image contralaterally,
+then transforming it back using the inverse transform, and subtracting
+the corresponding template component from the original image.  Additionally, 
+the template difference image is calculated by subtracting the warped template 
+image from the corresponding modality of the subject.
+
+\paragraph{Miscellaneous Feature Images}
+Two additional feature images are calculated which seem to aid in 
+classification as measured by the random forest variable importance
+(relative importance plots are provided in the Results section).
+First, the binary mask image is used to calculate the distance image 
+\citep{maurer2003}.  Second, from the symmetric template deformation 
+mapping we calculate the log Jacobian image.  Related to the earlier
+distance feature image, we calculate the distance mask image of the symmetric
+template and then warp that distance image to the individual subject.
 
@@
 
-\subsubsection{Prediction:  Applying Random Forest Models}
+For use with the Challenge and Leaderboard data, cohort-specific models (both 
+low-grade and 
+high-grade glioma) for both GMM and MAP-MRF stages were created using only the 
+supplied training data.  Initial testing of our proposed framework was performed 
+on the training data using a leave-one-out strategy.  Once the
+feature images are created for each subject, the resulting images of the entire
+training cohort are organized in a csv file for input into the R script
+{\tt createModel.R}.  Other possible input parameters include the requested 
+number of trees, number of samples per label, and number of threads for parallel
+processing.  The output is an {\tt RData} file describing the random forest
+model which can be used for future predictions.
+
+Additionally, the {\tt randomForest} package provides a couple measurements 
+for determining the importance of chosen features when constructing the model.  
+This aids in potential feature pruning or intuiting model behavior.  As mentioned
+earlier, we provide the R script {\tt plotVariableImportance.R} to render
+one such quantity denoted as {\tt MeanDecreaseAccuracy}.  During model construction
+(specifically the out-of-bag error calculation stage), the decrease in prediction accuracy
+with the omission of a single feature or variable is tracked and averaged.  Thus,
+those features which have the greatest decrease in mean accuracy are considered
+to be the most discriminative.%
+\footnote{
+There is an issue with consistently ranking correlated predictors as described at {\tt http://www.r-bloggers.com/random-forest-variable-importance/} since the permutation testing performed for predictive accuracy assessment assumes predictor independence.
+Correctives have been proposed but we ignore these issues for this particular application.
+}
+
+\subsubsection{Prediction:  Applying the Random Forest Models}
 
@@
   \begin{tabular}{c}
-    \includegraphics[width=85mm]{Figures/pipeline.pdf}
+    \includegraphics[width=90mm]{Figures/pipeline.pdf}
   \end{tabular}
-  \caption{Diagrammatic workflow for the proposed RF model prediction.  The feature
+  \caption{Diagrammatic workflow for the proposed random forest model prediction.  The feature
   images are first generated from the input MRI and symmetric multivariate template. 
-  After organization of the feature images in a csv file, the stage 1 RF model is 
+  After organization of the feature images in a csv file, the stage 1 random forest model is 
   applied which results in a first stage classification voting output.  This output
@@
 
-\begin{figure*}
-  \centering
-  \begin{tabular}{cc}
-  \includegraphics[width=80mm]{Figures/BRATS_HG_GMM.pdf} &
-  \includegraphics[width=80mm]{Figures/BRATS_HG_MAP_MRF.pdf} \\
-  \includegraphics[width=80mm]{Figures/BRATS_LG_GMM.pdf} &
-  \includegraphics[width=80mm]{Figures/BRATS_LG_MAP_MRF.pdf}
-  \end{tabular}
-  \caption{Feature images from the BRATS\_HG0004 data set.}
-\end{figure*}
+Once the models are created, classification of tumors in new subjects is performed
+as illustrated in Figure \ref{fig:pipeline}.  From the feature images and input 
+GMM model, a tentative set of random forest voting output confidence images are produced.
+As described, this is used as input to the second prediction round.  The 
+final probability output images are used to produce the maximum labeling.
+
+A final round of binary morphological operations were heuristically designed
+to improve the final segmentation results such as removal of small connected 
+components and morphological closure of certain regions.
+All these steps are included in the perl script 
+{\tt applyTumorSegmentationModelForCohort.pl} designed for parallel subject 
+processing on the computational cluster at the University of Virginia.%
+\footnote{
+http://www.uvacse.edu
+}
 
@@
 \begin{table*}
-\caption{Dice scores from the MICCAI 2012 BRATs Study}
+\caption{Results from the MICCAI 2013 BRATS Challenge}
+\label{table:results}
 \begin{center}
-\begin{tabular*}{0.975\textwidth}{@{\extracolsep{\fill} } c c c c c c c c c}
+\begin{tabular*}{0.99\textwidth}{@{\extracolsep{\fill} } c c c c c c c c c c c}
 \toprule
-{} & \multicolumn{2}{c}{High-grade (real)} & \multicolumn{2}{c}{Low-grade (real)} & \multicolumn{2}{c}{High-grade (simulated)} & \multicolumn{2}{c}{Low-grade (simulated)}\\
-{\bf Method} & Edema & Tumor & Edema & Tumor & Edema & Tumor & Edema & Tumor\\
+{\bf Data set} & {\bf Rank} & \multicolumn{3}{c}{\bf Dice} & \multicolumn{3}{c}{ \bf Positive Predictive Value} & \multicolumn{3}{c}{ \bf Sensitivity} \\
+{} & {} & complete & core & enhanced & complete & core & enhanced & complete & core & enhanced \\
 \midrule
-\cite{zikic2012} & {$0.70 \pm 0.09$} & {$0.71 \pm 0.24$} & {$0.44 \pm 0.18$} & {$0.62 \pm 0.27$} & {$0.65 \pm 0.27$} & {$0.90 \pm 0.05$} & {$0.55 \pm 0.23$} & {$0.71 \pm 0.20$} \\
-\cite{bauer2012} & {$0.61 \pm 0.15$} & {$0.62 \pm 0.27$} & {$0.35 \pm 0.18$} & {$0.49 \pm 0.26$} & {$0.68 \pm 0.26$} & {$0.90 \pm 0.06$} & {$0.57 \pm 0.24$} & {$0.74 \pm 0.10$} \\
-ANTsR & {$0.65 \pm 0.15$} & {$0.66 \pm 0.28$} & {$0.49 \pm 0.16$} & {$0.65 \pm 0.21$} & {$0.68 \pm 0.25$} & {$0.91 \pm 0.08$} & {$0.61 \pm 0.25$} & {$0.84 \pm 0.09$} \\
-w/ Atropos & {$0.68 \pm 0.15$} & {$0.67 \pm 0.30$} & {$0.50 \pm 0.15$} & {$0.67 \pm 0.23$} & {$0.74 \pm 0.26$} & {$0.92 \pm 0.09$} & {$0.65 \pm 0.26$} & {$0.84\pm 0.08$} \\
+Challenge & 1 & 0.87 (1) & 0.78 (1) & 0.74 (1) & 0.85 (2) & 0.74 (5) & 0.69 (4) & 0.89 (2) & 0.88 (1) & 0.83 (1) \\
+Leaderboard & 1 & 0.79 (2) & 0.65 (1) & 0.53 (3) & 0.83 (1) & 0.70 (2) & 0.51 (3) & 0.81 (4) & 0.73 (2) & 0.66 (2) \\
+Evaluation & 2 & 0.88 (2) & 0.76 (2) & 0.55 (3) & 0.88 (3) & 0.80 (5) & 0.65 (3) & 0.89 (3) & 0.79 (3) & 0.53 (3) \\
 \bottomrule
@@
 
+A total of four random forest models were created from the 30 training data 
+sets. Two models for Stage 1 and Stage 2 processing were generated from the 
+20 high-grade glioma evaluation data described earlier.  Similarly, two
+additional models were created from the 10 low-grade glioma data sets.  Following
+model construction, weighted importance feature plots described earlier were 
+rendered to provide feedback as to the individual potential predictive accuracy
+of each feature.  The high-grade plots are given in Figure \ref{fig:hgimportance}
+whereas the low-grade plots are given in Figure \ref{fig:lgimportance}.
+
+\begin{figure*}
+  \makebox[\textwidth][c]{
+  \begin{tabular}{cc}
+  \includegraphics[width=90mm]{Figures/BRATS_HG_GMM.pdf} &
+  \includegraphics[width=90mm]{Figures/BRATS_HG_MAP_MRF.pdf} \\
+  \end{tabular}
+  }
+  \caption{{\tt MeanDecreaseAccuracy} plots generated from the high-grade glioma
+  Stage 1 and Stage 2 random forest models.  These plots provide a weighted 
+  ranking describing the importance of each feature for predictive accuracy. 
+  Features are listed from most to least important.
+  }
+  \label{fig:hgimportance}
+\end{figure*}
+
+\begin{figure*}
+  \makebox[\textwidth][c]{
+  \begin{tabular}{cc}
+  \includegraphics[width=90mm]{Figures/BRATS_LG_GMM.pdf} &
+  \includegraphics[width=90mm]{Figures/BRATS_LG_MAP_MRF.pdf} \\
+  \end{tabular}
+  }
+  \caption{{\tt MeanDecreaseAccuracy} plots generated from the low-grade glioma
+  Stage 1 and Stage 2 random forest models.  These plots provide a weighted 
+  ranking describing the importance of each feature for predictive accuracy.
+  Features are listed from most to least important.
+  }
+  \label{fig:lgimportance}
+\end{figure*}
+
+Immediately apparent from these plots are the importance of certain features.
+In general, the features based on the symmetric template are quite
+discriminative thus justifying the increased computational 
+resources required to generate such features.  For single-threaded processing
+on the cluster, creating the feature images for a single subject required 
+approximately 2 hours of total processing time of which the template registration 
+component took approximately 75\%.  Additionally, the first order 
+statistical features also seemed fairly important which is something that
+previous work had demonstrated \citep[e.g.][]{bauer2012,zikic2012}.
+There are a number of differences, however, between low-grade and high-grade 
+importance outcomes which would also seem to justify the creation of separate
+glioma class models.
+
+As described earlier, the MICCAI 2013 BRATS challenge data was provided to the
+competitors in three sets.  The Evaluation set was used primarily for training
+although an initial ranking was performed for all the competitors to aid in
+determining participation in the workshop.  Shortly prior to the actual workshop
+date, the Leaderboard data was released to the competitors for posting of results
+and subsequent ranking.  Finally, the night before the competition, the Challenge
+data was made available and used to produce the final competitor ranking.  In addition
+to the Dice overlap measure, additional performance emeasures for producing the rankings 
+included the positive predictive value, and sensitivity (all of which can be calculated
+using open source tools such as \cite{tustison2009}).  
+For all three data sets, we provide these overlap measures and relative competitor 
+ranking in Table \ref{table:results}.  Full competition results can be viewed
+at {\tt http://www.virtualskeleton.ch}.
+
 \section{Discussion and Conclusions}
+
+Although previous research has employed random forests for supervised brain
+segmentation, our contribution of concatenated model application and use
+of symmetric multivariate templates demonstrated good performance 
+in the recent MICCAI 2013 Brain Tumor Segmentation challenge/workshop.  
+In terms of variable importance, the latter provided several highly 
+discriminative features which resulted in the top-performing algorithm 
+of the competition.  This confirmed what others have found in that
+random forests provide an excellent framework for prediction in certain
+medical image analysis problems.
+
+Although these competitions are extremely useful as they provide objective
+assessment of algorithmic performance given the prevalence of selection bias
+in reporting results in conventional publication venues, even more important
+is the availing of the actual algorithmic instantiation (i.e. code) so that
+others can more easily build upon and utilize what has proven effective. 
+This also provides an opportunity for the users to return
+constructive feedback to the authors thereby improving the original offering.
+
+However, as explained previously, the brain tumor segmentation 
+methodology that we have made available is only a small part of the 
+larger software package that we have created in ANTsR.  Not only does
+ANTsR significantly facilitate the development of the work discussed,
+but it provides an interface to one of the most powerful statistical
+packages available in R.  The combination of the well-known ANTs software
+package with R provides tremendous potential for future insightful analyses 
+and graphical displays.
 

### 2013-11-06 Scripts/plotVariableImportance.R

@@ -1,0 +1,56 @@
+library( ggplot2 )
+library( grid )
+library( randomForest )
+
+stopQuietly <- function(...)
+  {
+  blankMsg <- sprintf( "\r%s\r", paste( rep(" ", getOption( "width" ) - 1L ), collapse = " ") );
+  stop( simpleError( blankMsg ) );
+  } # stopQuietly()
+
+args <- commandArgs( trailingOnly = TRUE )
+
+###############################################
+#
+# Selected parameters
+#
+###############################################
+
+if( length( args ) < 2 )
+  {
+  cat( "Usage: Rscript plotVariableImportance.R inputModel outputFile <title>", sep = "" )
+  stopQuietly()
+  }
+
+inputModel <- args[1]
+outputFile <- args[2]
+plotTitle <- ''
+if( length( args ) > 2 )
+  {
+  plotTitle <- args[3]
+  }
+
+load( inputModel )
+
+forestImp <- importance( modelForest, type = 1 )
+forestImp.df <- data.frame( Statistic = names( forestImp[,1] ), Importance = as.numeric( forestImp[,1] )  )
+forestImp.df <- forestImp.df[order( forestImp.df$Importance ),]
+
+forestImp.df$Statistic <- factor( x = forestImp.df$Statistic, levels = forestImp.df$Statistic )
+
+vPlot <- ggplot( data = forestImp.df, aes( x = Importance, y = Statistic ) ) +
+         geom_point( aes( color = Importance ) ) +
+         labs( title = plotTitle ) +
+         ylab( "" ) +
+         xlab( "MeanDecreaseAccuracy" ) +
+         scale_x_continuous( limits = c( 0, 110 ) ) +
+         scale_color_continuous( low = "navyblue", high = "darkred" ) +
+         theme( axis.text.y = element_text( size = 5 ) ) +
+         theme( plot.margin = unit( c( 0.1, 0.1, 0.1, -0.5 ), "cm" ) ) +
+         theme( axis.title = element_text( size = 10 ) ) +
+         theme( legend.position = "none" )
+
+ggsave( file = outputFile, plot = vPlot, width = 4, height = 8 )
+
+
+

### 2013-11-06 SimpleExample/examplePredict.sh

@@ -1,0 +1,125 @@
+#################################
+#
+# examplePredict - use after calling "exampleTrain.sh"
+
+BASE_DIRECTORY=./
+
+SCRIPTS_DIRECTORY=${BASE_DIRECTORY}/../Scripts/
+
+DATA_DIRECTORY=${BASE_DIRECTORY}/BRATS2DData/
+TESTING_DATA_DIRECTORY=${DATA_DIRECTORY}/TestingData/
+PROCESSED_DIRECTORY=${DATA_DIRECTORY}/Processed
+
+TEMPLATE_DIRECTORY=${DATA_DIRECTORY}/SymmetricTemplate/
+TEMPLATE_FLAIR=${TEMPLATE_DIRECTORY}/S_templateFLAIR_RESCALED_slice141.nii.gz
+TEMPLATE_T1=${TEMPLATE_DIRECTORY}/S_templateT1_RESCALED_slice141.nii.gz
+TEMPLATE_T2=${TEMPLATE_DIRECTORY}/S_templateT2_RESCALED_slice141.nii.gz
+
+TESTING_SUBJECTS=( BRATS_HG0006 BRATS_HG0008 BRATS_HG0009 )
+TESTING_SLICES=( 103 96 110 )
+
+NUMBER_OF_CLASSES=7
+TUMOR_CORE_LABEL=4
+
+GMM_MODEL=${PROCESSED_DIRECTORY}/example_GMM_Model
+MAP_MRF_MODEL=${PROCESSED_DIRECTORY}/example_MAP_MRF_Model
+
+######################################################################################
+#
+# Apply "GMM" and then "MAP_MRF" RF models to each testing subject.
+#
+######################################################################################
+
+for (( i=0; i<${#TESTING_SUBJECTS[@]}; i++ ))
+  do
+    echo "Applying GMM model to ${TESTING_SUBJECTS[$i]}"
+
+    MASK=${TESTING_DATA_DIRECTORY}/${TESTING_SUBJECTS[$i]}/${TESTING_SUBJECTS[$i]}_MASK_slice${TESTING_SLICES[$i]}.nii.gz
+    FLAIR=${TESTING_DATA_DIRECTORY}/${TESTING_SUBJECTS[$i]}/${TESTING_SUBJECTS[$i]}_FLAIR_slice${TESTING_SLICES[$i]}.nii.gz
[…]

### 2013-11-06 paper.tex
ENH:  initial draft?

@@
 
-\title{Multivariate Neuroanalysis with ANTsR: Application to Supervised Brain Tumor Segmentation using Random Forests}
+\title{Multivariate Neuroimage Analysis with ANTsR: Application to Supervised Brain Tumor Segmentation using Random Forests}
 
@@
 of high-performance registration algorithms for neuroimage analysis
-\citep{avants2008a} built upon the mature and well-vetted Insight Toolkit (ITK)
+\citep{avants2008a} built upon the mature Insight Toolkit (ITK)
 of the National Institutes of Health.  Since then, ANTs has grown to include 
@@
 
-In addition to describing ANTsR basics and how it can be
-used generally in multivariate neuroimaging studies, we showcase its use in
+%In addition to describing ANTsR basics and how it can be
+%used generally in multivariate neuroimaging studies, 
+We showcase ANTsR with
 the particularly difficult application of supervised brain tumor
-segmentation \cite{bauer2013} for which we use the widely popular 
+segmentation \citep{bauer2013} for which we use the increasingly popular 
 random forests machine learning technique.  Although random forests have
@@
 of a fully functional, parallelizable workflow for such a task and 
-that the performance of our proposed framework has proven extremely in 
+that the performance of our proposed framework has proven extremely effective in 
 actual practice, specifically
@@
 ANTsR package.
-The R project for statistical computing,%
-\footnote{
-www.r-project.org
-}
- or more compactly
-`R', is an environment for statistical computation
-and data visualization.  R's open source code base
+%The R project for statistical computing,%
+%\footnote{
+%www.r-project.org
+%}
+% or more compactly
+%`R', is an environment for statistical computation
+%and data visualization.  
+R's open source code base
 and add-on packages coupled with its rapidly growing 
@@
 scripts (R, shell, and perl) used to coordinate the ANTsR processing 
-including (along with a brief description of each script):
+including:
 \begin{itemize}
@@
   a set of co-registered input MRI from a single subject. 
-  \item {\tt createModel.R}:  creates random forest model given the input csv 
+  \item {\tt createModel.R}:  creates a random forest model given the input csv 
   file of the feature image file names for all training data and set of truth
@@
   \item {\tt plotVariableImportance.R}:  produces a plot of the importance 
-  (`{\tt MeanDecreaseAccuracy}') of each feature variable used 
+  ({\tt MeanDecreaseAccuracy}) of each feature variable used 
   in constructing the model.
@@
 }
-organized by K. Farahani, M. Reyes,B. Menze, E. Gerstner, J. Kirby and J. Kalpathy-Cramer. The challenge database contains fully anonymized images from the following institutions: 
+organized by K. Farahani, M. Reyes, B. Menze, E. Gerstner, J. Kirby and J. Kalpathy-Cramer. 
+The challenge database contains fully anonymized images from the following institutions: 
 ETH Zurich, University of Bern, University of Debrecen, and University of Utah and 
@@
 \paragraph{Miscellaneous Feature Images}
-Two additional feature images are calculated which seem to aid in 
+Three additional feature images are calculated which seem to aid in 
 classification as measured by the random forest variable importance
@@
 distance feature image, we calculate the distance mask image of the symmetric
-template and then warp that distance image to the individual subject.
+template and then warp that distance image to the individual subject
+for a third feature image.
 
@@
 high-grade glioma) for both GMM and MAP-MRF stages were created using only the 
-supplied training data.  Initial testing of our proposed framework was performed 
+supplied training data.  Prior to training, we segmented normal brain tissue \cite{avants2011}
+for each data set by segmenting only the T1 image.  This was only to yield
+a rough estimate of normal brain tissue to augment the already provided 
+pathology labels.  This resulted in seven labels (csf, gray matter, white matter,
+necrosis, edema, enhancing, and non-enhancing tumor)
+characterizing each brain.
+
+Initial testing of our proposed framework was performed 
 on the training data using a leave-one-out strategy.  Once the
@@
 
-Additionally, the {\tt randomForest} package provides a couple measurements 
+Additionally, the {\tt randomForest} package provides  measurements 
 for determining the importance of chosen features when constructing the model.  
@@
 packages available in R.  The combination of the well-known ANTs software
-package with R provides tremendous potential for future insightful analyses 
-and graphical displays.
+package with R provides tremendous potential for future insightful analysis.
 

### 2014-01-17 paper.tex
WIP: about 1/2 way through

@@
 
-\usepackage{color}
+\usepackage{color,hyperref}
 \usepackage{multirow,booktabs,ctable,array}
@@
 
-\title{Multivariate Neuroimage Analysis with ANTsR: Application to Supervised Brain Tumor Segmentation using Random Forests}
+\title{Multivariate Neuroimage Analysis with \textit{ANTsR}: Application to Supervised Brain Tumor Segmentation using Random Forests}
 
@@
 Improvements in image acquisition and increasingly sophisticated statistical techniques 
-for neuroanalysis have motivated the development of advanced computational software 
-packages for studying the
-brain---many of which
-have been made publicly available.  One such popular toolkit is the open source Advanced
+for medical image analysis have motivated the development of advanced computational software 
+packages, many of which
+have been made publicly available.  One such popular toolkit is the
+free and open source Advanced
 Normalization Tools (ANTs) which contains a suite of well-vetted processing tools for
@@
 inhomogeneity field correction, and template building.  Despite the numerous
-solutions afforded by such packages, neuroscience (and data analysis in general) requires
+solutions afforded by medical image analysis software, neuroscience (and data analysis in general) requires
 robust, statistical tools for making predictive inferences with respect to given hypotheses
@@
 In this paper, we describe the
-coupling of ANTs with the widely-used R language---an open source environment for 
-statistical processing and data visualization---which we denote as ANTsR.  One of the
+coupling of ANTs with the widely-used \textit{R} language---an open source environment for 
+statistical processing and data visualization---which we denote as \textit{ANTsR}.  One of the
 many benefits from such integration is the accessibility to advanced statistical 
 and machine learning techniques as well as state-of-the-art visualization possibilities.  
-To showcase the flexibility and utility of ANTsR, we 
+To showcase the flexibility and utility of \textit{ANTsR}, we 
 apply it to the difficult problem of brain tumor segmentation from multi-modality
@@
 \begin{keyword}
-advanced normalization tools \sep BRATS \sep brain tumor segmentation \sep R project \sep random forests
+advanced normalization tools \sep BRATS \sep brain tumor segmentation \sep \textit{R} project \sep random forests
 %% keywords here, in the form: keyword \sep keyword
@@
 
-ANTs (Advanced Normalization Tools) originated with the open source availing
-of high-performance registration algorithms for neuroimage analysis
-\citep{avants2008a} built upon the mature Insight Toolkit (ITK)
-of the National Institutes of Health.  Since then, ANTs has grown to include 
-several algorithmic solutions for robust medical image analysis including bias 
+ANTs (Advanced Normalization Tools) was built, originally, to provide 
+high performance image registration for medical image analysis
+\citep{avants2008a} and based upon the mature Insight Toolkit (ITK)
+sponsored by the National Institutes of Health.  Since then, ANTs has grown to include 
+several robust medical image analysis solutions including bias 
 correction \citep{tustison2010}, $n$-tissue multivariate segmentation 
@@
 that robust statistical machinery was lacking for making inferences regarding
-the data produced during the course of ANTs processing.  ANTsR was developed
+the data produced during the course of ANTs processing.  \textit{ANTsR} was developed
 specifically to provide an interface between ANTs, a 
 powerful neuroimaging toolkit for producing reliable imaging data 
-transformations, and the R project%
+transformations, and the \textit{R} project%
 \footnote{
@@
 for statistical computing and visualization thus providing a complete
-set of tools for multivariate neuroimage analysis.
-
-%In addition to describing ANTsR basics and how it can be
+set of tools for multivariate neuroimage analysis \textit{ANTsR} intends to provide a modern framework for medical analytics, with a focus on imaging-assisted prediction and statistical power.
+
+%In addition to describing \textit{ANTsR} basics and how it can be
 %used generally in multivariate neuroimaging studies, 
-We showcase ANTsR with
+We showcase \textit{ANTsR} with
 the particularly difficult application of supervised brain tumor
@@
 supervised brain segmentation (e.g. \cite{geremia2011}), 
-we demonstrate that ANTsR significantly facilitates the construction 
+we demonstrate that \textit{ANTsR} significantly facilitates the construction 
 of a fully functional, parallelizable workflow for such a task and 
@@
 
-\subsection{ANTsR:  An ANTs/R Interface}
+\subsection{\textit{ANTsR}:  An ANTs/\textit{R} Interface}
 
@@
 
-Careful consideration of available options
-led to the adoption of R to complement ANTs capabilities resulting in the
-ANTsR package.
+Careful consideration of available statistical software 
+led to the adoption of \textit{R} to complement ANTs quantification resulting in the
+\textit{ANTsR} package.
 %The R project for statistical computing,%
@@
 %and data visualization.  
-R's open source code base
+\textit{R}'s open source code base, reliable software testing and distribution strategies,
 and add-on packages coupled with its rapidly growing 
@@
 
-The ANTsR package is publicly available on the github project hosting service.%
-\footnote{
-https://github.com/stnava/ANTsR
-}
-Prior to installation of ANTsR, several external R packages
+The \textit{ANTsR} package is publicly available on the github project hosting service.%
+\footnote{
+\href{https://github.com/stnava/ANTsR}{https://github.com/stnava/ANTsR}
+}
+Prior to installation of \textit{ANTsR}, several external R packages
 need to be installed including: \verb#Rcpp#, \verb#signal#, \verb#timeSeries#, 
 \verb#mFilter#, \verb#doParallel#, \verb#robust#, \verb#magic#, \verb#knitr#, \verb#pixmap#, 
-\verb#rgl#, and \verb#misc3d#.  Additionally, in order
+\verb#rgl#, and \verb#misc3d#.  
+See \href{http://stnava.github.io/software/2014/01/08/antsr/}{http://stnava.github.io/software/2014/01/08/antsr/} for current status. 
+Additionally, in order
 to perform the supervised brain segmentation as described 
@@
 \footnote{
-Packages are easily installed using the {\tt install.packages()} R mechanism.
-}
-
-In addition to R and the add-on packages previously mentioned, CMake is also 
+Packages are easily installed using the {\tt install.packages()} \textit{R} mechanism.
+}
+
+In addition to \textit{R} and the add-on packages previously mentioned, CMake is also 
 required.  CMake%
@@
 and ANTs.  Further instructions for download and
-installation can be found on the ANTsR github website.
+installation can be found on the \textit{ANTsR} github website.  Feel
+free to contact the authors if installation trouble occurs.  We note
+that \textit{ANTsR} is currently only tested on UNIX-alikes such OSX and Ubuntu
+operating systems.  \footnote{Windows installation should be possible
+  but, to our knowledge, has not been attempted.}.
 
 \subsubsection{Usage}
-The utility of ANTsR stems from the ability to couple ANTs core 
+\textit{ANTsR} is intended to not only allow easy interchange between
+medical imaging formats and \textit{R} but also to facilitate
+reproducible scientific studies and the type compilable analysis
+articles that are fundamental to journals such as
+\textit{Biostatistics}.  Both \verb#knitr# and \verb#sweave#
+facilitate integration of \textit{R}-code with the LaTeX document
+preparation system.
+
+An additional motivation for our development of \textit{ANTsR} (and
+hopefully its acceptance by the community) 
+stems from the ability to couple ANTs core 
 functionality, including IO tools such as \verb#antsImageRead#, 
-with the large number of R statistical and
+with the large number of \textit{R} statistical and
 visualization packages.  Due to this combination, several
@@
 constituent functions are invoked in the similar manner as other
-R libraries.
+\textit{R} libraries.
 
 As mentioned earlier, we have made this entire framework
-available as open source.  In addition to the ANTsR repository
-already on github which houses both ANTs and ANTsR functionality, 
+available as open source.  In addition to the \textit{ANTsR} repository
+already on github which houses both ANTs and \textit{ANTsR} functionality, 
 we created a special github repository specifically for this work
@@
 \footnote{
-https://github.com/ntustison/ANTsAndArboles
+\href{https://github.com/ntustison/ANTsAndArboles}{https://github.com/ntustison/ANTsAndArboles}
 }
 Also, we posted all
-scripts (R, shell, and perl) used to coordinate the ANTsR processing 
+scripts (\textit{R}, shell, and perl) used to coordinate the \textit{ANTsR} processing 
 including:
@@
 
+\textcolor{red}{FIXME---what is the first thing a user should do?  is
+  there a basic test verifying the 2D result?}
+
 \subsection{Supervised Brain Tumor Segmentation with Random Forests}
@@
 \footnote{
-One of the principal advantages of R is the extensive community of
+One of the principal advantages of \textit{R} is the extensive community of
 developers  who have contributed on the order of thousands of packages 
-extending R's capabilities beyond its core functionality.
+extending \textit{R} 's capabilities beyond its core functionality.
 Most relevant 
@@
 feature images are created for each subject, the resulting images of the entire
-training cohort are organized in a csv file for input into the R script
+training cohort are organized in a csv file for input into the \textit{R} script
 {\tt createModel.R}.  Other possible input parameters include the requested 
@@
 This aids in potential feature pruning or intuiting model behavior.  As mentioned
-earlier, we provide the R script {\tt plotVariableImportance.R} to render
+earlier, we provide the \textit{R} script {\tt plotVariableImportance.R} to render
 one such quantity denoted as {\tt MeanDecreaseAccuracy}.  During model construction
@@
 methodology that we have made available is only a small part of the 
-larger software package that we have created in ANTsR.  Not only does
-ANTsR significantly facilitate the development of the work discussed,
+larger software package that we have created in \textit{ANTsR}.  Not only does
+\textit{ANTsR} significantly facilitate the development of the work discussed,
 but it provides an interface to one of the most powerful statistical
-packages available in R.  The combination of the well-known ANTs software
-package with R provides tremendous potential for future insightful analysis.
+packages available in \textit{R}.  The combination of the well-known ANTs software
+package with \textit{R} provides tremendous potential for future insightful analysis.
 

### 2014-01-17 paper.tex
WIP: BA comments

@@
 One of the innovations that we provide in this work is
-the use of concatenated random forests for improved probabilistic estimation
+the use of \textit{concatenated random forests} \textcolor{red}{where
+  does the name come from?  it doesnt get mentioned again until the
+  discussion section ... expanding on this / defining it could help
+  ``novelty'' angle in neuroimage submission .... FIXME} for improved probabilistic estimation
 of the brain segmentation labels.  Briefly, Gaussian mixture
@@
 challenge data, such images have shown discriminative potential 
-\citep[e.g.][]{price2003} to warrant investigation in our future work.
+\citep[e.g.][]{price2003} to warrant investigation in our future
+work.  The general procedure for const ructing a symmetric template involves flipping data around the mid-sagittal
+plane, as outlined in prior work
+\citep{Hopkins2013}. \textcolor{red}{FIXME-nick, is that correct? is
+  this covered in the scripts?  if so, where?}
 
@@
 to isolate the brain \cite{avants2010a}.  We recently performed an evaluation of over 1200
-MRI which demonstrated perfect performance \cite{tustison2013a}.
-
-\paragraph{Stage 1 (GMM)}
+MRI which demonstrated performance that is comparable or better than
+other available methodology \cite{tustison2013a}.
+
+\paragraph{Stage 1 (GMM)} \textcolor{red}{Need more descriptive
+  section name ... lots going on in the paper , easy to get disoriented}
 Prior cluster centers for specific tissue types learned from training data \citep{reynolds2009} are used in the GMM random forest stage to create 
-multiple feature images.  
+multiple feature images.  \textcolor{red}{FIXME I think this should be
+  expanded upon a bit ... can you be more specific?  how are these
+  cluster centers passed to the RF?  why are these centers portable to
+  this new data? etcetera.}
 Given $M$ tissue types, a GMM formulates the 
@@
 estimation which can performed using the Atropos segmentation 
-tool \citep{avants2011} available in ANTs.
+tool \citep{avants2011} available in ANTs.  \textcolor{red}{FIXME
+  would prefer active voice version of the sentence to make clear that
+  Atropos uses EM} See Figure~\ref{fig:featureImages} for
+an example segmentation.
+
+\textcolor{red}{should there be an RF model equation here?  i.e. what
+  exactly (in R) was used ?}
 
@@
 forest voting classification maps giving a probabilistic estimate
-of the label value at each voxel.  These spatial probability
+of the label value at each voxel.  \textcolor{red}{FIXME ---i think
+  this sentence is confusing ... did you already run the RF at this
+  point?  seems no ... but ``results in a set of RF voting
+  classification maps'' sounds like you already have them.}  These spatial probability
 maps can be used as spatial priors in a MAP-MRF segmentation
@@
 forest models improves the resulting labeled output.
+
+\textcolor{red}{ok ... but i still don't understand this ... it gives
+  me a vague idea ... can you sketch out the equations for these
+  features/predictors? and give better intuition of what the
+  ``composite'' RF (previously referred to as concatenated?) is?}
 
@@
 }
+
+\textcolor{red}{do you actually use variable importance or no?
+  i.e. to trim the number of features?  overall, having a hard time
+  seeing what is possible versus what was actually done ... }
 
@@
 
+\textcolor{red}{how novel is this a/symmetry stuff?  i have the chimp
+  paper that briefly mentions it but could this be spun as a
+  ``contribution'' (with additional technical detail) in addition to pointing out the benefit of asymmetry
+  plus RF's?  also see
+  \href{http://stnava.github.io/asymmetry/}{http://stnava.github.io/asymmetry/}
+  which is sort of novel ... }
+
 As described earlier, the MICCAI 2013 BRATS challenge data was provided to the
@@
 Although previous research has employed random forests for supervised brain
-segmentation, our contribution of concatenated model application and use
+segmentation, our contribution of \textcolor{red}{concatenated model} application and use
 of symmetric multivariate templates demonstrated good performance 
@@
 random forests provide an excellent framework for prediction in certain
-medical image analysis problems.
+medical image analysis problems.  Even with relatively few training
+subjects relative to input variables, the random forest models
+perform generalizably.
+In addition, this work highlights the value of  a symmetric multiple
+modality template for clinical disorders in which asymmetry is a
+hallmark.  Furthermore, our challenge-leading results establish the
+value of such templates in multiple modality prediction
+tasks in which there is a small training set with an abundance of
+multivariate data.
+
+\textcolor{red}{\subsection{Discussion of clinical relevance of the multivariate
+  features we ``found''}}
 

### 2014-02-07 paper.tex
ENH:  Major revisions.

@@
 
-\title{Multivariate Neuroimage Analysis with \textit{ANTsR}: Application to Supervised Brain Tumor Segmentation using Random Forests}
+%\title{Multivariate Neuroimage Analysis with \textit{ANTsR}: Application to Supervised Brain Tumor Segmentation using Random Forests}
+\title{Optimal Symmetric Multimodal Templates and Concatenated Random Forests for Supervised Brain Tumor Segmentation (made easy) with \textit{ANTsR}}
 
@@
 \author[label1]{Christopher R.~Durst}
-\author[label2]{Jeffrey T.~Duda}
 \author[label1]{James C.~Gee}
@@
 \begin{abstract} 
-Improvements in image acquisition and increasingly sophisticated statistical techniques 
-for medical image analysis have motivated the development of advanced computational software 
-packages, many of which
-have been made publicly available.  One such popular toolkit is the
-free and open source Advanced
-Normalization Tools (ANTs) which contains a suite of well-vetted processing tools for
-core data transformation tasks such as image registration, image segmentation, 
-inhomogeneity field correction, and template building.  Despite the numerous
-solutions afforded by medical image analysis software, neuroscience (and data analysis in general) requires
-robust, statistical tools for making predictive inferences with respect to given hypotheses
-and for rendering quantitative information in clarifying and insightful ways.  
-In this paper, we describe the
-coupling of ANTs with the widely-used \textit{R} language---an open source environment for 
-statistical processing and data visualization---which we denote as \textit{ANTsR}.  One of the
-many benefits from such integration is the accessibility to advanced statistical 
-and machine learning techniques as well as state-of-the-art visualization possibilities.  
-To showcase the flexibility and utility of \textit{ANTsR}, we 
-apply it to the difficult problem of brain tumor segmentation from multi-modality
-MRI.  The reported algorithmic framework was the top-performing entry in the MICCAI 2013 Multimodal 
-Brain Tumor Segmentation (BRATS) challenge and, to our knowledge, is the only entry for
-both challenge years (2012, 2013) which has been made publicly available as open source.
+Automated brain tumor segmentation in MRI is complicated by the lack of prior knowledge 
+concerning tumor location, spatial extent, shape, possible displacement of normal 
+tissue, and intensity signature, all of which limit 
+the applicability of well-established techniques that have demonstrated good performance 
+in related tasks such as normal brain tissue segmentation.  Due to this 
+limitation, we localize our prior knowledge strategy to each subject through the 
+use of subject-specific multimodal asymmetrical features which demonstrate excellent
+discriminative properties within the random forest machine learning framework.  
+This approach is made possible with symmetric multimodal brain templates which
+are optimal in terms of both shape and multivariate intensity.
+Further performance improvements are demonstrated by a successive application of random forest models for increasing levels of feature refinement
+ to which we refer as {\it concatenated random forests}.
+Although the derived workflow involves many disparate components, the newly developed \textit{ANTsR} open source software package significantly facilitates the proposed methodology.
+\textit{ANTsR}, a comprehensive interface of the popular Advanced Normalization Tools (ANTs) repository with the \textit{R} statistical project,
+integrates well-performing image analysis tools such as image registration, segmentation, bias correction, and template building with access to advanced statistical and machine learning techniques as well as state-of-the-art visualization possibilities.
+The reported algorithmic framework was the top-performing entry in the MICCAI 2013 Multimodal Brain Tumor Segmentation challenge and, to our knowledge, is the only entry for both challenge years (2012 \& 2013) which has been made publicly available.
 \end{abstract}
@@
 \begin{keyword}
-advanced normalization tools \sep BRATS \sep brain tumor segmentation \sep \textit{R} project \sep random forests
+advanced normalization tools \sep BRATS \sep brain tumor segmentation \sep \textit{R} project 
 %% keywords here, in the form: keyword \sep keyword
@@
 \section{Introduction}
+Given the complexity of tumor growth and appearance and the need
+for precise volumetric measurements for tumor characterization, 
+much research has been invested in computational approaches for 
+automated segmentation of tumor regions in MR images.  Several 
+approaches have been previously proposed in the literature as
+detailed in recent reviews \citep{angelini2007,bauer2013}.  
+Any type of commensurable algorithmic evaluation, however, is
+extremely problematic due to widely varying performance assessments
+described in the corresponding publications, lack of publicly available 
+evaluation data, and privately kept algorithmic instantiations.  
+In response to these issues, the Multimodal Brain Tumor Segmentation 
+(BRATS) challenge was initiated in 2012 (and continued in 2013) under 
+the auspices of the Medical Image Computing and Computer Assisted 
+Intervention Society in association with their annual international 
+conference \citep{menze2014}.
+
+After observing the 2012 challenge to gain insight for our own research 
+needs, we found that no teams had availed their software in any 
+form which prompted our participation in the challenge which took place
+the following year in Nagoya, Japan.  Heavily influenced by the success 
+of participants in the 2012 challenge (specifically, the work of 
+\cite{bauer2012,geremia2012,zikic2012}), we adopted the increasingly popular 
+random forest machine learning framework \citep{breiman2001} which 
+permits the inclusion of various features which might prove saliently 
+discriminative.  Instrumental to our success in the 2013 challenge 
+was the use of features based on multimodal shape and appearance 
+asymmetries which were generated via the construction of symmetric, 
+multivariate brain templates.  Additional improvements in performance
+were due to the use of a two-stage random forest model approach
+whereby the output probability images from the application of the first
+random forest model seeds the construction of a subset of new feature images
+for input into the second stage.  The results of the second random forest model
+are then refined using a series of heuristically-derived binary morphological
+operations.
+
+In line with our philosophy regarding open science \citep{tustison2013,ince2012}, 
+we made the code available as open source%
+\footnote{
+https://github.com/ntustison/BRATS2013
+}
+and, to our knowledge, we are the only competitors out of the 20 or so teams 
+that participated in either years' challenges to do this.  All code is based
+on the well-vetted and open source Insight Toolkit (ITK)%
+\footnote{
+http://www.itk.org/
+} 
+of the National Institutes of Health and is available through our \textit{ANTsR} 
+package.
+
+In subsequent sections, we describe in greater detail the automated 
+supervised brain segmentation pipeline including the asymmetry
+analysis and the concatenated random forest construction.  We also
+describe the code and data used for the BRATS challenge so that 
+the interested reader can reproduce our results.  Finally, we 
+illustrate performance with results from the BRATS 2013 challenge.
+
+% which has demonstrated good performance
+%in other neuro-related tasks \citep[e.g.,][]{geremia2011}.
+
+%Although random forests have been proposed previously in the 
+%literature for, among other things, supervised brain segmentation (e.g. 
+%\cite{geremia2011}), 
+%we demonstrate that \textit{ANTsR} significantly facilitates the construction 
+%of a fully functional, parallelizable workflow for such a task and 
+%that the performance of our proposed framework has proven extremely effective in 
+%actual practice, specifically
+% in the recent MICCAI 2013 Multimodal Brain Tumor Segmentation
+%Challenge.%
+%\footnote{
+%http://martinos.org/qtim/miccai2013/
+%}
+%\begin{figure}
+%  \centering
+%  \includegraphics[width=90mm]{Figures/award.pdf}
+%  \caption{First place certificate from the MICCAI 2013 BRATS challenge.}
+%  \label{fig:award}
+%\end{figure}
+
+\section{Materials and Methods}
+
+Certain key elements characterize the supervised brain tumor segmentation
+protocol including:
+\begin{itemize}
+  \item construction of symmetric multivariate templates for multimodal 
+        asymmetry-based features,
+  \item generation of other image-based features,
+  \item training of random forest models, and
+  \item prediction using the proposed framework.
+\end{itemize}
+A description of each of these components is detailed below along.
+
+\subsection{Multimodal Asymmetry Features from Multivariate Symmetric Templates}
+
+\setlength{\tabcolsep}{2pt}
+
+\begin{figure}[htb]
+  \centering
+  \begin{tabular}{cccc}
+  \includegraphics[width=20mm]{Figures/BRATS_HG0001/BRATS_HG0001_FLAIR_slice93.png} &
+  \includegraphics[width=20mm]{Figures/BRATS_HG0001/BRATS_HG0001_T1_slice93.png} &
+  \includegraphics[width=20mm]{Figures/BRATS_HG0001/BRATS_HG0001_T1C_slice93.png} &
+  \includegraphics[width=20mm]{Figures/BRATS_HG0001/BRATS_HG0001_T2_slice93.png} \\
+  (a) & (b) &
+  (c) & (d) \\
+  \end{tabular}
+  \caption{Induced bilateral asymmetry due to tumor presence causing
+  distortion of the plane of symmetry.  Shown are mid-axial slices of
+  one of the BRATS 2013 training data 
+  (specifically {\tt BRATS\_HG0001}):  (a) FLAIR, (b) T1, (c) T1C, and
+  (d) T2.  }
+  \label{fig:asymmetry}
+\end{figure}
+
+\setlength{\tabcolsep}{6pt}
+
+In order to better characterize deviations from normal brain shape 
+and appearance, several image features were derived using symmetric 
+population-specific multivariate templates \citep{avants2008,avants2010}.  
+For normal neuroanatomy, the use of spatial prior information 
+coupled with image normalization capabilities has proven useful 
+in producing improved segmentation results of ``expected'' brain tissue
+such as cerebrospinal fluid, gray matter, and white matter 
+\citep[e.g.,][]{ashburner1997}.  In contrast, accommodating spatial 
+priors to model the presence of a possible tumor and its constituent tissue 
+components is difficult. However, since the normal brain 
+exhibits a bilaterally symmetric organization, we can 
+use the presence of asymmetries to differentiate abnormal 
+brain tissue (i.e. tumor).  A similar motivation prompted
+the identification of the mid-sagittal plane of symmetry \citep{prima2002}
+for feature generation in MS lesion \citep{geremia2011} and 
+tumor \citep{geremia2012} identification.  However, this
+approach does not take into account the displacement of 
+normal tissue due to tumor growth causing the mid-sagittal plane
+to deform from its planar structure (cf Figure \ref{fig:asymmetry} (a)).
+
+To take into account these potential asymmetries, we 
+require a data set with the same modalities as dictated
+by the subject image acquisition protocol.  Although it
+would be preferable to build population-specific multivariate
+templates from normal data using the same acquisition 
+parameters, we substituted publicly available data
+
+A recent neuroimaging reproducibility study
+by Landman et al. resulted in an open data cohort of 21
+normal individuals, each imaged twice, comprising several
+modalities including ASL, FLAIR, DTI, fMRI, T1, and T2 
+\citep{landman2011}.  These data (known as the
+``MMRR'' data set) were selected for deriving
+a multivariate template due to its public availability and
+inclusion of several modalities (even permitting future 
+incorporation of modalities 
+not currently included with the BRATS challenge into our 
+segmentation framework).
+
+As detailed in \cite{avants2008,avants2010}, 
+given $K$ image modalities for $N$ subjects,  
+${\mathbf I} = \{I_1,I_2,\ldots, I_K\}$, multivariate 
+template construction iterates between optimizing the set 
+of diffeomorphic transforms between the subjects and the 
+template, 
+$\left\{\left(\phi_1,\phi_1^{-1}\right),\ldots,\left(\phi_N,\phi_N^{-1}\right)\right\}$ 
+and constructing the 
+optimal multivariate template appearance 
+$\mathbf{J}=\{J_1,J_2,\ldots, J_K\}$ to minimize the
+following cost function:
+\begin{align}
+  \sum_{n=1}^N 
+        \Bigg[ D &\left( \psi(\mathbf{x}),\phi_1^n(\mathbf{x},1)\right) \\ \nonumber 
+        +& \sum_{k=1}^K \lambda_k \Pi_k \left(I_k^n\left(\phi_n(\mathbf{x},0.5)\right),J_k\left(\phi^{-1}_n(\mathbf{x},0.5)\right)\right)\Bigg].
+\end{align}
+$D$ is the diffeomorphic shape distance,
+\begin{align}
+D\left( \phi( \mathbf{x},0),\phi( \mathbf{x},1)\right) = \int_0^1 \| \nu(\mathbf{x},t)\|_L dt
+\end{align}
+dependent on the choice of linear operator, $L$, and $\nu$
+is the velocity field
+\begin{align}
+\nu\left( \phi(\mathbf{x},t) \right) = \frac{d\phi(\mathbf{x},t)}{dt},\,\,\, \phi(\mathbf{x},0) = \mathbf{x}.
+\end{align}
+Each pairwise registration employing the similarity metric $\Pi_k$ can 
+be assigned a relative weighting, $\lambda_k$, to weight a particular
+modality's influence in the construction process.  Once the multivariate
+template has converged (typically four iterations are used), we symmetrize
+the template by flipping each image contralaterally and then running the
+multivariate template construction a second time using only the multivariate
+template and its symmetrical version.
+
+In terms of implementation, this template building algorithm is 
+encapsulated in the script \verb#antsMultivariateTemplateConstruction.sh#,
+available in the ANTs repository, which permits parallel processing on
+an individual workstation or on a large computational cluster.  In
+Figure \ref{fig:symmetrictemplates} we show mid-axial slices from
+the MMRR multivariate symmetrical template consisting of FA%
+\footnote{
+Although DTI-based images were not included in the
+challenge data, such images have shown discriminative potential 
+\citep[e.g.,][]{price2003} warranting investigation in our future
+work.  
+}, FLAIR,  
+T1, and T2 components.  These images have been made publicly available%
+\footnote{
+http://figshare.com/articles/Kirby\_multivariate\_template/852989.
+All 7 components are stored under the more informal ``Kirby'' moniker.
+}
+along with other processed images such as corresponding
+tissue priors and structural labels.
+
+{\color{red}{Need to describe the actual feature images}}
+
+\begin{figure}[htb]
+  \centering
+    \includegraphics[width=85mm]{Figures/templateKirby.pdf}
+  \caption{Multivariate symmetric template created from the MMRR 
+           data described in \cite{landman2011}.  Shown are the
+           (a) FA, (b) FLAIR, (c) MPRAGE, 
+           and (d) T2 template components.
+          }
+  \label{fig:symmetrictemplates}
+\end{figure}
+
+\begin{figure}[!htb]
+  \centering
+    \includegraphics[width=85mm]{Figures/asymmetryFeatureImages2.pdf}
+  \caption{ $S \underset{b}{\leftrightsquigarrow} \leftrightarrow T$
+          }
+  \label{fig:asymmetryFeatures}
+\end{figure}
+
+\subsection{Multi-Modal Feature Image Generation}
+
+%For both training and prediction protocols, feature images are generated
+%for each subject using the selected input MRI.  
+Key to any supervised regression or classification protocol are the 
+selected features for training and subsequent prediction.  Based on previous
+work and our own experience, the following feature images are generated
+for use in both training and prediction:
+\begin{itemize}
+  \item Per modality (FLAIR, T1, T1C, T2)
+    \begin{itemize}
+      \item First-order neighborhood statistical images:
+            mean, variance, skewness, and entropy. 
+            Neighborhood radius $\in \{1,3\}$.
+    \item GMM (stage 1) and MAP-MRF (stage 2) posteriors: CSF, gray matter, white 
+          matter, necrosis, edema, non-enhancing tumor and enhancing tumor (or a
+          subset for the simulated data).
+    \item GMM (stage 1) and MAP-MRF (stage 2) connected component geometry 
+          features:  distance to tumor core label, volume, volume to surface area ratio, eccentricity, and elongation
+    \item Template-based:  symmetric template difference and 
+          contralateral difference with Gaussian smoothing ($\sigma = 4$mm).
+    \end{itemize}
+  \item Miscellaneous: normalized Euclidean distance based on cerebral mask,
+    log Jacobian image, and  (T1 - T1C) difference image.
+\end{itemize}
+The generation of all feature images is performed using the available script 
+{\tt createFeatureImages.sh} which also preprocesses the input images performing N4
+bias correction \citep{tustison2010} and intensity rescaling to $[0,1]$.  Note that 
+it is easily expandable to include 
+additional modalities and/or neighborhood and smoothing parameters.
+In Figure \ref{fig:featureImages}
+we provide sample mid-axial slices from the BRATS\_HG0301
+data set from the Challenge cohort.
+
+We previously mentioned that all data was provided already 
+skull-stripped.  However, we note to the reader that an
+available ANTs-based solution to the brain extraction problem
+is found in the script {\tt antsBrainExtraction.sh} which uses
+a hybrid approach based on image registration and binary morphology
+to isolate the brain \cite{avants2010a}.  We recently performed an evaluation of over 1200
+MRI which demonstrated performance that is comparable or better than
+other available methodology \cite{tustison2013a}.
+
+\begin{figure*}[htb]
+  \centering
+  \makebox[\textwidth][c]{
+  \includegraphics[width=180mm]{Figures/featureImages.pdf}
+  }
+  \caption{Representative feature images derived from the 
+           BRATS\_HG0301  ``challenge'' data set.
+           Neighborhood statistical images for each modality were 
+           generated by calculating a given statistic within
+           a specified neighborhood radius.  Also calculated for each modality were feature
+           images based on either the GMM or the MAP-MRF segmentation.  For the former, we
+           show the probability maps for each of the seven labels which are used as feature
+           images.  From the resulting hard segmentation, we calculate various geometric 
+           measures per connected component of each of the seven labels.  Similarly, the 
+           registration to the symmetric template produces the modality-specific 
+           difference images with the
+           corresponding symmetric template itself and with respect to the 
+           contralateral side.   
+           This mapping is also used to produce the log Jacobian image.  Finally, the 
+           (T1 - TC) image is calculated and, from the
+           cerebral mask, we calculate the normalized distance image.  
+           }
+  \label{fig:featureImages}         
+\end{figure*}
+
+\paragraph{Stage 1 (GMM)} \textcolor{red}{Need more descriptive
+  section name ... lots going on in the paper , easy to get disoriented}
+Prior cluster centers for specific tissue types learned from training data \citep{reynolds2009} are used in the GMM random forest stage to create 
+multiple feature images.  \textcolor{red}{FIXME I think this should be
+  expanded upon a bit ... can you be more specific?  how are these
+  cluster centers passed to the RF?  why are these centers portable to
+  this new data? etcetera.}
+Given $M$ tissue types, a GMM formulates the 
+probability distribution at each voxel, $\mathbf{x}$, as the
+sum of Gaussian components, $\mathcal{N}(\mathbf{x}|\mu,\sigma)$, i.e.
+\begin{align}
+p\left(\mathbf{x}|\mu_m,\sigma_m,\lambda_m\right) = \sum_{i=1}^M \lambda_m \mathcal{N}(\mathbf{x}|\mu_m,\sigma_m)
+\end{align}
+where $\sum_{m=1}^M \lambda_m = 1$.  One popular method for 
+determining the parameters of the GMM is maximum likelihood 
+estimation which can performed using the Atropos segmentation 
+tool \citep{avants2011} available in ANTs.  \textcolor{red}{FIXME
+  would prefer active voice version of the sentence to make clear that
+  Atropos uses EM} See Figure~\ref{fig:featureImages} for
+an example segmentation.
+
+\textcolor{red}{should there be an RF model equation here?  i.e. what
+  exactly (in R) was used ?}
+
+\paragraph{Stage 2 (MAP-MRF):}
+
+Prediction during the GMM stage results in a set of random 
+forest voting classification maps giving a probabilistic estimate
+of the label value at each voxel.  \textcolor{red}{FIXME ---i think
+  this sentence is confusing ... did you already run the RF at this
+  point?  seems no ... but ``results in a set of RF voting
+  classification maps'' sounds like you already have them.}  These spatial probability
+maps can be used as spatial priors in a MAP-MRF segmentation
+estimation protocol also performed using the Atropos segmentation 
+tool \citep{avants2011} available in ANTs although we couple
+it with iterative bias correction as contained in the script
+{\tt antsAtroposN4.sh}.  The probabilistic output (and corresponding
+feature images) 
+for each modality are used to replace the analogous GMM-based feature images used 
+during the first stage of prediction along with the other feature
+images previously generated (such as the template-based data).
+We discovered that this composite application of random
+forest models improves the resulting labeled output.
+
+\textcolor{red}{ok ... but i still don't understand this ... it gives
+  me a vague idea ... can you sketch out the equations for these
+  features/predictors? and give better intuition of what the
+  ``composite'' RF (previously referred to as concatenated?) is?}
+
+For both stages, in contrast to previous generative
+modeling approaches for multi-modal tumor segmentation 
+\citep[e.g.,][]{prastawa2003}, we do not use multivariate 
+Gaussians to specify tissue probabilities but rather incorporate each
+univariate probability map into the feature vector of the training
+data.  As pointed out in \cite{menze2010}, multivariate modeling
+might obscure the distinct biological information provided by each 
+modality.  Instead, we let the random forest construction 
+process determine the optimal combination of such multivariate
+information.  Additionally, maximum posterior labeling from both stages
+is used to determine the connected components for each label.  
+Geometric features (assigned voxel-wise) include the physical volumes 
+of each connected component, the volume to surface area ratio, 
+the elongation, and eccentricity.
+
+\paragraph{Neighborhood First-Order Statistic Feature Images}
+Similar to other work \cite{bauer2012,zikic2012} we calculated
+feature images using first order statistics such as mean, 
+variance, skewness, and entropy.  Images were created for 
+neighborhood radii of 1 and 3 voxels.
+
+\paragraph{Template-based Feature Images}
+Additional feature images are derived from the multivariate 
+symmetric template described earlier. The T1 images of the 
+subject and symmetric template are co-registered 
+\citep{tustison2013a} using the {\tt antsRegistration} 
+tool.  The motivation is similar in spirit to the method
+using reflection about the mid-sagittal plane described in 
+\cite{geremia2012}.  However, tumor growth often causes 
+distortion of the mid-plane thereby limiting the accuracy of
+this method.  By mapping to the symmetric template using the
+highly accurate SyN algorithm \cite{avants2011a}, we can 
+account for any distortion for comparing contralateral
+sides.
+
+Two images per modality use this derived mapping.  The modality-specific
+contralateral difference feature image is calculated by transforming
+the image to the symmetric template, flipping the image contralaterally,
+then transforming it back using the inverse transform, and subtracting
+the corresponding template component from the original image.  Additionally, 
+the template difference image is calculated by subtracting the warped template 
+image from the corresponding modality of the subject.
+
+\paragraph{Miscellaneous Feature Images}
+Three additional feature images are calculated which seem to aid in 
+classification as measured by the random forest variable importance
+(relative importance plots are provided in the Results section).
+First, the binary mask image is used to calculate the distance image 
+\citep{maurer2003}.  Second, from the symmetric template deformation 
+mapping we calculate the log Jacobian image.  Related to the earlier
+distance feature image, we calculate the distance mask image of the symmetric
+template and then warp that distance image to the individual subject
+for a third feature image.
+
+\subsection{Supervised Brain Tumor Segmentation with Concatenated Random Forests}
+\subsection{\textit{ANTsR}:  An ANTs/\textit{R} Interface}
+\subsection{MICCAI BRATS 2013 Brain Tumor Data}
+
+\clearpage
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{\textit{ANTsR}:  An ANTs/\textit{R} Interface}
+
+\subsubsection{Overview}
+
+The complexity of neuroimaging research necessitates 
+commensurable numerical analysis capabilities.  Similarly, concomitant
+with the era of ``big data'' (specifically with respect to neuroimaging
+\citep{vanhorn2013}) are new visualization needs and challenges
+\citep{childs2013,kehrer2013}.
+In response, various software packages have been developed to
+integrate tools specific to neuroimaging research with more general
+numerical and visualization software packages.
+The well-known neuroimaging de facto standard SPM%
+\footnote{
+http://www.fil.ion.ucl.ac.uk/spm/
+}
+is a significant extension of the commercial computing and visualization environment
+Matlab.  Open source neuroimaging packages, such as NIPY (neuroimaging in Python),%
+\footnote{
+http://nipy.org
+} 
+rely on other open source packages for numerical/statistical analysis.  NIPY,
+for example, uses the more generic packages NumPy and SciPy for numerical analysis and 
+optimization.%
+\footnote{
+http://www.numpy.org
+}
 
@@
 set of tools for multivariate neuroimage analysis \textit{ANTsR} intends to provide a modern framework for medical analytics, with a focus on imaging-assisted prediction and statistical power.
-
-%In addition to describing \textit{ANTsR} basics and how it can be
-%used generally in multivariate neuroimaging studies, 
-We showcase \textit{ANTsR} with
-the particularly difficult application of supervised brain tumor
-segmentation \citep{bauer2013} for which we use the increasingly popular 
-random forests machine learning technique.  Although random forests have
-been proposed previously in the literature for, among other things, 
-supervised brain segmentation (e.g. \cite{geremia2011}), 
-we demonstrate that \textit{ANTsR} significantly facilitates the construction 
-of a fully functional, parallelizable workflow for such a task and 
-that the performance of our proposed framework has proven extremely effective in 
-actual practice, specifically
- in the recent MICCAI 2013 Multimodal Brain Tumor Segmentation
-Challenge (see Figure \ref{fig:award}).%
-\footnote{
-http://martinos.org/qtim/miccai2013/
-}
-\begin{figure}
-  \centering
-  \includegraphics[width=90mm]{Figures/award.pdf}
-  \caption{First place certificate from the MICCAI 2013 BRATS challenge.}
-  \label{fig:award}
-\end{figure}
-
-It is interesting to note that our participation in the 2013 challenge
-was the result of observing the 2012
-challenge, wanting to take advantage of the same methods as the winners,
-and finding that no teams had availed their software in any form.  In contrast, 
-and very much in line with our philosophy 
-regarding open science \citep{tustison2013,ince2012}, we have made this framework available as open 
-source (even releasing the code before the actual 2013 contest).%
-\footnote{
-https://github.com/ntustison/BRATS2013
-}
-To our knowledge, we are the only competitors out of the 20 or so teams 
-that participated in either years' challenges to do this.
-
-\section{Materials and Methods}
-
-\subsection{\textit{ANTsR}:  An ANTs/\textit{R} Interface}
-
-\subsubsection{Overview}
-
-The complexity of neuroimaging research necessitates 
-commensurable numerical analysis capabilities.  Similarly, concomitant
-with the era of ``big data'' (specifically with respect to neuroimaging
-\citep{vanhorn2013}) are new visualization needs and challenges
-\citep{childs2013,kehrer2013}.
-In response, various software packages have been developed to
-integrate tools specific to neuroimaging research with more general
-numerical and visualization software packages.
-The well-known neuroimaging de facto standard SPM%
-\footnote{
-http://www.fil.ion.ucl.ac.uk/spm/
-}
-is a significant extension of the commercial computing and visualization environment
-Matlab.  Open source neuroimaging packages, such as NIPY (neuroimaging in Python),%
-\footnote{
-http://nipy.org
-} 
-rely on other open source packages for numerical/statistical analysis.  NIPY,
-for example, uses the more generic packages NumPy and SciPy for numerical analysis and 
-optimization.%
-\footnote{
-http://www.numpy.org
-}
 
@@
 a set of training data consisting of labeled brain images 
-(e.g. csf, gray matter, white matter, 
+(e.g., csf, gray matter, white matter, 
 necrosis, edema, enhancing and  non-enhancing tumor) is
@@
 classification techniques have been used to segment
-brain tumors (e.g. support vector machines \citep{bauer2011}),
+brain tumors (e.g., support vector machines \citep{bauer2011}),
 random forest models have proven particularly successful.
@@
 
-\begin{figure}
-  \centering
-  \begin{tabular}{cc}
-    \includegraphics[width=40mm]{Figures/S_templateFA_140.png} &
-    \includegraphics[width=40mm]{Figures/S_templateFLAIR_140.png} \\
-    (a) & (b) \\
-    \includegraphics[width=40mm]{Figures/S_templateT1_140.png} &
-    \includegraphics[width=40mm]{Figures/S_templateT2_140.png} \\
-    (c) & (d) 
-  \end{tabular}
-  \caption{Multivariate symmetric template created from the Kirby 
-           21 data described in \cite{landman2011}.  Shown are the
-           (a) fractional anisotropy (FA), (b) FLAIR, (c) MPRAGE, 
-           and (d) T2 template components.
-          }
-  \label{fig:symmetrictemplates}
-\end{figure}
-
-In order to better characterize deviations from normal
-multi-modal brain shape and appearance, several features were derived 
-using population-specific multivariate template 
-construction. A recent neuroimaging reproducibility study
-by Landman et al. resulted in an open data cohort of 21
-normal individuals, each imaged twice, comprising several
-modalities including ASL, FLAIR, DTI, fMRI, T1, and T2 
-\cite{landman2011}.  These data (known informally as the
-``Kirby'' data set) were selected for deriving
-a multivariate template due to its public availability and
-inclusion of several modalities (even permitting future 
-incorporation into our segmentation framework of modalities 
-not currently included with the BRATS challenge).
-
-As detailed in \cite{avants2008,avants2010}, 
-given $K$ image modalities for $N$ subjects,  
-${\mathbf I} = \{I_1,I_2,\ldots, I_K\}$, multivariate 
-template construction iterates between optimizing the set 
-of diffeomorphic transforms between the subjects and the 
-template, 
-$\left\{\left(\phi_1,\phi_1^{-1}\right),\ldots,\left(\phi_N,\phi_N^{-1}\right)\right\}$ 
-and constructing the 
-optimal multivariate template appearance 
-$\mathbf{J}=\{J_1,J_2,\ldots, J_K\}$ to minimize the
-following cost function:
-\begin{align}
-  \sum_{n=1}^N 
-        \Bigg[ D &\left( \psi(\mathbf{x}),\phi_1^n(\mathbf{x},1)\right) \\ \nonumber 
-        +& \sum_{k=1}^K \lambda_k \Pi_k \left(I_k^n\left(\phi_n(\mathbf{x},0.5)\right),J_k\left(\phi^{-1}_n(\mathbf{x},0.5)\right)\right)\Bigg].
-\end{align}
-$D$ is the diffeomorphic shape distance,
-\begin{align}
-D\left( \phi( \mathbf{x},0),\phi( \mathbf{x},1)\right) = \int_0^1 \| \nu(\mathbf{x},t)\|_L dt
-\end{align}
-dependent on the choice of linear operator, $L$, and $\nu$
-is the velocity field
-\begin{align}
-\nu\left( \phi(\mathbf{x},t) \right) = \frac{d\phi(\mathbf{x},t)}{dt},\,\,\, \phi(\mathbf{x},0) = \mathbf{x}.
-\end{align}
-Each pairwise registration employing the similarity metric $\Pi_k$ can 
-be assigned a relative weighting, $\lambda_k$, to weight a particular
-modality's influence in the construction process.
-
-In terms of implementation, this algorithm is 
-encapsulated in the script \verb#antsMultivariateTemplateConstruction.sh#,
-available in the ANTs repository, which permits parallel processing on
-an individual workstation or on a large computational cluster.  In
-Figure \ref{fig:symmetrictemplates} we show mid-axial slices from
-the Kirby multivariate symmetrical template consisting of FA, FLAIR,  
-T1, and T2 components.  Although DTI-based images were not included in the
-challenge data, such images have shown discriminative potential 
-\citep[e.g.][]{price2003} to warrant investigation in our future
-work.  The general procedure for const ructing a symmetric template involves flipping data around the mid-sagittal
-plane, as outlined in prior work
-\citep{Hopkins2013}. \textcolor{red}{FIXME-nick, is that correct? is
-  this covered in the scripts?  if so, where?}
-
-\subsubsection{Multi-Modal Feature Image Generation}
-
-\begin{figure*}
-  \centering
-  \makebox[\textwidth][c]{
-  \includegraphics[width=180mm]{Figures/featureImages.pdf}
-  }
-  \caption{Representative feature images derived from the 
-           BRATS\_HG0301  ``challenge'' data set.
-           Neighborhood statistical images for each modality were 
-           generated by calculating a given statistic within
-           a specified neighborhood radius.  Also calculated for each modality were feature
-           images based on either the GMM or the MAP-MRF segmentation.  For the former, we
-           show the probability maps for each of the seven labels which are used as feature
-           images.  From the resulting hard segmentation, we calculate various geometric 
-           measures per connected component of each of the seven labels.  Similarly, the 
-           registration to the symmetric template produces the modality-specific 
-           difference images with the
-           corresponding symmetric template itself and with respect to the 
-           contralateral side.   
-           This mapping is also used to produce the log Jacobian image.  Finally, the 
-           (T1 - TC) image is calculated and, from the
-           cerebral mask, we calculate the normalized distance image.  
-           }
-  \label{fig:featureImages}         
-\end{figure*}
-
-%For both training and prediction protocols, feature images are generated
-%for each subject using the selected input MRI.  
-Key to any supervised regression or classification protocol are the 
-selected features for training and subsequent prediction.  Based on previous
-work and our own experience, the following feature images are generated
-for use in both training and prediction:
-\begin{itemize}
-  \item Per modality (FLAIR, T1, T1C, T2)
-    \begin{itemize}
-      \item First-order neighborhood statistical images:
-            mean, variance, skewness, and entropy. 
-            Neighborhood radius $\in \{1,3\}$.
-    \item GMM (stage 1) and MAP-MRF (stage 2) posteriors: CSF, gray matter, white 
-          matter, necrosis, edema, non-enhancing tumor and enhancing tumor (or a
-          subset for the simulated data).
-    \item GMM (stage 1) and MAP-MRF (stage 2) connected component geometry 
-          features:  distance to tumor core label, volume, volume to surface area ratio, eccentricity, and elongation
-    \item Template-based:  symmetric template difference and 
-          contralateral difference with Gaussian smoothing ($\sigma = 4$mm).
-    \end{itemize}
-  \item Miscellaneous: normalized Euclidean distance based on cerebral mask,
-    log Jacobian image, and  (T1 - T1C) difference image.
-\end{itemize}
-The generation of all feature images is performed using the available script 
-{\tt createFeatureImages.sh} which also preprocesses the input images performing N4
-bias correction \citep{tustison2010} and intensity rescaling to $[0,1]$.  Note that 
-it is easily expandable to include 
-additional modalities and/or neighborhood and smoothing parameters.
-In Figure \ref{fig:featureImages}
-we provide sample mid-axial slices from the BRATS\_HG0301
-data set from the Challenge cohort.
-
-We previously mentioned that all data was provided already 
-skull-stripped.  However, we note to the reader that an
-available ANTs-based solution to the brain extraction problem
-is found in the script {\tt antsBrainExtraction.sh} which uses
-a hybrid approach based on image registration and binary morphology
-to isolate the brain \cite{avants2010a}.  We recently performed an evaluation of over 1200
-MRI which demonstrated performance that is comparable or better than
-other available methodology \cite{tustison2013a}.
-
-\paragraph{Stage 1 (GMM)} \textcolor{red}{Need more descriptive
-  section name ... lots going on in the paper , easy to get disoriented}
-Prior cluster centers for specific tissue types learned from training data \citep{reynolds2009} are used in the GMM random forest stage to create 
-multiple feature images.  \textcolor{red}{FIXME I think this should be
-  expanded upon a bit ... can you be more specific?  how are these
-  cluster centers passed to the RF?  why are these centers portable to
-  this new data? etcetera.}
-Given $M$ tissue types, a GMM formulates the 
-probability distribution at each voxel, $\mathbf{x}$, as the
-sum of Gaussian components, $\mathcal{N}(\mathbf{x}|\mu,\sigma)$, i.e.
-\begin{align}
-p\left(\mathbf{x}|\mu_m,\sigma_m,\lambda_m\right) = \sum_{i=1}^M \lambda_m \mathcal{N}(\mathbf{x}|\mu_m,\sigma_m)
-\end{align}
-where $\sum_{m=1}^M \lambda_m = 1$.  One popular method for 
-determining the parameters of the GMM is maximum likelihood 
-estimation which can performed using the Atropos segmentation 
-tool \citep{avants2011} available in ANTs.  \textcolor{red}{FIXME
-  would prefer active voice version of the sentence to make clear that
-  Atropos uses EM} See Figure~\ref{fig:featureImages} for
-an example segmentation.
-
-\textcolor{red}{should there be an RF model equation here?  i.e. what
-  exactly (in R) was used ?}
-
-\paragraph{Stage 2 (MAP-MRF):}
-
-Prediction during the GMM stage results in a set of random 
-forest voting classification maps giving a probabilistic estimate
-of the label value at each voxel.  \textcolor{red}{FIXME ---i think
-  this sentence is confusing ... did you already run the RF at this
-  point?  seems no ... but ``results in a set of RF voting
-  classification maps'' sounds like you already have them.}  These spatial probability
-maps can be used as spatial priors in a MAP-MRF segmentation
-estimation protocol also performed using the Atropos segmentation 
-tool \citep{avants2011} available in ANTs although we couple
-it with iterative bias correction as contained in the script
-{\tt antsAtroposN4.sh}.  The probabilistic output (and corresponding
-feature images) 
-for each modality are used to replace the analogous GMM-based feature images used 
-during the first stage of prediction along with the other feature
-images previously generated (such as the template-based data).
-We discovered that this composite application of random
-forest models improves the resulting labeled output.
-
-\textcolor{red}{ok ... but i still don't understand this ... it gives
-  me a vague idea ... can you sketch out the equations for these
-  features/predictors? and give better intuition of what the
-  ``composite'' RF (previously referred to as concatenated?) is?}
-
-For both stages, in contrast to previous generative
-modeling approaches for multi-modal tumor segmentation 
-\citep[e.g.][]{prastawa2003}, we do not use multivariate 
-Gaussians to specify tissue probabilities but rather incorporate each
-univariate probability map into the feature vector of the training
-data.  As pointed out in \cite{menze2010}, multivariate modeling
-might obscure the distinct biological information provided by each 
-modality.  Instead, we let the random forest construction 
-process determine the optimal combination of such multivariate
-information.  Additionally, maximum posterior labeling from both stages
-is used to determine the connected components for each label.  
-Geometric features (assigned voxel-wise) include the physical volumes 
-of each connected component, the volume to surface area ratio, 
-the elongation, and eccentricity.
-
-\paragraph{Neighborhood First-Order Statistic Feature Images}
-Similar to other work \cite{bauer2012,zikic2012} we calculated
-feature images using first order statistics such as mean, 
-variance, skewness, and entropy.  Images were created for 
-neighborhood radii of 1 and 3 voxels.
-
-\paragraph{Template-based Feature Images}
-Additional feature images are derived from the multivariate 
-symmetric template described earlier. The T1 images of the 
-subject and symmetric template are co-registered 
-\citep{tustison2013a} using the {\tt antsRegistration} 
-tool.  The motivation is similar in spirit to the method
-using reflection about the mid-sagittal plane described in 
-\cite{geremia2012}.  However, tumor growth often causes 
-distortion of the mid-plane thereby limiting the accuracy of
-this method.  By mapping to the symmetric template using the
-highly accurate SyN algorithm \cite{avants2011a}, we can 
-account for any distortion for comparing contralateral
-sides.
-
-Two images per modality use this derived mapping.  The modality-specific
-contralateral difference feature image is calculated by transforming
-the image to the symmetric template, flipping the image contralaterally,
-then transforming it back using the inverse transform, and subtracting
-the corresponding template component from the original image.  Additionally, 
-the template difference image is calculated by subtracting the warped template 
-image from the corresponding modality of the subject.
-
-\paragraph{Miscellaneous Feature Images}
-Three additional feature images are calculated which seem to aid in 
-classification as measured by the random forest variable importance
-(relative importance plots are provided in the Results section).
-First, the binary mask image is used to calculate the distance image 
-\citep{maurer2003}.  Second, from the symmetric template deformation 
-mapping we calculate the log Jacobian image.  Related to the earlier
-distance feature image, we calculate the distance mask image of the symmetric
-template and then warp that distance image to the individual subject
-for a third feature image.
-
 \subsubsection{Training:  Random Forest Model Creation}
@@
 statistical features also seemed fairly important which is something that
-previous work had demonstrated \citep[e.g.][]{bauer2012,zikic2012}.
+previous work had demonstrated \citep[e.g.,][]{bauer2012,zikic2012}.
 There are a number of differences, however, between low-grade and high-grade 

### 2014-02-15 paper.tex
ENH: Fixed figures.
ENH:  Reconfiguration.

@@
 
-%\title{Multivariate Neuroimage Analysis with \textit{ANTsR}: Application to Supervised Brain Tumor Segmentation using Random Forests}
-\title{Optimal Symmetric Multimodal Templates and Concatenated Random Forests for Supervised Brain Tumor Segmentation (made easy) with \textit{ANTsR}}
+%\title{Multivariate Neuroimage Analysis with ANTsR: Application to Supervised Brain Tumor Segmentation using Random Forests}
+\title{Optimal Symmetric Multimodal Templates and Concatenated Random Forests for Supervised Brain Tumor Segmentation (made easy) with ANTsR}
 
@@
  to which we refer as {\it concatenated random forests}.
-Although the derived workflow involves many disparate components, the newly developed \textit{ANTsR} open source software package significantly facilitates the proposed methodology.
-\textit{ANTsR}, a comprehensive interface of the popular Advanced Normalization Tools (ANTs) repository with the \textit{R} statistical project,
+Although the derived workflow involves many disparate components, the newly developed ANTsR open source software package significantly facilitates the proposed methodology.
+ANTsR, a comprehensive interface of the popular Advanced Normalization Tools (ANTs) repository with the R statistical project,
 integrates well-performing image analysis tools such as image registration, segmentation, bias correction, and template building with access to advanced statistical and machine learning techniques as well as state-of-the-art visualization possibilities.
@@
 \begin{keyword}
-advanced normalization tools \sep BRATS \sep brain tumor segmentation \sep \textit{R} project 
+advanced normalization tools \sep BRATS \sep brain tumor segmentation \sep R project for statistical computing
 %% keywords here, in the form: keyword \sep keyword
@@
 detailed in recent reviews \citep{angelini2007,bauer2013}.  
-Any type of commensurable algorithmic evaluation, however, is
+Commensurable algorithmic evaluation, however, is
 extremely problematic due to widely varying performance assessments
@@
 } 
-of the National Institutes of Health and is available through our \textit{ANTsR} 
+of the National Institutes of Health and is available through our ANTsR 
 package.
@@
 %\cite{geremia2011}), 
-%we demonstrate that \textit{ANTsR} significantly facilitates the construction 
+%we demonstrate that ANTsR significantly facilitates the construction 
 %of a fully functional, parallelizable workflow for such a task and 
@@
 and appearance, several image features were derived using symmetric 
-population-specific multivariate templates \citep{avants2008,avants2010}.  
+population-specific multivariate templates.  
 For normal neuroanatomy, the use of spatial prior information 
@@
 normal tissue due to tumor growth causing the mid-sagittal plane
-to deform from its planar structure (cf Figure \ref{fig:asymmetry} (a)).
+to deform from its planar structure (cf Figure \ref{fig:asymmetry}).
 
@@
 templates from normal data using the same acquisition 
-parameters, we substituted publicly available data
-
+parameters \citep{avants2010}, we substituted well-known,
+publicly available data since such normal data was not made available
+for the BRATS 2013 challenge. 
 A recent neuroimaging reproducibility study
@@
 
+\begin{figure*}[htb]
+  \centering
+    \includegraphics[width=160mm]{Figures/templateKirby.pdf}
+  \caption{Multivariate symmetric template created from the MMRR 
+           data \citep{landman2011}.  Of the seven modalities 
+           comprising the set of study acquisitions, we illustrate the
+           (a) FA, (b) FLAIR, (c) MPRAGE, and (d) T2 template components.
+           The optimal transformation and averaging of the individual 
+           subject images result in the asymmetric template represented at 
+           the top of the middle column.  A horizontal reflection 
+           perpendicular to the mid-sagittal
+           plane, $T_R$, resulted in the contralateral counterpart represented
+           at the bottom of the second column.  The final template seen on
+           the right is a result of repeating the template construction using
+           the two asymmetric templates as input.
+          }
+  \label{fig:symmetrictemplates}
+\end{figure*}
+
 As detailed in \cite{avants2008,avants2010}, 
-given $K$ image modalities for $N$ subjects,  
-${\mathbf I} = \{I_1,I_2,\ldots, I_K\}$, multivariate 
+given $K$ image modalities, ${\mathbf I} = \{I_1,I_2,\ldots, I_K\}$, 
+for $N$ subjects,  multivariate 
 template construction iterates between optimizing the set 
@@
 multivariate template construction a second time using only the multivariate
-template and its symmetrical version.
+template and its symmetrical version.  This is illustrated conceptually in
+Figure \ref{fig:symmetrictemplates}.
 
@@
 available in the ANTs repository, which permits parallel processing on
-an individual workstation or on a large computational cluster.  In
+an individual workstation or on a computational cluster.  In
 Figure \ref{fig:symmetrictemplates} we show mid-axial slices from
@@
 }, FLAIR,  
-T1, and T2 components.  These images have been made publicly available%
+T1, and T2 components.  These images have been made publicly available
+along with other processed images such as corresponding
+tissue priors and structural labels.%
 \footnote{
 http://figshare.com/articles/Kirby\_multivariate\_template/852989.
-All 7 components are stored under the more informal ``Kirby'' moniker.
-}
-along with other processed images such as corresponding
-tissue priors and structural labels.
-
-{\color{red}{Need to describe the actual feature images}}
-
-\begin{figure}[htb]
-  \centering
-    \includegraphics[width=85mm]{Figures/templateKirby.pdf}
-  \caption{Multivariate symmetric template created from the MMRR 
-           data described in \cite{landman2011}.  Shown are the
-           (a) FA, (b) FLAIR, (c) MPRAGE, 
-           and (d) T2 template components.
-          }
-  \label{fig:symmetrictemplates}
-\end{figure}
+All 7 components are stored under the more informal ``Kirby'' data set moniker.
+}
+
+After constructing the template offline, each data set is processed
+by first registering the non-contrast T1-weighted image to the T1-weighted
+component of the symmetric template.  To do this we use a recently developed 
+SyN \citep{avants2011a} variant based on B-spline regularization 
+which has demonstrated good performance in normal brain registration.
+Good performance also extends to this pathological data scenario as 
+indicated by visual inspection and the fact that the derived features
+were amongst the most informative in our winning entry.
+
+We denote the mapping from the subject, $S$, to the template, $T$,
+space as $\phi_b: S \leftrightarrow \underset{b}{\leftrightsquigarrow}  T$
+which consists of both affine and diffeomorphic components.  Transform
+invertibility is essential for the features we describe below.  The first 
+set of feature images is generated by warping the template components 
+to the subject space and calculating the difference image (see the 
+top portion of Figure \ref{fig:asymmetryFeatures}).  For example,
+the T2 symmetric template difference image is calculated from
+\begin{align}
+  \mathrm{T2}\,\,\mathrm{symmetric}\,\,\mathrm{template}\,\,\mathrm{difference} = S_{\mathrm{T2}} - T_{\mathrm{T2}}\left(\phi_b^{-1}\right).
+\end{align}
+Note that the T1-weighted contrast and non-contrast images are paired with
+the T1-weighted component of the symmetric template.  The second
+set of template-based feature images is generated per modality 
+as the difference image with the contralateral reflection.  This is achieved by calculating the reflection transform, $T_R$, in the symmetric template space and composing transforms
+as follows to create the non-Euclidean contralateral modality counterpart:
+\begin{align}
+  S_{contralateral} = S\left( \phi_b^{-1}\left( T_R\left( \phi_b \right)\right)\right).
+\end{align}
+This process is illustrated in the bottom portion of Figure \ref{fig:asymmetryFeatures}. A final, related feature is the Jacobian
+of the transform.  The basis for this final feature is that 
+relatively larger Jacobian values might be indicative of tumor
+expansion.
 
@@
     \includegraphics[width=85mm]{Figures/asymmetryFeatureImages2.pdf}
-  \caption{ $S \underset{b}{\leftrightsquigarrow} \leftrightarrow T$
+  \caption{ Given the mapping between the template, $T$, and subject, $S$, domains ($\phi_b: S  \leftrightarrow \underset{b}{\leftrightsquigarrow} T$), various features can be calculated which demonstrate good discriminative qualities.
+Feature images used are outlined in a dashed line.
+Top:  Difference images with the symmetric multivariate template are created by warping the template to the subject space and performing a voxelwise subtraction from the original modality image.
+Bottom:  Similarly, contralateral difference images are calculated from each modality per subject by generating the non-Euclidean contralateral image via the diffeomorphic transform $\phi_b$.  
           }
@@
 
-\subsection{Multi-Modal Feature Image Generation}
-
-%For both training and prediction protocols, feature images are generated
-%for each subject using the selected input MRI.  
-Key to any supervised regression or classification protocol are the 
-selected features for training and subsequent prediction.  Based on previous
-work and our own experience, the following feature images are generated
-for use in both training and prediction:
+\subsection{Other Multi-Modal Feature Image Generation}
+
+In addition to the feature images outlined in the previous section, 
+several others were generated as described as follows:
 \begin{itemize}
@@
 In Figure \ref{fig:featureImages}
-we provide sample mid-axial slices from the BRATS\_HG0301
+we provide sample mid-axial slices from the {\tt BRATS\_HG0301}
 data set from the Challenge cohort.
@@
 \subsection{Supervised Brain Tumor Segmentation with Concatenated Random Forests}
-\subsection{\textit{ANTsR}:  An ANTs/\textit{R} Interface}
+\subsection{ANTsR:  An ANTs/R Interface}
 \subsection{MICCAI BRATS 2013 Brain Tumor Data}
@@
 
-\subsection{\textit{ANTsR}:  An ANTs/\textit{R} Interface}
+\subsection{ANTsR:  An ANTs/R Interface}
 
@@
 that robust statistical machinery was lacking for making inferences regarding
-the data produced during the course of ANTs processing.  \textit{ANTsR} was developed
+the data produced during the course of ANTs processing.  ANTsR was developed
 specifically to provide an interface between ANTs, a 
 powerful neuroimaging toolkit for producing reliable imaging data 
-transformations, and the \textit{R} project%
+transformations, and the R project%
 \footnote{
@@
 for statistical computing and visualization thus providing a complete
-set of tools for multivariate neuroimage analysis \textit{ANTsR} intends to provide a modern framework for medical analytics, with a focus on imaging-assisted prediction and statistical power.
+set of tools for multivariate neuroimage analysis ANTsR intends to provide a modern framework for medical analytics, with a focus on imaging-assisted prediction and statistical power.
 
 Careful consideration of available statistical software 
-led to the adoption of \textit{R} to complement ANTs quantification resulting in the
-\textit{ANTsR} package.
+led to the adoption of R to complement ANTs quantification resulting in the
+ANTsR package.
 %The R project for statistical computing,%
@@
 %and data visualization.  
-\textit{R}'s open source code base, reliable software testing and distribution strategies,
+R's open source code base, reliable software testing and distribution strategies,
 and add-on packages coupled with its rapidly growing 
@@
 
-The \textit{ANTsR} package is publicly available on the github project hosting service.%
+The ANTsR package is publicly available on the github project hosting service.%
 \footnote{
@@
 }
-Prior to installation of \textit{ANTsR}, several external R packages
+Prior to installation of ANTsR, several external R packages
 need to be installed including: \verb#Rcpp#, \verb#signal#, \verb#timeSeries#, 
@@
 \footnote{
-Packages are easily installed using the {\tt install.packages()} \textit{R} mechanism.
-}
-
-In addition to \textit{R} and the add-on packages previously mentioned, CMake is also 
+Packages are easily installed using the {\tt install.packages()} R mechanism.
+}
+
+In addition to R and the add-on packages previously mentioned, CMake is also 
 required.  CMake%
@@
 and ANTs.  Further instructions for download and
-installation can be found on the \textit{ANTsR} github website.  Feel
+installation can be found on the ANTsR github website.  Feel
 free to contact the authors if installation trouble occurs.  We note
-that \textit{ANTsR} is currently only tested on UNIX-alikes such OSX and Ubuntu
+that ANTsR is currently only tested on UNIX-alikes such OSX and Ubuntu
 operating systems.  \footnote{Windows installation should be possible
@@
 \subsubsection{Usage}
-\textit{ANTsR} is intended to not only allow easy interchange between
-medical imaging formats and \textit{R} but also to facilitate
+ANTsR is intended to not only allow easy interchange between
+medical imaging formats and R but also to facilitate
 reproducible scientific studies and the type compilable analysis
@@
 \textit{Biostatistics}.  Both \verb#knitr# and \verb#sweave#
-facilitate integration of \textit{R}-code with the LaTeX document
+facilitate integration of R-code with the LaTeX document
 preparation system.
 
-An additional motivation for our development of \textit{ANTsR} (and
+An additional motivation for our development of ANTsR (and
 hopefully its acceptance by the community) 
@@
 functionality, including IO tools such as \verb#antsImageRead#, 
-with the large number of \textit{R} statistical and
+with the large number of R statistical and
 visualization packages.  Due to this combination, several
@@
 constituent functions are invoked in the similar manner as other
-\textit{R} libraries.
+R libraries.
 
 As mentioned earlier, we have made this entire framework
-available as open source.  In addition to the \textit{ANTsR} repository
-already on github which houses both ANTs and \textit{ANTsR} functionality, 
+available as open source.  In addition to the ANTsR repository
+already on github which houses both ANTs and ANTsR functionality, 
 we created a special github repository specifically for this work
@@
 Also, we posted all
-scripts (\textit{R}, shell, and perl) used to coordinate the \textit{ANTsR} processing 
+scripts (R, shell, and perl) used to coordinate the ANTsR processing 
 including:
@@
 \footnote{
-One of the principal advantages of \textit{R} is the extensive community of
+One of the principal advantages of R is the extensive community of
 developers  who have contributed on the order of thousands of packages 
-extending \textit{R} 's capabilities beyond its core functionality.
+extending R 's capabilities beyond its core functionality.
 Most relevant 
@@
 feature images are created for each subject, the resulting images of the entire
-training cohort are organized in a csv file for input into the \textit{R} script
+training cohort are organized in a csv file for input into the R script
 {\tt createModel.R}.  Other possible input parameters include the requested 
@@
 This aids in potential feature pruning or intuiting model behavior.  As mentioned
-earlier, we provide the \textit{R} script {\tt plotVariableImportance.R} to render
+earlier, we provide the R script {\tt plotVariableImportance.R} to render
 one such quantity denoted as {\tt MeanDecreaseAccuracy}.  During model construction
@@
 methodology that we have made available is only a small part of the 
-larger software package that we have created in \textit{ANTsR}.  Not only does
-\textit{ANTsR} significantly facilitate the development of the work discussed,
+larger software package that we have created in ANTsR.  Not only does
+ANTsR significantly facilitate the development of the work discussed,
 but it provides an interface to one of the most powerful statistical
-packages available in \textit{R}.  The combination of the well-known ANTs software
-package with \textit{R} provides tremendous potential for future insightful analysis.
+packages available in R.  The combination of the well-known ANTs software
+package with R provides tremendous potential for future insightful analysis.
 

### 2014-02-25 paper.tex
ENH:  Complete rewrite.

@@
 
-% \documentclass[preprint,authoryear,review,12pt]{elsarticle}
-\documentclass[final,5p,times,twocolumn]{elsarticle}
+\documentclass[preprint,authoryear,review,12pt]{elsarticle}
+%\documentclass[final,5p,times,twocolumn]{elsarticle}
 
@@
 \usepackage{rccol}
+\usepackage{pbox}
 %\usepackage[table]{xcolor}
@@
 
-%\title{Multivariate Neuroimage Analysis with ANTsR: Application to Supervised Brain Tumor Segmentation using Random Forests}
-\title{Optimal Symmetric Multimodal Templates and Concatenated Random Forests for Supervised Brain Tumor Segmentation (made easy) with ANTsR}
+\title{Optimal Symmetric Multimodal Templates and Concatenated Random Forests for Supervised Brain Tumor Segmentation (Simplified) with ANTsR}
 
@@
 in related tasks such as normal brain tissue segmentation.  Due to this 
-limitation, we localize our prior knowledge strategy to each subject through the 
+limitation, we localize our prior knowledge strategy through the 
 use of subject-specific multimodal asymmetrical features which demonstrate excellent
@@
 %\input{methods}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%
+% Introduction
+%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 After observing the 2012 challenge to gain insight for our own research 
-needs, we found that no teams had availed their software in any 
+needs \citep{durst2014}, we found that no teams had availed their software in any 
 form which prompted our participation in the challenge which took place
@@
 \cite{bauer2012,geremia2012,zikic2012}), we adopted the increasingly popular 
-random forest machine learning framework \citep{breiman2001} which 
-permits the inclusion of various features which might prove saliently 
-discriminative.  Instrumental to our success in the 2013 challenge 
-was the use of features based on multimodal shape and appearance 
+random forest (RF) machine learning framework \citep{breiman2001} which 
+permits the inclusion of potentially discriminative image features.  
+Supervised segmentation techniques generally consist of a training phase
+for model construction using image-based feature data followed by prediction using the 
+generated model.  For supervised brain tumor segmentation, 
+a set of training data consisting of labeled brain images 
+is used to construct a predictive model.  Although other 
+classification techniques have been used to segment
+brain tumors (e.g., support vector machines \citep{bauer2011}),
+RF models have proven particularly successful.
+
+Several machine learning concepts were integrated to create 
+the RF framework first articulated in its entirety by Breiman
+et al. \citep{breiman2001} for performing classification/regression.  
+Although decision trees had been previously explored in the literature, 
+it was the success of ``boosting''-style machine learning 
+techniques, such as AdaBoost \citep{schapire1990,freund1997}, which influenced 
+the aggregation of such decision trees into ``forests'' 
+with randomized node optimization for improved
+classification/regression performance \citep{ho1995,amit1997}.
+The final element of bootstrap aggregating or ``bagging'' (i.e.
+random sampling of the training data) was
+introduced by Breiman \citep{breiman1996} to achieve improved
+accuracy.%
+\footnote{
+One of the principal advantages of R is the extensive community of
+developers  who have contributed on the order of thousands of packages 
+extending R 's capabilities beyond its core functionality.
+Most relevant here
+is the {\tt randomForest} package developed from Breiman's original
+Fortran code by Liaw and Wiener \citep{liaw2002}.
+}
+
+Early adoption \cite{viola2005} and success in the
+computer vision community
+has led to a recent surge within the medical image analysis
+community of using RFs for handling complex 
+classification/regression tasks including
+normal brain segmentation \citep{yi2009},
+MS lesion segmentation \cite{geremia2011}, 
+multimodal brain tumor segmentationp
+\citep{bauer2012,zikic2012}, brain extraction \citep{iglesias2010}, 
+anatomy detection in computed tomography \citep{criminisi2013}, and
+segmentation of echocardiographic images \citep{verhoek2011}. 
+A thorough introduction for those interested in delving deeper 
+into the more theoretical aspects of RFs can be found
+in \cite{criminisi2011}.
+
+%One of the innovations that we provide in this work is
+%the use of \textit{concatenated RFs}  for improved probabilistic estimation
+%of the brain segmentation labels.  Briefly, Gaussian mixture
+%modeling (GMM) is used to create a subset of feature images which are
+%used to train and predict during the first classification stage.  
+%The RF voting output probabilistic maps are then used as 
+%spatial probability priors for performing a MAP-MRF segmentation
+%and subsequent feature image creation
+%thereby replacing the GMM-derived feature images for a second
+%classification stage.  This dual RF model generation
+%and application improves results and is detailed in later sections.
+
+Instrumental to our success in the 2013 challenge 
+was the use of image features based on multimodal shape and appearance 
 asymmetries which were generated via the construction of symmetric, 
 multivariate brain templates.  Additional improvements in performance
-were due to the use of a two-stage random forest model approach
+were due to the use of a two-stage RF model approach
 whereby the output probability images from the application of the first
-random forest model seeds the construction of a subset of new feature images
-for input into the second stage.  The results of the second random forest model
+RF model seeds the construction of a subset of new feature images
+for input into the second stage.  The results of the second RF model
 are then refined using a series of heuristically-derived binary morphological
@@
 }
-and, to our knowledge, we are the only competitors out of the 20 or so teams 
+and, to our knowledge, we are the only competitors out of the approximately 20 teams 
 that participated in either years' challenges to do this.  All code is based
@@
 package.
-
-In subsequent sections, we describe in greater detail the automated 
-supervised brain segmentation pipeline including the asymmetry
-analysis and the concatenated random forest construction.  We also
-describe the code and data used for the BRATS challenge so that 
+In subsequent sections we describe in greater detail the automated 
+supervised brain segmentation pipeline including the generation of asymmetry
+feature images and the concatenated RF construction.  We also
+describe the code and data%
+\footnote{
+Although all data was provided having been skull-stripped, 
+we note to the reader that an
+available ANTs-based solution to the brain extraction problem
+is found in the script {\tt antsBrainExtraction.sh} which uses
+a hybrid approach based on image registration and binary morphology
+to isolate the brain \citep{avants2010a}.  An evaluation with other
+methodologies demonstrated comparable to superior performance 
+\citep{avants2010a}.
+} 
+used for the BRATS challenge so that 
 the interested reader can reproduce our results.  Finally, we 
@@
 
-% which has demonstrated good performance
-%in other neuro-related tasks \citep[e.g.,][]{geremia2011}.
-
-%Although random forests have been proposed previously in the 
-%literature for, among other things, supervised brain segmentation (e.g. 
-%\cite{geremia2011}), 
-%we demonstrate that ANTsR significantly facilitates the construction 
-%of a fully functional, parallelizable workflow for such a task and 
-%that the performance of our proposed framework has proven extremely effective in 
-%actual practice, specifically
-% in the recent MICCAI 2013 Multimodal Brain Tumor Segmentation
-%Challenge.%
-%\footnote{
-%http://martinos.org/qtim/miccai2013/
-%}
-%\begin{figure}
-%  \centering
-%  \includegraphics[width=90mm]{Figures/award.pdf}
-%  \caption{First place certificate from the MICCAI 2013 BRATS challenge.}
-%  \label{fig:award}
-%\end{figure}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%
+% Materials and Methods
+%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
   \item generation of other image-based features,
-  \item training of random forest models, and
+  \item training of RF models, and
   \item prediction using the proposed framework.
 \end{itemize}
-A description of each of these components is detailed below along.
-
-\subsection{Multimodal Asymmetry Features from Multivariate Symmetric Templates}
+In subsequent sections, we describe these items and detail the
+ANTsR framework which coordinates all steps.
+
+\subsection{Image Features for RF Model Training and Prediction}
+
+\subsubsection{Multimodal Asymmetry Features from Multivariate Symmetric Templates}
 
@@
 
-\begin{figure}[htb]
+\begin{figure}[h]
   \centering
   \begin{tabular}{cccc}
-  \includegraphics[width=20mm]{Figures/BRATS_HG0001/BRATS_HG0001_FLAIR_slice93.png} &
-  \includegraphics[width=20mm]{Figures/BRATS_HG0001/BRATS_HG0001_T1_slice93.png} &
-  \includegraphics[width=20mm]{Figures/BRATS_HG0001/BRATS_HG0001_T1C_slice93.png} &
-  \includegraphics[width=20mm]{Figures/BRATS_HG0001/BRATS_HG0001_T2_slice93.png} \\
+  \includegraphics[width=30mm]{Figures/BRATS_HG0001/BRATS_HG0001_FLAIR_slice93.png} &
+  \includegraphics[width=30mm]{Figures/BRATS_HG0001/BRATS_HG0001_T1_slice93.png} &
+  \includegraphics[width=30mm]{Figures/BRATS_HG0001/BRATS_HG0001_T1C_slice93.png} &
+  \includegraphics[width=30mm]{Figures/BRATS_HG0001/BRATS_HG0001_T2_slice93.png} \\
   (a) & (b) &
@@
 \end{figure}
-
-\setlength{\tabcolsep}{6pt}
 
@@
 exhibits a bilaterally symmetric organization, we can 
-use the presence of asymmetries to differentiate abnormal 
-brain tissue (i.e. tumor).  A similar motivation prompted
+use the presence of asymmetries to potentially differentiate abnormal 
+brain tissue.  A similar motivation prompted
 the identification of the mid-sagittal plane of symmetry \citep{prima2002}
-for feature generation in MS lesion \citep{geremia2011} and 
+for feature generation in multiple sclerosis lesion \citep{geremia2011} and 
 tumor \citep{geremia2012} identification.  However, this
@@
 to deform from its planar structure (cf Figure \ref{fig:asymmetry}).
+
+\begin{figure}[h]
+    \centerline{\includegraphics[width=130mm]{Figures/templateKirby.pdf}}
+  \caption{Multivariate symmetric template created from the MMRR 
+           data \citep{landman2011}.  Of the seven modalities 
+           comprising the set of study acquisitions, we illustrate the
+           (a) FA, (b) FLAIR, (c) MPRAGE, and (d) T2 template components.
+           The optimal transformation and averaging of the individual 
+           subject images result in the asymmetric template represented at 
+           the top of the middle column.  A horizontal reflection 
+           perpendicular to the mid-sagittal
+           plane, $T_R$, resulted in the contralateral counterpart represented
+           at the bottom of the second column.  The final template seen on
+           the right is a result of repeating the template construction using
+           the two asymmetric templates as input.
+          }
+  \label{fig:symmetrictemplates}
+\end{figure}
 
@@
 
-\begin{figure*}[htb]
+\begin{figure}[!h]
   \centering
-    \includegraphics[width=160mm]{Figures/templateKirby.pdf}
-  \caption{Multivariate symmetric template created from the MMRR 
-           data \citep{landman2011}.  Of the seven modalities 
-           comprising the set of study acquisitions, we illustrate the
-           (a) FA, (b) FLAIR, (c) MPRAGE, and (d) T2 template components.
-           The optimal transformation and averaging of the individual 
-           subject images result in the asymmetric template represented at 
-           the top of the middle column.  A horizontal reflection 
-           perpendicular to the mid-sagittal
-           plane, $T_R$, resulted in the contralateral counterpart represented
-           at the bottom of the second column.  The final template seen on
-           the right is a result of repeating the template construction using
-           the two asymmetric templates as input.
-          }
-  \label{fig:symmetrictemplates}
-\end{figure*}
+    \includegraphics[width=100mm]{Figures/asymmetryFeatureImages.pdf}
+  \caption{ Given the mapping between the template, $T$, and subject, $S$, domains ($\phi_b: S  \leftrightarrow \underset{b}{\leftrightsquigarrow} T$), various features can be calculated which demonstrate good discriminative qualities.
+Feature images used are outlined in a dashed line.
+Top:  Difference images with the symmetric multivariate template are created by warping the template to the subject space and performing a voxelwise subtraction from the original modality image.
+Bottom:  Similarly, contralateral difference images are calculated from each modality per subject by generating the non-Euclidean contralateral image via the diffeomorphic transform $\phi_b$.  
+          }
+  \label{fig:asymmetryFeatures}
+\end{figure}
 
 As detailed in \cite{avants2008,avants2010}, 
-given $K$ image modalities, ${\mathbf I} = \{I_1,I_2,\ldots, I_K\}$, 
+given $K$ multimodality images, ${\mathbf I} = \{I_1,I_2,\ldots, I_K\}$, 
 for $N$ subjects,  multivariate 
@@
   \sum_{n=1}^N 
-        \Bigg[ D &\left( \psi(\mathbf{x}),\phi_1^n(\mathbf{x},1)\right) \\ \nonumber 
-        +& \sum_{k=1}^K \lambda_k \Pi_k \left(I_k^n\left(\phi_n(\mathbf{x},0.5)\right),J_k\left(\phi^{-1}_n(\mathbf{x},0.5)\right)\right)\Bigg].
+        \left[ D \left( \psi(\mathbf{x}),\phi_1^n(\mathbf{x},1)\right)  
+        + \sum_{k=1}^K \lambda_k \Pi_k \left(I_k^n\left(\phi_n(\mathbf{x},0.5)\right),J_k\left(\phi^{-1}_n(\mathbf{x},0.5)\right)\right)\right]
 \end{align}
@@
 modality's influence in the construction process.  Once the multivariate
-template has converged (typically four iterations are used), we symmetrize
-the template by flipping each image contralaterally and then running the
+template has converged (typically in four iterations), we symmetrize
+the template by flipping each asymmetric template component contralaterally 
+and then running the
 multivariate template construction a second time using only the multivariate
-template and its symmetrical version.  This is illustrated conceptually in
+template and its symmetric analog.  This is illustrated conceptually in
 Figure \ref{fig:symmetrictemplates}.
@@
 encapsulated in the script \verb#antsMultivariateTemplateConstruction.sh#,
-available in the ANTs repository, which permits parallel processing on
+available in the ANTs repository, which permits serial or parallel processing on
 an individual workstation or on a computational cluster.  In
@@
 \footnote{
-Although DTI-based images were not included in the
+Although DWI-based images were not included in the
 challenge data, such images have shown discriminative potential 
-\citep[e.g.,][]{price2003} warranting investigation in our future
+\citep{price2003,cha2005} warranting investigation in our future
 work.  
@@
 SyN \citep{avants2011a} variant based on B-spline regularization 
-which has demonstrated good performance in normal brain registration.
+which has demonstrated good performance in normal brain registration \citep{tustison2013a}.
 Good performance also extends to this pathological data scenario as 
@@
 which consists of both affine and diffeomorphic components.  Transform
-invertibility is essential for the features we describe below.  The first 
+invertibility is essential for the template-based features.  The first 
 set of feature images is generated by warping the template components 
@@
 top portion of Figure \ref{fig:asymmetryFeatures}).  For example,
-the T2 symmetric template difference image is calculated from
+the T2 symmetric template voxelwise difference image is calculated from
 \begin{align}
@@
 as the difference image with the contralateral reflection.  This is achieved by calculating the reflection transform, $T_R$, in the symmetric template space and composing transforms
-as follows to create the non-Euclidean contralateral modality counterpart:
+as follows to create the non-Euclidean contralateral counterpart per modality:
 \begin{align}
@@
 
-\begin{figure}[!htb]
+\subsubsection{Image Features for Concatenated Random Forests}
+
+\begin{figure}[!h]
   \centering
-    \includegraphics[width=85mm]{Figures/asymmetryFeatureImages2.pdf}
-  \caption{ Given the mapping between the template, $T$, and subject, $S$, domains ($\phi_b: S  \leftrightarrow \underset{b}{\leftrightsquigarrow} T$), various features can be calculated which demonstrate good discriminative qualities.
-Feature images used are outlined in a dashed line.
-Top:  Difference images with the symmetric multivariate template are created by warping the template to the subject space and performing a voxelwise subtraction from the original modality image.
-Bottom:  Similarly, contralateral difference images are calculated from each modality per subject by generating the non-Euclidean contralateral image via the diffeomorphic transform $\phi_b$.  
-          }
-  \label{fig:asymmetryFeatures}
+  \begin{tabular}{c}
+    \includegraphics[width=100mm]{Figures/pipeline2.pdf}
+  \end{tabular}
+  \caption{Diagrammatic workflow for the proposed RF model training and prediction.  The feature
+  images are first generated from the input MRI and symmetric multivariate template. If training,
+  the set of feature images derived from the training data are used to create the RF model
+  for the first stage.  The model is then applied to the training or prediction data to generate 
+  a second set of feature images to create the second stage RF model (if training) and/or generate
+  the second RF model.  
+  For prediction, the final refinement process entails a series of heuristically-derived
+  binary morphological resulting in the final labeled brain.  
+  }
+  \label{fig:pipeline}
 \end{figure}
 
-\subsection{Other Multi-Modal Feature Image Generation}
-
-In addition to the feature images outlined in the previous section, 
-several others were generated as described as follows:
+Our segmentation protocol involves training and application
+of two RF models in succession 
+(see Figure \ref{fig:pipeline}) to which we refer as ``stages.''  The basic idea is that
+we generate a set of feature images used as input to the
+first RF model (or first stage) which produces a voxelwise 
+probabilistic tissue estimate%
+\footnote{
+More precisely, in the RF framework, each prediction sample (i.e. the feature vector at each voxel),
+is propagated through each tree of the ensemble where it is labeled as belonging to a specific
+class.  These ``votes'' are converted to voxelwise probabilistic estimates for each class.
+}
+ for the seven brain/tumor tissue types:
 \begin{itemize}
-  \item Per modality (FLAIR, T1, T1C, T2)
-    \begin{itemize}
-      \item First-order neighborhood statistical images:
-            mean, variance, skewness, and entropy. 
-            Neighborhood radius $\in \{1,3\}$.
-    \item GMM (stage 1) and MAP-MRF (stage 2) posteriors: CSF, gray matter, white 
-          matter, necrosis, edema, non-enhancing tumor and enhancing tumor (or a
-          subset for the simulated data).
-    \item GMM (stage 1) and MAP-MRF (stage 2) connected component geometry 
-          features:  distance to tumor core label, volume, volume to surface area ratio, eccentricity, and elongation
-    \item Template-based:  symmetric template difference and 
-          contralateral difference with Gaussian smoothing ($\sigma = 4$mm).
+\item cerebrospinal fluid,
+\item grey matter,
+\item white matter,
+\item edema, 
+\item non-enhancing tumor (including low-grade tumor center), 
+\item enhancing tumor (excluding necrotic center), and 
+\item abnormal necrotic center or necrocyst in high-grade gliomas.
 \end{itemize}
-  \item Miscellaneous: normalized Euclidean distance based on cerebral mask,
-    log Jacobian image, and  (T1 - T1C) difference image.
-\end{itemize}
-The generation of all feature images is performed using the available script 
-{\tt createFeatureImages.sh} which also preprocesses the input images performing N4
-bias correction \citep{tustison2010} and intensity rescaling to $[0,1]$.  Note that 
-it is easily expandable to include 
-additional modalities and/or neighborhood and smoothing parameters.
-In Figure \ref{fig:featureImages}
-we provide sample mid-axial slices from the {\tt BRATS\_HG0301}
-data set from the Challenge cohort.
-
-We previously mentioned that all data was provided already 
-skull-stripped.  However, we note to the reader that an
-available ANTs-based solution to the brain extraction problem
-is found in the script {\tt antsBrainExtraction.sh} which uses
-a hybrid approach based on image registration and binary morphology
-to isolate the brain \cite{avants2010a}.  We recently performed an evaluation of over 1200
-MRI which demonstrated performance that is comparable or better than
-other available methodology \cite{tustison2013a}.
-
-\begin{figure*}[htb]
+We then use these output tissue map estimates as spatial 
+priors for generating a second set of image features.  These
+are used as input for a second RF model application (or
+second stage),
+the output of which constitute the final tumor segmentation estimate.
+Note that some feature images are included in both stages such as
+the asymmetry features described in the previous section.
+
+For both stages, in contrast to previous generative
+modeling approaches for multimodal tumor segmentation 
+\citep[e.g.,][]{prastawa2003}, we do not use multivariate 
+Gaussians to specify tissue probabilities but rather incorporate each
+univariate probability map into the feature vector of the training
+data.  As pointed out in \cite{menze2010}, multivariate modeling
+might obscure the distinct biological information provided by each 
+modality.  Instead, we let the RF construction 
+process determine the optimal combination of such multivariate
+information.  Additionally, maximum posterior labeling from both stages
+is used to determine the connected components for each label.  
+Geometric features (assigned voxel-wise) include the physical volumes 
+of each connected component, the volume to surface area ratio, 
+the elongation, and eccentricity.
+
+\paragraph{Stage 1 (GMM)}
+Gaussian mixture modeling (GMM) is used to assign voxelwise
+tissue probabilities which comprise an additional set of image features 
+where 
+prior cluster centers for specific tissue types for each
+modality are learned from training data \citep{reynolds2009}.  
+Intensity normalization
+across the training cohort was originally performed using 
+\cite{nyul2000}.  However, despite good performance on normal
+data, application of this intensity normalization algorithm 
+(after a thorough parameter search) on BRATS data led to slightly 
+worse performance than simply winsorizing the intensity values to the quantile
+range $[0.01, 0.99]$ and then rescaling the resulting intensities to 
+$[0,1]$ so we opted for the latter, more simplistic intensity
+normalization protocol.  The cluster centers are defined as the mean
+normalized intensity value for each tissue type of each modality 
+image over all the training data.  It should be noted that 
+initializing with these values is optional and that results
+seem to be robust to initialization values.
+
+A GMM formulates the 
+probability distribution at each voxel, $\mathbf{x}$, as the
+sum of $M$ Gaussian components, $\mathcal{N}(\mathbf{x}|\mu,\sigma)$, i.e.
+\begin{align}
+p\left(\mathbf{x}|\mu_m,\sigma_m,\lambda_m\right) = \sum_{i=1}^M \lambda_m \mathcal{N}(\mathbf{x}|\mu_m,\sigma_m)
+\end{align}
+where $\sum_{m=1}^M \lambda_m = 1$.  The parameters of the GMM 
+are determined with maximum likelihood estimation using the 
+Atropos segmentation tool \citep{avants2011} available in ANTs.  
+We include the resulting seven posterior probability tissue images 
+along with five additional geometric features based on the connected 
+components of the segmentation image (distance image to tumor core,
+physical volume image, volume to surface area ratio, eccentricity,
+and elongation).
+During the training phase (or RF model construction),
+these feature images, $\left\{F_i: i \in GMM\right\}$, generated during Stage 1 are used to create
+the Stage 1 RF model using the following relationship
+\begin{align}
+\label{eq:gmm}
+\mathrm{Tissue}\,\,\mathrm{classes} \sim \sum_{i \in GMM} F_i + \sum_{i \in Asym} F_i + \sum_{i \in Misc} F_i
+\end{align}
+given in the standard R notation of \cite{wilkinson1973}.  The
+set of $\left\{F_i: i \in Asym\right\}$ are the feature images 
+generated from the symmetric multivariate template 
+described in the previous section and $\left\{F_i: i \in Misc\right\}$
+are miscellaneous  feature images described in Section \ref{sec:misc}.
+
+\paragraph{Stage 2 (MAP-MRF)}
+
+Application of the Stage 1 RF model results in a set of
+probabilistic estimates for each of the seven tissue types.  
+At the beginning of the second stage, these spatial probability maps are 
+used as spatial priors in a 
+MAP-MRF segmentation also performed using the Atropos 
+segmentation tool \citep{avants2011} available in ANTs although we couple
+it with N4 bias correction \citep{tustison2010} as contained in the script
+{\tt antsAtroposN4.sh}.  The probabilistic segmentation output and corresponding
+connected-component feature images for each modality replace the analogous 
+GMM-based feature images used 
+during the first stage of training/prediction along with the other feature
+images generated during the feature stage (such as the template-based data).
+The second stage RF model is then produced similarly as the previous stage: 
+\begin{align}
+\label{eq:mapmrf}
+\mathrm{Tissue}\,\,\mathrm{classes} \sim \sum_{i \in MAP-MRF} F_i + \sum_{i \in Asym} F_i + \sum_{i \in Misc} F_i.
+\end{align}
+We discovered that this composite application of random
+forest models improves the resulting labeled output.
+
+\subsubsection{Miscellaneous Multimodal Feature Image Generation}
+\label{sec:misc}
+
+In addition to the feature images outlined in the previous sections, 
+several other feature images were generated largely based on what 
+had been used in previous work.  For each modality, several images
+were generated based on calculation of voxelwise neighborhood 
+first-order statistics including mean, variance, skewness, and entropy
+using neighborhood block radii
+of 1 and 3 voxels.  We also calculated two Euclidean distance
+transforms \citep{maurer2003} to be used for each subject.  One was
+calculated from the subject's own cerebral mask and the other was the 
+distance transform of the symmetric template cerebral mask warped
+to the subject.  Finally, we also generated the  (T1 - T1C) difference 
+image \citep{prastawa2003}.
+The script {\tt createFeatureImages.sh} performs all preprocessing
+and generates all feature images.  In Figure \ref{fig:featureImages} we provide sample mid-axial 
+image slices from features generated from the {\tt BRATS\_HG0301} data set.
+
+\begin{figure*}[h]
   \centering
-  \makebox[\textwidth][c]{
-  \includegraphics[width=180mm]{Figures/featureImages.pdf}
+  \makebox[0.9\textwidth][c]{
+  \includegraphics[width=150mm]{Figures/featureImages.pdf}
   }
@@
 
-\paragraph{Stage 1 (GMM)} \textcolor{red}{Need more descriptive
-  section name ... lots going on in the paper , easy to get disoriented}
-Prior cluster centers for specific tissue types learned from training data \citep{reynolds2009} are used in the GMM random forest stage to create 
-multiple feature images.  \textcolor{red}{FIXME I think this should be
-  expanded upon a bit ... can you be more specific?  how are these
-  cluster centers passed to the RF?  why are these centers portable to
-  this new data? etcetera.}
-Given $M$ tissue types, a GMM formulates the 
-probability distribution at each voxel, $\mathbf{x}$, as the
-sum of Gaussian components, $\mathcal{N}(\mathbf{x}|\mu,\sigma)$, i.e.
-\begin{align}
-p\left(\mathbf{x}|\mu_m,\sigma_m,\lambda_m\right) = \sum_{i=1}^M \lambda_m \mathcal{N}(\mathbf{x}|\mu_m,\sigma_m)
-\end{align}
-where $\sum_{m=1}^M \lambda_m = 1$.  One popular method for 
-determining the parameters of the GMM is maximum likelihood 
-estimation which can performed using the Atropos segmentation 
-tool \citep{avants2011} available in ANTs.  \textcolor{red}{FIXME
-  would prefer active voice version of the sentence to make clear that
-  Atropos uses EM} See Figure~\ref{fig:featureImages} for
-an example segmentation.
-
-\textcolor{red}{should there be an RF model equation here?  i.e. what
-  exactly (in R) was used ?}
-
-\paragraph{Stage 2 (MAP-MRF):}
-
-Prediction during the GMM stage results in a set of random 
-forest voting classification maps giving a probabilistic estimate
-of the label value at each voxel.  \textcolor{red}{FIXME ---i think
-  this sentence is confusing ... did you already run the RF at this
-  point?  seems no ... but ``results in a set of RF voting
-  classification maps'' sounds like you already have them.}  These spatial probability
-maps can be used as spatial priors in a MAP-MRF segmentation
-estimation protocol also performed using the Atropos segmentation 
-tool \citep{avants2011} available in ANTs although we couple
-it with iterative bias correction as contained in the script
-{\tt antsAtroposN4.sh}.  The probabilistic output (and corresponding
-feature images) 
-for each modality are used to replace the analogous GMM-based feature images used 
-during the first stage of prediction along with the other feature
-images previously generated (such as the template-based data).
-We discovered that this composite application of random
-forest models improves the resulting labeled output.
-
-\textcolor{red}{ok ... but i still don't understand this ... it gives
-  me a vague idea ... can you sketch out the equations for these
-  features/predictors? and give better intuition of what the
-  ``composite'' RF (previously referred to as concatenated?) is?}
-
-For both stages, in contrast to previous generative
-modeling approaches for multi-modal tumor segmentation 
-\citep[e.g.,][]{prastawa2003}, we do not use multivariate 
-Gaussians to specify tissue probabilities but rather incorporate each
-univariate probability map into the feature vector of the training
-data.  As pointed out in \cite{menze2010}, multivariate modeling
-might obscure the distinct biological information provided by each 
-modality.  Instead, we let the random forest construction 
-process determine the optimal combination of such multivariate
-information.  Additionally, maximum posterior labeling from both stages
-is used to determine the connected components for each label.  
-Geometric features (assigned voxel-wise) include the physical volumes 
-of each connected component, the volume to surface area ratio, 
-the elongation, and eccentricity.
-
-\paragraph{Neighborhood First-Order Statistic Feature Images}
-Similar to other work \cite{bauer2012,zikic2012} we calculated
-feature images using first order statistics such as mean, 
-variance, skewness, and entropy.  Images were created for 
-neighborhood radii of 1 and 3 voxels.
-
-\paragraph{Template-based Feature Images}
-Additional feature images are derived from the multivariate 
-symmetric template described earlier. The T1 images of the 
-subject and symmetric template are co-registered 
-\citep{tustison2013a} using the {\tt antsRegistration} 
-tool.  The motivation is similar in spirit to the method
-using reflection about the mid-sagittal plane described in 
-\cite{geremia2012}.  However, tumor growth often causes 
-distortion of the mid-plane thereby limiting the accuracy of
-this method.  By mapping to the symmetric template using the
-highly accurate SyN algorithm \cite{avants2011a}, we can 
-account for any distortion for comparing contralateral
-sides.
-
-Two images per modality use this derived mapping.  The modality-specific
-contralateral difference feature image is calculated by transforming
-the image to the symmetric template, flipping the image contralaterally,
-then transforming it back using the inverse transform, and subtracting
-the corresponding template component from the original image.  Additionally, 
-the template difference image is calculated by subtracting the warped template 
-image from the corresponding modality of the subject.
-
-\paragraph{Miscellaneous Feature Images}
-Three additional feature images are calculated which seem to aid in 
-classification as measured by the random forest variable importance
-(relative importance plots are provided in the Results section).
-First, the binary mask image is used to calculate the distance image 
-\citep{maurer2003}.  Second, from the symmetric template deformation 
-mapping we calculate the log Jacobian image.  Related to the earlier
-distance feature image, we calculate the distance mask image of the symmetric
-template and then warp that distance image to the individual subject
-for a third feature image.
-
-\subsection{Supervised Brain Tumor Segmentation with Concatenated Random Forests}
-\subsection{ANTsR:  An ANTs/R Interface}
-\subsection{MICCAI BRATS 2013 Brain Tumor Data}
-
-\clearpage
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 \subsection{ANTsR:  An ANTs/R Interface}
@@
 numerical and visualization software packages.
-The well-known neuroimaging de facto standard SPM%
+The well-known neuroimaging package SPM%
 \footnote{
@@
 for statistical computing and visualization thus providing a complete
-set of tools for multivariate neuroimage analysis ANTsR intends to provide a modern framework for medical analytics, with a focus on imaging-assisted prediction and statistical power.
+set of tools for multivariate neuroimage analysis.  
+ ANTsR intends to provide a modern framework for medical analytics, 
+ with a focus on imaging-assisted prediction and statistical power.
 
@@
 \verb#mFilter#, \verb#doParallel#, \verb#robust#, \verb#magic#, \verb#knitr#, \verb#pixmap#, 
-\verb#rgl#, and \verb#misc3d#.  
-See \href{http://stnava.github.io/software/2014/01/08/antsr/}{http://stnava.github.io/software/2014/01/08/antsr/} for current status. 
+\verb#rgl#, and \verb#misc3d#.%  
+\footnote{
+See \href{http://stnava.github.io/software/2014/01/08/antsr/}{http://stnava.github.io/software/2014/01/08/antsr/} for current status.}
 Additionally, in order
@@
 that ANTsR is currently only tested on UNIX-alikes such OSX and Ubuntu
-operating systems.  \footnote{Windows installation should be possible
-  but, to our knowledge, has not been attempted.}.
+operating systems.%  
+\footnote{Windows installation should be possible
+but, to our knowledge, has not been attempted.}
 
@@
 \begin{itemize}
-  \item {\tt applyModel.R}:  applies a random forest model to a new 
+  \item {\tt applyModel.R}:  applies a RF model to a new 
   feature data set from a testing subject resulting in a set of probability
@@
   organizes the file names in a csv file (via {\tt createCSVFileFromModel.R}),
-  and applies the random forest model using {\tt applyModel.R}. 
+  and applies the RF model using {\tt applyModel.R}. 
   \item {\tt applyTumorSegmentationModelForCohort.pl}:  Coordinates tumor 
@@
   a set of co-registered input MRI from a single subject. 
-  \item {\tt createModel.R}:  creates a random forest model given the input csv 
+  \item {\tt createModel.R}:  creates a RF model given the input csv 
   file of the feature image file names for all training data and set of truth
@@
   \item {\tt plotVariableImportance.R}:  produces a plot of the importance 
-  ({\tt MeanDecreaseAccuracy}) of each feature variable used 
+   of each feature variable used 
   in constructing the model.
@@
 We also include a fully functional 2-D example which performs
-both testing and training on sample challenge data.
-
-\textcolor{red}{FIXME---what is the first thing a user should do?  is
-  there a basic test verifying the 2D result?}
-
-\subsection{Supervised Brain Tumor Segmentation with Random Forests}
-
-Supervised techniques generally consist of a training phase
-for model construction followed by prediction using the 
-generated model.  For supervised brain tumor segmentation, 
-a set of training data consisting of labeled brain images 
-(e.g., csf, gray matter, white matter, 
-necrosis, edema, enhancing and  non-enhancing tumor) is
-used to construct a predictive model.  Although other 
-classification techniques have been used to segment
-brain tumors (e.g., support vector machines \citep{bauer2011}),
-random forest models have proven particularly successful.
-
-Several machine learning concepts were integrated to create 
-the random forests framework first articulated in its entirety by Breiman
-et al. \cite{breiman2001} for performing classification/regression.  
-Although decision trees had been previously explored in the literature, 
-it was the success of ``boosting''-style machine learning 
-techniques, such as AdaBoost \cite{schapire1990,freund1997}, which influenced 
-the aggregation of such decision trees into ``forests'' 
-with randomized node optimization for improved
-classification/regression performance \cite{ho1995,amit1997}.
-The final element of bootstrap aggregating or ``bagging'' (i.e.
-random sampling of the training data) was
-introduced by Breiman \cite{breiman1996} to achieve improved
-accuracy.%
-\footnote{
-One of the principal advantages of R is the extensive community of
-developers  who have contributed on the order of thousands of packages 
-extending R 's capabilities beyond its core functionality.
-Most relevant 
-is the {\tt randomForest} package developed from Breiman's original
-Fortran code by Liaw and Wiener \citep{liaw2002}.
-}
-
-Early adoption \cite{viola2005} and success in the
-computer vision community
-has led to a recent surge within the medical image analysis
-community of using random forests for handling complex 
-classification/regression tasks including
-normal brain segmentation \cite{yi2009},
-MS lesion segmentation \cite{geremia2011}, 
-multimodal brain tumor segmentation
-\cite{bauer2012,zikic2012}, brain extraction \cite{iglesias2010}, 
-anatomy detection in computed tomography \cite{criminisi2013}, and
-segmentation of echocardiographic images \cite{verhoek2011}. 
-A thorough introduction for those interested in delving deeper 
-into the more theoretical aspects of random forests can be found
-in \cite{criminisi2011}.
-
-%In subsequent sections, we detail the generic
-%random forest modeling framework which takes as input
-%a set of feature images (also described in a later section)
-%and labels for each training subject 
-%and outputs a statistical model.  This model, in turn, can then
-%be used to provide a probabilistic estimate of the labels in an
-%unlabeled subject.
-
-One of the innovations that we provide in this work is
-the use of \textit{concatenated random forests} \textcolor{red}{where
-  does the name come from?  it doesnt get mentioned again until the
-  discussion section ... expanding on this / defining it could help
-  ``novelty'' angle in neuroimage submission .... FIXME} for improved probabilistic estimation
-of the brain segmentation labels.  Briefly, Gaussian mixture
-modeling (GMM) is used to create a subset of feature images which are
-used to train and predict during the first classification stage.  
-The random forest voting output probabilistic maps are then used as 
-spatial probability priors for performing a MAP-MRF segmentation
-and subsequent feature image creation
-thereby replacing the GMM-derived feature images for a second
-classification stage.  This dual random forest model generation
-and application
-improves results and is detailed in later sections.
-
-%As we demonstrate in the Results section, the set of feature
-%images employed work sufficiently well for good performance
-%on the training data.  However, we discovered
-%that further improvements could be gained by using the probabilistic 
-%label estimates as input to enhance preprocessing before a 
-%second round of feature image generation including 
-%modality-specific, prior-based $n$-tissue segmentation and 
-%random forest model creation.
-
-\subsubsection{Brain Tumor Data}
+both testing and training on sample challenge data.%
+\footnote{
+https://github.com/ntustison/ANTsAndArboles/tree/master/SimpleExample
+}
+After pulling the repository, one can run the scripts {\tt exampleTrain.sh}
+and {\tt examplePredict.sh} to get the sample results.  Output includes
+several overlap measures describing the performance.
+
+\subsection{Brain Tumor Data}
 
@@
 
-Training data consisted of multi-modal brain MRI (T1, T2, FLAIR, and 
+Training data consisted of multimodal brain MRI (T1, T2, FLAIR, and 
 post-Gadolinium T1) from 30 glioma patients (both low, $n=10$, and high-grade, $n=20$,
@@
 \footnote{
-http://martinos.org/qtim/miccai2013/data.html.
+http://martinos.org/qtim/miccai2013/data.html
 }
@@
 
-%\subsubsection{Multi-Modal Feature Image Preprocessing}
-%Although several studies have pointed out the importance of
-%intensity normalization and bias correction, our experience 
-%with the training data illustrated a degradation in performance
-%when one or both steps (using \cite{nyul2000} and N4 \cite{tustison2010},
-%respectively) were performed due to the presence of the tumor/edema complex. 
-% 
-%As a corrective, for the first stage we simply windowed the image intensity
-%for all images to be between the quantiles $[0.01,0.99]$ and
-%subsequently rescaled to $[0,1]$.  From these ``corrected'' images,
-%the first set of feature images were derived.  For the training cohort, these
-%data were used to create the random forest regression model for the first
-%stage.  During the second stage, the probabilistic estimates
-%of the white matter and gray matter labels were used to generate a
-%``pure tissue weight mask'' to estimate the bias field 
-%using N4 (although the resulting bias field estimation was used
-%to correct the image within the entire cerebral mask).  Formally, this 
-%involved generation of a probabilistic map defined as:
-%\begin{align}
-%  P_{pure\,\,tissue}(\mathbf{x}) = \sum_{i=1}^N P_i(\mathbf{x}) \prod_{j=1, j \neq i}^N \left( 1 - P_j(\mathbf{x}) \right)
-%\end{align}
-%where $N$ is the set of user-selected tissue labels (in our
-%case $N=2$ consisting of the gray and white matter probability
-%maps).
-%
-%Both rescaling and weighted bias correction were applied to produce
-%the ``corrected images'' for the second stage resulting in
-%modified features images for the second stage.  Note that we
-%perform a similar iterative scenario for normal brain 
-%segmentation \cite{avants2011} (encapsulated in the ANTs script 
-%\verb#antsAtroposN4.sh#).
-
-\subsubsection{Symmetric Multivariate Template Construction}
-
-\subsubsection{Training:  Random Forest Model Creation}
+\subsubsection{Training: RF Creation for the BRATS 2013 Challenge}
 
@@
 high-grade glioma) for both GMM and MAP-MRF stages were created using only the 
-supplied training data.  Prior to training, we segmented normal brain tissue \cite{avants2011}
+supplied training data.  Prior to training, we segmented normal brain tissue \citep{avants2011}
 for each data set by segmenting only the T1 image.  This was only to yield
 a rough estimate of normal brain tissue to augment the already provided 
-pathology labels.  This resulted in seven labels (csf, gray matter, white matter,
-necrosis, edema, enhancing, and non-enhancing tumor)
-characterizing each brain.
+pathology labels.  This resulted in seven labels for tissues described
+earlier i.e., csf, gray matter, white matter,
+necrosis, edema, enhancing, and non-enhancing tumor characterizing each brain.
 
@@
 number of trees, number of samples per label, and number of threads for parallel
-processing.  The output is an {\tt RData} file describing the random forest
+processing.  The output is an {\tt RData} file describing the RF
 model which can be used for future predictions.
@@
 }
-
-\textcolor{red}{do you actually use variable importance or no?
-  i.e. to trim the number of features?  overall, having a hard time
-  seeing what is possible versus what was actually done ... }
-
-\subsubsection{Prediction:  Applying the Random Forest Models}
-
-\begin{figure}
-  \centering
-  \begin{tabular}{c}
-    \includegraphics[width=90mm]{Figures/pipeline.pdf}
-  \end{tabular}
-  \caption{Diagrammatic workflow for the proposed random forest model prediction.  The feature
-  images are first generated from the input MRI and symmetric multivariate template. 
-  After organization of the feature images in a csv file, the stage 1 random forest model is 
-  applied which results in a first stage classification voting output.  This output
-  is used to generate an additional set of feature images for the second classification
-  stage.  The final refinement process entails a series of binary morphological 
-  operations which results in the final labeled brain.  In the diagram we also list
-  the files used to perform each set of functions which are available as open source.
-  }
-  \label{fig:pipeline}
-\end{figure}
+In this work, we do not use these measurements for feature pruning.  However,
+we plot them in the Results section (see Figures \ref{fig:hgimportance} and
+\ref{fig:lgimportance}) as they demonstrate the relative importance of our
+selected features including that of the proposed asymmetry images.
+
+\subsubsection{Prediction:  Applying the RFs for the BRATS 2013 Challenge}
 
@@
 as illustrated in Figure \ref{fig:pipeline}.  From the feature images and input 
-GMM model, a tentative set of random forest voting output confidence images are produced.
+GMM model, a tentative set of RF voting output confidence images are produced.
 As described, this is used as input to the second prediction round.  The 
-final probability output images are used to produce the maximum labeling.
+final probability output images are used to produce the maximum probability labeling.
 
@@
 components and morphological closure of certain regions.
-All these steps are included in the perl script 
+All steps are included in the script 
 {\tt applyTumorSegmentationModelForCohort.pl} designed for parallel subject 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%
+% Results
+%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \section{Results}
 
+\begin{figure}[!htb]
+  \centering
+  \includegraphics[width=120mm]{Figures/challengeResults.pdf}
+  \caption{Visual results (every tenth slice sampled from the axial center of the
+  tumor region) from the BRATS 2013 challenge using the proposed
+  framework (off-white = non-enhancing tumor, cyan = edema, violet = enhancing 
+  tumor, yellow = necrosis).  Tumor components were combined into
+  three regions for performance assessment ((1) complete tumor: all four tissue
+  classes; (2) tumor core:  necrosis, non-enhancing tumor, and enhancing tumor;
+  and (3) enhancing tumor).  See Table \ref{table:results} for quantitative measures
+  and challenge rankings.
+  }
+  \label{fig:challengeresults}
+\end{figure}
+
+%\begin{table*}
+%\caption{Results from the MICCAI 2013 BRATS Challenge}
+%\label{table:results}
+%\begin{center}
+%\begin{tabular*}{0.99\textwidth}{@{\extracolsep{\fill} } c c c c c c c c c c c}
+%\toprule
+%{\bf Data set} & {\bf Rank} & \multicolumn{3}{c}{\bf Dice} & \multicolumn{3}{c}{ \bf Positive Predictive Value} & \multicolumn{3}{c}{ \bf Sensitivity} \\
+%{} & {} & complete & core & enhanced & complete & core & enhanced & complete & core & enhanced \\
+%\midrule
+%Challenge & 1 & 0.87 (1) & 0.78 (1) & 0.74 (1) & 0.85 (2) & 0.74 (5) & 0.69 (4) & 0.89 (2) & 0.88 (1) & 0.83 (1) \\
+%Leaderboard & 1 & 0.79 (2) & 0.65 (1) & 0.53 (3) & 0.83 (1) & 0.70 (2) & 0.51 (3) & 0.81 (4) & 0.73 (2) & 0.66 (2) \\
+%Evaluation & 2 & 0.88 (2) & 0.76 (2) & 0.55 (3) & 0.88 (3) & 0.80 (5) & 0.65 (3) & 0.89 (3) & 0.79 (3) & 0.53 (3) \\
+%\bottomrule
+%\end{tabular*}
+%\end{center}
+%\end{table*}
+
 \begin{table*}
-\caption{Results from the MICCAI 2013 BRATS Challenge}
+\caption{Assessment measures from the MICCAI 2013 BRATS Challenge including
+ Dice overlap, positive prediction value (PPV), and sensitivity.}
 \label{table:results}
 \begin{center}
-\begin{tabular*}{0.99\textwidth}{@{\extracolsep{\fill} } c c c c c c c c c c c}
+\begin{tabular*}{0.99\textwidth}{@{\extracolsep{\fill} } l c c c c}
 \toprule
-{\bf Data set} & {\bf Rank} & \multicolumn{3}{c}{\bf Dice} & \multicolumn{3}{c}{ \bf Positive Predictive Value} & \multicolumn{3}{c}{ \bf Sensitivity} \\
-{} & {} & complete & core & enhanced & complete & core & enhanced & complete & core & enhanced \\
+\multicolumn{1}{c}{ \bf Data Set} & {\bf Rank}$^\dagger$ & {\bf Dice} & {\bf PPV} & {\bf Sensitivity} \\
 \midrule
-Challenge & 1 & 0.87 (1) & 0.78 (1) & 0.74 (1) & 0.85 (2) & 0.74 (5) & 0.69 (4) & 0.89 (2) & 0.88 (1) & 0.83 (1) \\
-Leaderboard & 1 & 0.79 (2) & 0.65 (1) & 0.53 (3) & 0.83 (1) & 0.70 (2) & 0.51 (3) & 0.81 (4) & 0.73 (2) & 0.66 (2) \\
-Evaluation & 2 & 0.88 (2) & 0.76 (2) & 0.55 (3) & 0.88 (3) & 0.80 (5) & 0.65 (3) & 0.89 (3) & 0.79 (3) & 0.53 (3) \\
+Challenge (complete) & 1 & 0.87 (1) & 0.85 (2) & 0.89 (2) \\
+Challenge (core) & {} & 0.78 (1) & 0.74 (5) & 0.88 (1) \\
+Challenge (enhanced) & {} & 0.74 (1) & 0.69 (4) & 0.83 (1) \\
+\midrule
+Leaderboard (complete) &  1 & 0.79 (2) & 0.83 (1) & 0.81 (4) \\
+Leaderboard (core)     & {} & 0.65 (1) & 0.70 (2) & 0.73 (2) \\
+Leaderboard (enhanced) & {} & 0.53 (3) & 0.51 (3) & 0.66 (2) \\
+\midrule
+Evaluation (complete) &  2 & 0.88 (2) & 0.88 (3) & 0.89 (3) \\
+Evaluation (core)     & {} & 0.76 (2) & 0.80 (5) & 0.79 (3) \\
+Evaluation (enhanced) & {} & 0.55 (3) & 0.65 (3) & 0.53 (3) \\
 \bottomrule
+\multicolumn{5}{l}{\scriptsize$^\dagger$Indicates performance
+for all measures over all regions.  Component rankings are 
+given in parentheses.
+}
 \end{tabular*}
@@
 
-A total of four random forest models were created from the 30 training data 
+A total of four RF models were created from the 30 training data 
 sets. Two models for Stage 1 and Stage 2 processing were generated from the 
@@
   \caption{{\tt MeanDecreaseAccuracy} plots generated from the high-grade glioma
-  Stage 1 and Stage 2 random forest models.  These plots provide a weighted 
+  Stage 1 and Stage 2 RF models.  These plots provide a weighted 
   ranking describing the importance of each feature for predictive accuracy. 
@@
   \caption{{\tt MeanDecreaseAccuracy} plots generated from the low-grade glioma
-  Stage 1 and Stage 2 random forest models.  These plots provide a weighted 
+  Stage 1 and Stage 2 RF models.  These plots provide a weighted 
   ranking describing the importance of each feature for predictive accuracy.
@@
 glioma class models.
-
-\textcolor{red}{how novel is this a/symmetry stuff?  i have the chimp
-  paper that briefly mentions it but could this be spun as a
-  ``contribution'' (with additional technical detail) in addition to pointing out the benefit of asymmetry
-  plus RF's?  also see
-  \href{http://stnava.github.io/asymmetry/}{http://stnava.github.io/asymmetry/}
-  which is sort of novel ... }
 
@@
 
-Although previous research has employed random forests for supervised brain
-segmentation, our contribution of \textcolor{red}{concatenated model} application and use
+One of the difficulties with the focus of this particular competition 
+is the lack of consensus, even clinically, as to tumor type and extent
+in the context of medical imaging \citep{cha2005}.  In fact, current 
+consensus guidelines from the World Health Organization for brain 
+tumor classification are strictly histopathological \citep{louis2007}
+which limit clinical application.  Such limitations motivate the
+use of medical imaging for treatment planning and outcome analyses 
+\citep{cha2005} including more automated methods such as that proposed
+in this work.
+
+Although previous research has employed RFs for supervised brain
+segmentation, our contribution of concatenated RF model application and use
 of symmetric multivariate templates demonstrated good performance 
@@
 of the competition.  This confirmed what others have found in that
-random forests provide an excellent framework for prediction in certain
+RFs provide an excellent framework for prediction in certain
 medical image analysis problems.  Even with relatively few training
-subjects relative to input variables, the random forest models
-perform generalizably.
+subjects relative to input variables, the RF models
+perform well.
 In addition, this work highlights the value of  a symmetric multiple
@@
 multivariate data.
-
-\textcolor{red}{\subsection{Discussion of clinical relevance of the multivariate
-  features we ``found''}}
 
@@
 %\section*{Acknowledgments}
-
+\clearpage
 \section*{References}

### 2014-02-25 Figures/ChallengeResults/makeFigures.sh

@@ -1,0 +1,66 @@
+
+# for(( j=1; j < 10; j++ ))
+#   {
+#   echo $j
+#   ConvertScalarImageToRGB 3 RefinedLabelsChallenge/BRATS_HG030${j}_REFINE_RF_LABELS.nii.gz RefinedLabelsChallenge/BRATS_HG030${j}_REFINE_RF_LABELS.mha none custom ~/Pkg/ANTs/Examples/CustomColormaps/itkSnap.txt
+#   ThresholdImage 3 RefinedLabelsChallenge/BRATS_HG030${j}_REFINE_RF_LABELS.nii.gz RefinedLabelsChallenge/BRATS_HG030${j}_REFINE_RF_LABELS_MASK.nii.gz 0 0 0 1
+#   }
+# ConvertScalarImageToRGB 3 RefinedLabelsChallenge/BRATS_HG0310_REFINE_RF_LABELS.nii.gz RefinedLabelsChallenge/BRATS_HG0310_REFINE_RF_LABELS.mha none custom ~/Pkg/ANTs/Examples/CustomColormaps/itkSnap.txt
+# ThresholdImage 3 RefinedLabelsChallenge/BRATS_HG0310_REFINE_RF_LABELS.nii.gz RefinedLabelsChallenge/BRATS_HG0310_REFINE_RF_LABELS_MASK.nii.gz 0 0 0 1
+
+GEOM='-1x6'
+
+CreateTiledMosaic -i NiftiChallenge/BRATS_HG0301/BRATS_HG0301_FLAIR.nii.gz -r RefinedLabelsChallenge/BRATS_HG0301_REFINE_RF_LABELS.mha -a 1.0 -o RefinedLabelsChallenge/BRATS_HG0301_REFINE_RF_LABELS.png -t $GEOM -d 2 -s [7,40,78] -x RefinedLabelsChallenge/BRATS_HG0301_REFINE_RF_LABELS_MASK.nii.gz
[…]

### 2014-03-14 paper.tex
WIP: mod to abstract

@@
 
-\title{Optimal Symmetric Multimodal Templates and Concatenated Random Forests for Supervised Brain Tumor Segmentation (Simplified) with ANTsR}
+\title{Optimal Symmetric Multimodal Templates and Concatenated Random Forests for Supervised Brain Tumor Segmentation (Simplified) with \textit{ANTsR}}
 
@@
 \begin{abstract} 
-Automated brain tumor segmentation in MRI is complicated by the lack of prior knowledge 
-concerning tumor location, spatial extent, shape, possible displacement of normal 
-tissue, and intensity signature, all of which limit 
-the applicability of well-established techniques that have demonstrated good performance 
-in related tasks such as normal brain tissue segmentation.  Due to this 
-limitation, we localize our prior knowledge strategy through the 
-use of subject-specific multimodal asymmetrical features which demonstrate excellent
-discriminative properties within the random forest machine learning framework.  
-This approach is made possible with symmetric multimodal brain templates which
-are optimal in terms of both shape and multivariate intensity.
-Further performance improvements are demonstrated by a successive application of random forest models for increasing levels of feature refinement
- to which we refer as {\it concatenated random forests}.
-Although the derived workflow involves many disparate components, the newly developed ANTsR open source software package significantly facilitates the proposed methodology.
-ANTsR, a comprehensive interface of the popular Advanced Normalization Tools (ANTs) repository with the R statistical project,
-integrates well-performing image analysis tools such as image registration, segmentation, bias correction, and template building with access to advanced statistical and machine learning techniques as well as state-of-the-art visualization possibilities.
-The reported algorithmic framework was the top-performing entry in the MICCAI 2013 Multimodal Brain Tumor Segmentation challenge and, to our knowledge, is the only entry for both challenge years (2012 \& 2013) which has been made publicly available.
+Manual brain tumor segmentation in MRI is time consuming and
+inconsistent both longitudinally and across different raters.
+Automated methods have the promise of resolving these issues but are
+complicated by the lack of prior knowledge concerning tumor location,
+spatial extent, shape, possible displacement of normal tissue, and
+intensity signature.  These factors all limit the applicability of
+traditional segmentation techniques that have demonstrated good
+performance in related tasks such as normal brain tissue segmentation.
+The multi-year BRATS challenge seeks to
+resolve these issues via high-quality labeled data which may be used
+to both guide prior-based segmentation algorithms and also to assess performance in a
+blinded, international competition.  In this work, we present the
+theoretical and practical background that represents the winning
+team's entry into the 2013 BRATs challenge.  This winning approach
+combined the features of high-performance methods in both medical
+image processing and in machine learning within a new prior-based and automated
+tumor segmentation protocol.  This approach employs template-based
+deformation models, prior-based segmentation and image geometry
+measurements to compute rich, subject-specific multiple modality feature sets which drive a random
+forest (RF) prediction model.   The RF model demonstrates excellent
+classification performance and also, via variable importance ranking,
+identifies key discriminative properties from the rich input feature
+set.  Some of the most salient features are those that quantify
+asymmetry across different imaging modalities.  We describe how we
+employ symmetric multimodal brain templates to derive these
+performance critical features and how these features are used within
+successive applications of random forest models for increasing levels
+of feature refinement, i.e.
+\textit{concatenated random forests}.
+Although the derived workflow involves many components, the newly developed \textit{ANTsR} open source software package significantly facilitates the proposed methodology.
+\textit{ANTsR}
+integrates Advanced Normalization Tools (ANTs) with \textit{R}'s
+cutting-edge statistical and machine learning techniques and state-of-the-art visualization possibilities.  
+The reported algorithmic framework was the top-performing entry in the MICCAI 2013 Multimodal Brain Tumor Segmentation challenge and, to our knowledge, is the only entry for both challenge years (2012 and 2013) which has been made publicly available.
 \end{abstract}
@@
 } 
-of the National Institutes of Health and is available through our ANTsR 
+of the National Institutes of Health and is available through our \textit{ANTsR} 
 package.
@@
 In subsequent sections, we describe these items and detail the
-ANTsR framework which coordinates all steps.
+\textit{ANTsR} framework which coordinates all steps.
 
@@
 \begin{figure*}[h]
-  \centering
-  \makebox[0.9\textwidth][c]{
+%  \centering
+%  \makebox[0.9\textwidth][c]{
   \includegraphics[width=150mm]{Figures/featureImages.pdf}
-  }
+%  }
   \caption{Representative feature images derived from the 
@@
 
-\subsection{ANTsR:  An ANTs/R Interface}
+\subsection{\textit{ANTsR}:  An ANTs/R Interface}
 
@@
 that robust statistical machinery was lacking for making inferences regarding
-the data produced during the course of ANTs processing.  ANTsR was developed
+the data produced during the course of ANTs processing.  \textit{ANTsR} was developed
 specifically to provide an interface between ANTs, a 
@@
 set of tools for multivariate neuroimage analysis.  
- ANTsR intends to provide a modern framework for medical analytics, 
+ \textit{ANTsR} intends to provide a modern framework for medical analytics, 
  with a focus on imaging-assisted prediction and statistical power.
@@
 led to the adoption of R to complement ANTs quantification resulting in the
-ANTsR package.
+\textit{ANTsR} package.
 %The R project for statistical computing,%
@@
 
-The ANTsR package is publicly available on the github project hosting service.%
-\footnote{
-\href{https://github.com/stnava/ANTsR}{https://github.com/stnava/ANTsR}
-}
-Prior to installation of ANTsR, several external R packages
+The \textit{ANTsR} package is publicly available on the github project hosting service.%
+\footnote{
+\href{https://github.com/stnava/\textit{ANTsR}}{https://github.com/stnava/\textit{ANTsR}}
+}
+Prior to installation of \textit{ANTsR}, several external R packages
 need to be installed including: \verb#Rcpp#, \verb#signal#, \verb#timeSeries#, 
@@
 and ANTs.  Further instructions for download and
-installation can be found on the ANTsR github website.  Feel
+installation can be found on the \textit{ANTsR} github website.  Feel
 free to contact the authors if installation trouble occurs.  We note
-that ANTsR is currently only tested on UNIX-alikes such OSX and Ubuntu
+that \textit{ANTsR} is currently only tested on UNIX-alikes such OSX and Ubuntu
 operating systems.%  
@@
 \subsubsection{Usage}
-ANTsR is intended to not only allow easy interchange between
+\textit{ANTsR} is intended to not only allow easy interchange between
 medical imaging formats and R but also to facilitate
@@
 
-An additional motivation for our development of ANTsR (and
+An additional motivation for our development of \textit{ANTsR} (and
 hopefully its acceptance by the community) 
@@
 As mentioned earlier, we have made this entire framework
-available as open source.  In addition to the ANTsR repository
-already on github which houses both ANTs and ANTsR functionality, 
+available as open source.  In addition to the \textit{ANTsR} repository
+already on github which houses both ANTs and \textit{ANTsR} functionality, 
 we created a special github repository specifically for this work
@@
 Also, we posted all
-scripts (R, shell, and perl) used to coordinate the ANTsR processing 
+scripts (R, shell, and perl) used to coordinate the \textit{ANTsR} processing 
 including:
@@
 methodology that we have made available is only a small part of the 
-larger software package that we have created in ANTsR.  Not only does
-ANTsR significantly facilitate the development of the work discussed,
+larger software package that we have created in \textit{ANTsR}.  Not only does
+\textit{ANTsR} significantly facilitate the development of the work discussed,
 but it provides an interface to one of the most powerful statistical
@@
 \bibliography{references}
-
-%% Authors are advised to submit their bibtex database files. They are
-%% requested to list a bibtex style file in the manuscript if they do
-%% not want to use model1-num-names.bst.
-
-%% References without bibTeX database:
-
-% \begin{thebibliography}{00}
-
-%% \bibitem must have the following form:
-%%   \bibitem{key}...
-%%
-
-% \bibitem{}
-
-% \end{thebibliography}
-
 \end{document}
-
-%%
-%% End of file `elsarticle-template-1-num.tex'.

### 2014-03-14 paper.tex
BUG:  Remove italics from web address.

@@
 successive applications of random forest models for increasing levels
-of feature refinement, i.e.
-\textit{concatenated random forests}.
-Although the derived workflow involves many components, the newly developed \textit{ANTsR} open source software package significantly facilitates the proposed methodology.
-\textit{ANTsR}
+of feature refinement, i.e. \textit{concatenated random forests}.  Although the derived 
+workflow involves many components, the newly developed \textit{ANTsR} open source software 
+package significantly facilitates the proposed methodology.  \textit{ANTsR}
 integrates Advanced Normalization Tools (ANTs) with \textit{R}'s
 cutting-edge statistical and machine learning techniques and state-of-the-art visualization possibilities.  
-The reported algorithmic framework was the top-performing entry in the MICCAI 2013 Multimodal Brain Tumor Segmentation challenge and, to our knowledge, is the only entry for both challenge years (2012 and 2013) which has been made publicly available.
+The reported algorithmic framework was the top-performing entry in the MICCAI 2013 Multimodal Brain Tumor 
+Segmentation challenge and, to our knowledge, is the only entry for both challenge years (2012 and 2013) 
+which has been made publicly available.
 \end{abstract}
@@
 set of template-based feature images is generated per modality 
-as the difference image with the contralateral reflection.  This is achieved by calculating the reflection transform, $T_R$, in the symmetric template space and composing transforms
+as the difference image with the contralateral reflection.  This is achieved by calculating the 
+reflection transform, $T_R$, in the symmetric template space and composing transforms
 as follows to create the non-Euclidean contralateral counterpart per modality:
@@
 \footnote{
-\href{https://github.com/stnava/\textit{ANTsR}}{https://github.com/stnava/\textit{ANTsR}}
+\href{https://github.com/stnava/ANTsR}{https://github.com/stnava/ANTsR}
 }

### 2014-04-01 paper.tex
WIP: minor abstract edit
ENH: another abstract revison

--- 2014-02-25 paper.tex
@@
 \begin{abstract} 
-Automated brain tumor segmentation in MRI is complicated by the lack of prior knowledge 
-concerning tumor location, spatial extent, shape, possible displacement of normal 
-tissue, and intensity signature, all of which limit 
-the applicability of well-established techniques that have demonstrated good performance 
-in related tasks such as normal brain tissue segmentation.  Due to this 
-limitation, we localize our prior knowledge strategy through the 
-use of subject-specific multimodal asymmetrical features which demonstrate excellent
-discriminative properties within the random forest machine learning framework.  
-This approach is made possible with symmetric multimodal brain templates which
-are optimal in terms of both shape and multivariate intensity.
-Further performance improvements are demonstrated by a successive application of random forest models for increasing levels of feature refinement
- to which we refer as {\it concatenated random forests}.
-Although the derived workflow involves many disparate components, the newly developed ANTsR open source software package significantly facilitates the proposed methodology.
-ANTsR, a comprehensive interface of the popular Advanced Normalization Tools (ANTs) repository with the R statistical project,
-integrates well-performing image analysis tools such as image registration, segmentation, bias correction, and template building with access to advanced statistical and machine learning techniques as well as state-of-the-art visualization possibilities.
-The reported algorithmic framework was the top-performing entry in the MICCAI 2013 Multimodal Brain Tumor Segmentation challenge and, to our knowledge, is the only entry for both challenge years (2012 \& 2013) which has been made publicly available.
+Automated brain tumor segmentation in MRI is complicated by the lack
+of prior knowledge concerning tumor location, spatial extent, shape,
+possible displacement of normal tissue, and intensity signature.
+These factors limit the applicability of techniques with established
+performance in related tasks such as normal brain tissue segmentation.
+We introduce a template-based framework for computing geometric and
+asymmetry-driven feature sets from multiple modality neuroimaging.  We
+use these features to drive a supervised whole-brain and tumor segmentation
+approach based on random forests.  The asymmetry-related features 
+demonstrate excellent discriminative properties within this framework. 
+This approach is made possible by symmetric multimodality brain templates which are optimal
+in terms of both shape and multivariate intensity.  Further
+performance improvements are demonstrated by a successive application
+of random forest models for increasing levels of feature refinement to
+which we refer as {\it concatenated random forests}.  Although the
+derived workflow involves many disparate components, the newly
+developed \textit{ANTsR} open source software package significantly facilitates
+the proposed methodology.  \textit{ANTsR} is a comprehensive interface between the
+popular Advanced Normalization Tools (ANTs) and the \textit{R}
+statistical project. \textit{ANTsR} integrates proven image analysis methods
+for image registration, segmentation, bias correction, and
+template building with \textit{R}'s statistical and machine
+learning techniques and state-of-the-art visualization possibilities.  
+The reported algorithmic framework was the
+top-performing entry in the MICCAI 2013 Multimodal Brain Tumor
+Segmentation challenge and, to our knowledge, is the only entry for
+both challenge years (2012 \& 2013) which has been made publicly
+available.
 \end{abstract}

### 2014-04-01 paper.tex

--- 2014-03-14 paper.tex
@@
 \begin{abstract} 
-Manual brain tumor segmentation in MRI is time consuming and
-inconsistent both longitudinally and across different raters.
-Automated methods have the promise of resolving these issues but are
-complicated by the lack of prior knowledge concerning tumor location,
-spatial extent, shape, possible displacement of normal tissue, and
-intensity signature.  These factors all limit the applicability of
-traditional segmentation techniques that have demonstrated good
+Automated brain tumor segmentation in MRI is complicated by the lack
+of prior knowledge concerning tumor location, spatial extent, shape,
+possible displacement of normal tissue, and intensity signature.
+These factors limit the applicability of techniques with established
 performance in related tasks such as normal brain tissue segmentation.
-The multi-year BRATS challenge seeks to
-resolve these issues via high-quality labeled data which may be used
-to both guide prior-based segmentation algorithms and also to assess performance in a
-blinded, international competition.  In this work, we present the
-theoretical and practical background that represents the winning
-team's entry into the 2013 BRATs challenge.  This winning approach
-combined the features of high-performance methods in both medical
-image processing and in machine learning within a new prior-based and automated
-tumor segmentation protocol.  This approach employs template-based
-deformation models, prior-based segmentation and image geometry
-measurements to compute rich, subject-specific multiple modality feature sets which drive a random
-forest (RF) prediction model.   The RF model demonstrates excellent
-classification performance and also, via variable importance ranking,
-identifies key discriminative properties from the rich input feature
-set.  Some of the most salient features are those that quantify
-asymmetry across different imaging modalities.  We describe how we
-employ symmetric multimodal brain templates to derive these
-performance critical features and how these features are used within
-successive applications of random forest models for increasing levels
-of feature refinement, i.e. \textit{concatenated random forests}.  Although the derived 
-workflow involves many components, the newly developed \textit{ANTsR} open source software 
-package significantly facilitates the proposed methodology.  \textit{ANTsR}
-integrates Advanced Normalization Tools (ANTs) with \textit{R}'s
-cutting-edge statistical and machine learning techniques and state-of-the-art visualization possibilities.  
-The reported algorithmic framework was the top-performing entry in the MICCAI 2013 Multimodal Brain Tumor 
-Segmentation challenge and, to our knowledge, is the only entry for both challenge years (2012 and 2013) 
-which has been made publicly available.
+We introduce a template-based framework for computing geometric and
+asymmetry-driven feature sets from multiple modality neuroimaging.  We
+use these features to drive a supervised whole-brain and tumor segmentation
+approach based on random forests.  The asymmetry-related features 
+demonstrate excellent discriminative properties within this framework. 
+This approach is made possible by symmetric multimodality brain templates which are optimal
+in terms of both shape and multivariate intensity.  Further
+performance improvements are demonstrated by a successive application
+of random forest models for increasing levels of feature refinement to
+which we refer as {\it concatenated random forests}.  Although the
+derived workflow involves many disparate components, the newly
+developed \textit{ANTsR} open source software package significantly facilitates
+the proposed methodology.  \textit{ANTsR} is a comprehensive interface between the
+popular Advanced Normalization Tools (ANTs) and the \textit{R}
+statistical project. \textit{ANTsR} integrates proven image analysis methods
+for image registration, segmentation, bias correction, and
+template building with \textit{R}'s statistical and machine
+learning techniques and state-of-the-art visualization possibilities.  
+The reported algorithmic framework was the
+top-performing entry in the MICCAI 2013 Multimodal Brain Tumor
+Segmentation challenge and, to our knowledge, is the only entry for
+both challenge years (2012 \& 2013) which has been made publicly
+available.
 \end{abstract}

### 2014-04-01 paper.tex
WIP... small mods

@@
 described in the corresponding publications, lack of publicly available 
-evaluation data, and privately kept algorithmic instantiations.  
+evaluation data, and private algorithmic instantiations.  
 In response to these issues, the Multimodal Brain Tumor Segmentation 
@@
 needs \citep{durst2014}, we found that no teams had availed their software in any 
-form which prompted our participation in the challenge which took place
+form.  This prompted our participation in the challenge which took place
 the following year in Nagoya, Japan.  Heavily influenced by the success 
@@
 random forest (RF) machine learning framework \citep{breiman2001} which 
-permits the inclusion of potentially discriminative image features.  
+permits the inclusion of many potentially discriminative image features.  
 Supervised segmentation techniques generally consist of a training phase

### 2014-04-01 paper.tex
WIP: mild comments

@@
 are then refined using a series of heuristically-derived binary morphological
-operations.
+operations.  \textcolor{red}{The two-stage approach is motivated by
+the fact that the RF is a powerful tool for ranking the features
+in a high-dimensional model i.e. can be used for model selection.
+However, the RF outputs are not geometrically constrained, which is
+the motivation for our second refinement step.}.
 
@@
 $\mathbf{J}=\{J_1,J_2,\ldots, J_K\}$ to minimize the
-following cost function:
+following cost function \textcolor{red}{not sure the 2nd term is correct...}:
 \begin{align}
@@
 the asymmetry features described in the previous section.
+\textcolor{red}{Nick, I think the motivation for this is still unclear
+  .... doesnt it have something to do with the MRF of atropos
+  smoothing out the importance features resulting from the RF step?
+  Figure 4 tells us a little bit about the procedure of the code but
+  the idea is not conveyed ... would it be possible to show exactly
+  what is happening in a separate figure?  e.g. just conceptually in a small 
+  region of voxels with 2 modalities ....?}
 
@@
 
-\paragraph{Stage 1 (GMM)}
+\paragraph{Stage 1 (GMM)} \textcolor{red}{more descriptive name}
 Gaussian mixture modeling (GMM) is used to assign voxelwise
@@
 
+\textcolor{red}{still feel unclear about it ...  does it go like this?
+\begin{enumerate}
+\item use tissue prior intensity models (not spatial) to initialize
+  Atropos multivariate EM-MRF segmentation of new image  w/o spatial
+  priors 
+\item get out segmentation and posteriors ... run some shape stats on
+  the segmentation 
+\item separately compute the asymmetry features by mapping to the MV
+  template 
+\item input  asymmFeatures + atroposPosteriors + shapeFeatures to the
+  RF model  ... with or w/o ground truth outcomes ?  same either way?
+\end{enumerate}
+
 \paragraph{Stage 2 (MAP-MRF)}

### 2014-04-01 paper.tex
WIP: mild comments

@@
   RF model  ... with or w/o ground truth outcomes ?  same either way?
+  what do you get out of the RF?
 \end{enumerate}
@@
 probabilistic estimates for each of the seven tissue types.  
+\textcolor{red}{can you say more here --- what are these in RF and R
+  terms?  how the RF generate probabilities ?}
 At the beginning of the second stage, these spatial probability maps are 
@@
 forest models improves the resulting labeled output.
+\textcolor{red}{Is this so much a discovery or is it that the MRF
+  features ``fix'' the RF output by introducing spatial coherence to
+  the results .... at least , that's what it seems to me that you are
+  doing ...}
 

### 2014-04-01 paper.tex
WIP: mild comments

@@
 \label{sec:misc}
-
+Our strategy for feature selection was to, first, generate an ample feature
+set and to, second, use the training phase to prioritize the features
+appropriately.  In principal, this will eliminate weak features and
+retain strong ones even in the presence of substantial noise (cite
+some RF paper here ...).
 In addition to the feature images outlined in the previous sections, 

### 2014-04-01 paper.tex
WIP: mild comments

--- 2014-04-01 paper.tex
@@
 \label{sec:misc}
-
-In addition to the feature images outlined in the previous sections, 
-several other feature images were generated largely based on what 
-had been used in previous work.  For each modality, several images
+Our strategy for feature selection was to, first, generate an ample feature
+set and, second, to use the training phase to prioritize the features
+appropriately.  In principal, this will eliminate weak features and
+retain strong ones even in the presence of substantial noise (cite
+some RF paper here ...).  For each modality, several images
 were generated based on calculation of voxelwise neighborhood 

### 2014-04-02 paper.tex
WIP: mild comments
WIP: minor comments

@@
 \begin{keyword}
-advanced normalization tools \sep BRATS \sep brain tumor segmentation \sep R project for statistical computing
+advanced normalization tools \sep BRATS \sep brain tumor segmentation \sep \textit{R} project for statistical computing
 %% keywords here, in the form: keyword \sep keyword
@@
 \footnote{
-One of the principal advantages of R is the extensive community of
+One of the principal advantages of \textit{R} is the extensive community of
 developers  who have contributed on the order of thousands of packages 
-extending R 's capabilities beyond its core functionality.
+extending \textit{R} 's capabilities beyond its core functionality.
 Most relevant here
@@
 the asymmetry features described in the previous section.
+
 \textcolor{red}{Nick, I think the motivation for this is still unclear
@@
 \end{align}
-given in the standard R notation of \cite{wilkinson1973}.  The
+given in the standard \textit{R} notation of \cite{wilkinson1973}.  The
 set of $\left\{F_i: i \in Asym\right\}$ are the feature images 
@@
 \end{enumerate}
+}
 
@@
 \subsection{\textit{ANTsR}:  An ANTs/R Interface}
+We briefly describe our imaging-specific interface between
+\textit{ANTs} and \textit{R}.
 
@@
 powerful neuroimaging toolkit for producing reliable imaging data 
-transformations, and the R project%
+transformations, and the \textit{R} project%
 \footnote{
@@
 Careful consideration of available statistical software 
-led to the adoption of R to complement ANTs quantification resulting in the
+led to the adoption of \textit{R} to complement ANTs quantification resulting in the
 \textit{ANTsR} package.
-%The R project for statistical computing,%
+%The \textit{R} project for statistical computing,%
 %\footnote{
@@
 %and data visualization.  
-R's open source code base, reliable software testing and distribution strategies,
+\textit{R}'s open source code base, reliable software testing and distribution strategies,
 and add-on packages coupled with its rapidly growing 
@@
 }
-Prior to installation of \textit{ANTsR}, several external R packages
+Prior to installation of \textit{ANTsR}, several external \textit{R} packages
 need to be installed including: \verb#Rcpp#, \verb#signal#, \verb#timeSeries#, 
@@
 \textit{ANTsR} is intended to not only allow easy interchange between
-medical imaging formats and R but also to facilitate
+medical imaging formats and \textit{R} but also to facilitate
 reproducible scientific studies and the type compilable analysis
@@
 functionality, including IO tools such as \verb#antsImageRead#, 
-with the large number of R statistical and
+with the large number of \textit{R} statistical and
 visualization packages.  Due to this combination, several
@@
 packages available in R.  The combination of the well-known ANTs software
-package with R provides tremendous potential for future insightful analysis.
+package with \textit{R} provides tremendous potential for future insightful analysis.
 

### 2014-04-25 paper.tex
ENH:  Addressing Brian's comments.

@@
 are then refined using a series of heuristically-derived binary morphological
-operations.  \textcolor{red}{The two-stage approach is motivated by
+operations.  The two-stage approach is motivated by
 the fact that the RF is a powerful tool for ranking the features
@@
 However, the RF outputs are not geometrically constrained, which is
-the motivation for our second refinement step.}.
+the motivation for our second refinement step.
 
 In line with our philosophy regarding open science \citep{tustison2013,ince2012}, 
-we made the code available as open source%
+we have made the code available as open source%
 \footnote{
@@
 describe the code and data%
-\footnote{
-Although all data was provided having been skull-stripped, 
-we note to the reader that an
-available ANTs-based solution to the brain extraction problem
-is found in the script {\tt antsBrainExtraction.sh} which uses
-a hybrid approach based on image registration and binary morphology
-to isolate the brain \citep{avants2010a}.  An evaluation with other
-methodologies demonstrated comparable to superior performance 
-\citep{avants2010a}.
-} 
+%\footnote{
+%Although all data was provided having been skull-stripped, 
+%we note to the reader that an
+%available ANTs-based solution to the brain extraction problem
+%is found in the script {\tt antsBrainExtraction.sh} which uses
+%a hybrid approach based on image registration and binary morphology
+%to isolate the brain \citep{avants2010a}.  An evaluation with other
+%methodologies demonstrated comparable to superior performance 
+%\citep{avants2010a}.
+%} 
 used for the BRATS challenge so that 
@@
 
-Certain key elements characterize the supervised brain tumor segmentation
+Certain key elements characterize the proposed supervised brain tumor segmentation
 protocol including:
@@
 \setlength{\tabcolsep}{2pt}
-
-\begin{figure}[h]
-  \centering
-  \begin{tabular}{cccc}
-  \includegraphics[width=30mm]{Figures/BRATS_HG0001/BRATS_HG0001_FLAIR_slice93.png} &
-  \includegraphics[width=30mm]{Figures/BRATS_HG0001/BRATS_HG0001_T1_slice93.png} &
-  \includegraphics[width=30mm]{Figures/BRATS_HG0001/BRATS_HG0001_T1C_slice93.png} &
-  \includegraphics[width=30mm]{Figures/BRATS_HG0001/BRATS_HG0001_T2_slice93.png} \\
-  (a) & (b) &
-  (c) & (d) \\
-  \end{tabular}
-  \caption{Induced bilateral asymmetry due to tumor presence causing
-  distortion of the plane of symmetry.  Shown are mid-axial slices of
-  one of the BRATS 2013 training data 
-  (specifically {\tt BRATS\_HG0001}):  (a) FLAIR, (b) T1, (c) T1C, and
-  (d) T2.  }
-  \label{fig:asymmetry}
-\end{figure}
 
@@
 for feature generation in multiple sclerosis lesion \citep{geremia2011} and 
-tumor \citep{geremia2012} identification.  However, this
+tumor \citep{geremia2012} identification.  However, this earlier
 approach does not take into account the displacement of 
@@
 
-\begin{figure}[h]
-    \centerline{\includegraphics[width=130mm]{Figures/templateKirby.pdf}}
+\begin{figure}
+  \centering
+  \begin{tabular}{cccc}
+  \includegraphics[width=30mm]{BRATS_HG0001/BRATS_HG0001_FLAIR_slice93.png} &
+  \includegraphics[width=30mm]{BRATS_HG0001/BRATS_HG0001_T1_slice93.png} &
+  \includegraphics[width=30mm]{BRATS_HG0001/BRATS_HG0001_T1C_slice93.png} &
+  \includegraphics[width=30mm]{BRATS_HG0001/BRATS_HG0001_T2_slice93.png} \\
+%  \includegraphics[width=20mm]{BRATS_HG0001/BRATS_HG0001_FLAIR_slice93.png} &
+%  \includegraphics[width=20mm]{BRATS_HG0001/BRATS_HG0001_T1_slice93.png} &
+%  \includegraphics[width=20mm]{BRATS_HG0001/BRATS_HG0001_T1C_slice93.png} &
+%  \includegraphics[width=20mm]{BRATS_HG0001/BRATS_HG0001_T2_slice93.png} \\
+  (a) & (b) &
+  (c) & (d) \\
+  \end{tabular}
+  \caption{Induced bilateral asymmetry due to tumor presence causing
+  distortion of the plane of symmetry.  Shown are mid-axial slices of
+  one of the BRATS 2013 training data 
+  (specifically {\tt BRATS\_HG0001}):  (a) FLAIR, (b) T1, (c) T1C, and
+  (d) T2.  }
+  \label{fig:asymmetry}
+\end{figure}
+
+\begin{figure*}
+    \centerline{\includegraphics[width=130mm]{templateKirby.pdf}}
   \caption{Multivariate symmetric template created from the MMRR 
@@
            (a) FA, (b) FLAIR, (c) MPRAGE, and (d) T2 template components.
+           Although DWI-based images were not included in the challenge data, 
+           such images have shown discriminative potential \citep{price2003,cha2005} 
+           warranting investigation in our future work.  
            The optimal transformation and averaging of the individual 
@@
   \label{fig:symmetrictemplates}
-\end{figure}
+\end{figure*}
 
@@
 
-\begin{figure}[!h]
+\begin{figure}
   \centering
-    \includegraphics[width=100mm]{Figures/asymmetryFeatureImages.pdf}
+%    \includegraphics[width=80mm]{asymmetryFeatureImages.pdf}
+    \includegraphics[width=120mm]{asymmetryFeatureImages.pdf}
   \caption{ Given the mapping between the template, $T$, and subject, $S$, domains ($\phi_b: S  \leftrightarrow \underset{b}{\leftrightsquigarrow} T$), various features can be calculated which demonstrate good discriminative qualities.
@@
 optimal multivariate template appearance 
-$\mathbf{J}=\{J_1,J_2,\ldots, J_K\}$ to minimize the
+$\mathbf{J}=\{J_1,J_2,\ldots, J_K\}$ (with corresponding 
+coordinate system $\psi(\mathbf{x})$) to minimize the
 following cost function \textcolor{red}{not sure the 2nd term is correct...}:
 \begin{align}
-  \sum_{n=1}^N 
-        \left[ D \left( \psi(\mathbf{x}),\phi_1^n(\mathbf{x},1)\right)  
-        + \sum_{k=1}^K \lambda_k \Pi_k \left(I_k^n\left(\phi_n(\mathbf{x},0.5)\right),J_k\left(\phi^{-1}_n(\mathbf{x},0.5)\right)\right)\right]
+  \label{test}
+  \sum_{n=1}^N \Bigg[ &D \left( \psi(\mathbf{x}),\phi_1^n(\mathbf{x},1)\right) \nonumber  \\
+           &+ \sum_{k=1}^K \lambda_k \Pi_k \left(I_k^n\left(\phi_n(\mathbf{x},0.5)\right),J_k   
+           \left(\phi^{-1}_n(\mathbf{x},0.5)\right)\right)\Bigg]
 \end{align}
+%\begin{align}
+%  \sum_{n=1}^N 
+%        \left[ D \left( \psi(\mathbf{x}),\phi_1^n(\mathbf{x},1)\right)   
+%        + \sum_{k=1}^K \lambda_k \Pi_k \left(I_k^n\left(\phi_n(\mathbf{x},0.5)\right),J_k\left(\phi^{-1}_n(\mathbf{x},0.5)\right)\right)\right]
+%\end{align}
 $D$ is the diffeomorphic shape distance,
@@
 
-In terms of implementation, this template building algorithm is 
-encapsulated in the script \verb#antsMultivariateTemplateConstruction.sh#,
-available in the ANTs repository, which permits serial or parallel processing on
-an individual workstation or on a computational cluster.  In
-Figure \ref{fig:symmetrictemplates} we show mid-axial slices from
-the MMRR multivariate symmetrical template consisting of FA%
-\footnote{
-Although DWI-based images were not included in the
-challenge data, such images have shown discriminative potential 
-\citep{price2003,cha2005} warranting investigation in our future
-work.  
-}, FLAIR,  
-T1, and T2 components.  These images have been made publicly available
-along with other processed images such as corresponding
-tissue priors and structural labels.%
-\footnote{
-http://figshare.com/articles/Kirby\_multivariate\_template/852989.
-All 7 components are stored under the more informal ``Kirby'' data set moniker.
-}
+%In terms of implementation, this template building algorithm is 
+%encapsulated in the script \verb#antsMultivariateTemplateConstruction.sh#,
+%available in the ANTs repository, which permits serial or parallel processing on
+%an individual workstation or on a computational cluster.  In
+%Figure \ref{fig:symmetrictemplates} we show mid-axial slices from
+%the MMRR multivariate symmetrical template consisting of FA%
+%\footnote{
+%Although DWI-based images were not included in the
+%challenge data, such images have shown discriminative potential 
+%\citep{price2003,cha2005} warranting investigation in our future
+%work.  
+%}, FLAIR,  
+%T1, and T2 components.  These images have been made publicly available
+%along with other processed images such as corresponding
+%tissue priors and structural labels.%
+%\footnote{
+%http://figshare.com/articles/Kirby\_multivariate\_template/852989.
+%All 7 components are stored under the more informal ``Kirby'' data set moniker.
+%}
 
@@
 
+\begin{figure*}
+\centering
+  \makebox[0.9\textwidth][c]{
+  \includegraphics[width=150mm]{featureImages.pdf}
+  }
+  \caption{Representative feature images derived from the 
+           BRATS\_HG0301  ``challenge'' data set.
+           Neighborhood statistical images for each modality were 
+           generated by calculating a given statistic within
+           a specified neighborhood radius.  Also calculated for each modality were feature
+           images based on either the GMM or the MAP-MRF segmentation.  For the former, we
+           show the probability maps for each of the seven labels which are used as feature
+           images.  From the resulting hard segmentation, we calculate various geometric 
+           measures per connected component of each of the seven labels.  Similarly, the 
+           registration to the symmetric template produces the modality-specific 
+           difference images with the
+           corresponding symmetric template itself and with respect to the 
+           contralateral side.   
+           This mapping is also used to produce the log Jacobian image.  Finally, the 
+           (T1 - TC) image is calculated and, from the
+           cerebral mask, we calculate the normalized distance image.  
+           }
+  \label{fig:featureImages}         
+\end{figure*}
+
 \subsubsection{Image Features for Concatenated Random Forests}
 
-\begin{figure}[!h]
+\begin{figure}
   \centering
   \begin{tabular}{c}
-    \includegraphics[width=100mm]{Figures/pipeline2.pdf}
+%    \includegraphics[width=88mm]{pipeline2.pdf}
+    \includegraphics[width=110mm]{pipeline2.pdf}
   \end{tabular}
@@
 
-\textcolor{red}{Nick, I think the motivation for this is still unclear
-  .... doesnt it have something to do with the MRF of atropos
-  smoothing out the importance features resulting from the RF step?
-  Figure 4 tells us a little bit about the procedure of the code but
-  the idea is not conveyed ... would it be possible to show exactly
-  what is happening in a separate figure?  e.g. just conceptually in a small 
-  region of voxels with 2 modalities ....?}
+\begin{figure}
+  \centering
+  \begin{tabular}{c}
+    \includegraphics[width=88mm]{gmm_map_mrf.pdf}
+  \end{tabular}
+  \caption{Visual comparison between the first and second RF stages
+  on the evaluation {\tt BRATS\_HG0001} data set.
+  Although the GMM intensity modeling does provide certain 
+  discriminative benefit (see Figures \ref{fig:hgimportance}
+  and \ref{fig:lgimportance}), the use of the RF-derived posteriors 
+  coupled with the described MAP-MRF segmentation framework enhances 
+  the accuracy of the intensity-based modeling features.    
+  }
+  \label{fig:stageComparison}
+\end{figure}
+
+%\textcolor{red}{Nick, I think the motivation for this is still unclear
+%  .... doesnt it have something to do with the MRF of atropos
+%  smoothing out the importance features resulting from the RF step?
+%  Figure 4 tells us a little bit about the procedure of the code but
+%  the idea is not conveyed ... would it be possible to show exactly
+%  what is happening in a separate figure?  e.g. just conceptually in a small 
+%  region of voxels with 2 modalities ....?}
 
@@
 is used to determine the connected components for each label.  
-Geometric features (assigned voxel-wise) include the physical volumes 
+Geometric features (assigned voxelwise) include the physical volumes 
 of each connected component, the volume to surface area ratio, 
@@
 
-\paragraph{Stage 1 (GMM)} \textcolor{red}{more descriptive name}
+\paragraph{Stage 1 (Initial voxelwise classification)} 
 Gaussian mixture modeling (GMM) is used to assign voxelwise
@@
 
-\textcolor{red}{still feel unclear about it ...  does it go like this?
-\begin{enumerate}
-\item use tissue prior intensity models (not spatial) to initialize
-  Atropos multivariate EM-MRF segmentation of new image  w/o spatial
-  priors 
-\item get out segmentation and posteriors ... run some shape stats on
-  the segmentation 
-\item separately compute the asymmetry features by mapping to the MV
-  template 
-\item input  asymmFeatures + atroposPosteriors + shapeFeatures to the
-  RF model  ... with or w/o ground truth outcomes ?  same either way?
-  what do you get out of the RF?
-\end{enumerate}
-}
-
-\paragraph{Stage 2 (MAP-MRF)}
-
-Application of the Stage 1 RF model results in a set of
-probabilistic estimates for each of the seven tissue types.  
-\textcolor{red}{can you say more here --- what are these in RF and R
-  terms?  how the RF generate probabilities ?}
+\paragraph{Stage 2 (Refined spatial coherence classification)}
+
+Application of the Stage 1 RF model to a set of feature images 
+from an individual subjects results in a set of
+images estimating the spatial probabilities for each 
+of the seven tissue types. At each voxel, the corresponding multi-feature 
+vector is propagated through each tree of the random forest 
+resulting in a single vote/classification per tree.  The
+classification probabilities are derived by normalizing 
+these votes to $[0,1]$.
+
 At the beginning of the second stage, these spatial probability maps are 
@@
 \end{align}
-We discovered that this composite application of random
-forest models improves the resulting labeled output.
-\textcolor{red}{Is this so much a discovery or is it that the MRF
-  features ``fix'' the RF output by introducing spatial coherence to
-  the results .... at least , that's what it seems to me that you are
-  doing ...}
+The use of these MRF features improves the accuracy of the 
+intensity-based modeling over the GMM approach by introducing
+a spatial coherence constraint to those features in addition to 
+providing a better initialization with the RF posteriors from 
+the initial estimation stage.
+
+%\textcolor{red}{Is this so much a discovery or is it that the MRF
+%  features ``fix'' the RF output by introducing spatial coherence to
+%  the results .... at least , that's what it seems to me that you are
+%  doing ...}
 
@@
 
-\begin{figure*}[h]
-%  \centering
-%  \makebox[0.9\textwidth][c]{
-  \includegraphics[width=150mm]{Figures/featureImages.pdf}
-%  }
-  \caption{Representative feature images derived from the 
-           BRATS\_HG0301  ``challenge'' data set.
-           Neighborhood statistical images for each modality were 
-           generated by calculating a given statistic within
-           a specified neighborhood radius.  Also calculated for each modality were feature
-           images based on either the GMM or the MAP-MRF segmentation.  For the former, we
-           show the probability maps for each of the seven labels which are used as feature
-           images.  From the resulting hard segmentation, we calculate various geometric 
-           measures per connected component of each of the seven labels.  Similarly, the 
-           registration to the symmetric template produces the modality-specific 
-           difference images with the
-           corresponding symmetric template itself and with respect to the 
-           contralateral side.   
-           This mapping is also used to produce the log Jacobian image.  Finally, the 
-           (T1 - TC) image is calculated and, from the
-           cerebral mask, we calculate the normalized distance image.  
-           }
-  \label{fig:featureImages}         
-\end{figure*}
+\subsection{Overview}
+
+In summary, the proposed RF classification approach per subject consists of the 
+following steps:
+
+\begin{enumerate}
+  \item Calculate stage 1 features based on:  
+  \begin{itemize}
+    \item GMM intensity modeling,
+    \item asymmetry, and
+    \item miscellaneous.
+  \end{itemize}
+  \item Apply stage 1 RF model to the above calculated features resulting in seven 
+        probability maps for the seven tissue types.  
+  \item Use the probability maps to calculate stage 2 features based on MAP-MRF intensity modeling.
+  \item Apply stage 2 RF model to MAP-MRF, asymmetry, and miscellaneous features to 
+  get final RF classification. 
+\end{enumerate}
+
+%\textcolor{red}{still feel unclear about it ...  does it go like this?
+%\begin{enumerate}
+%\item use tissue prior intensity models (not spatial) to initialize
+%  Atropos multivariate EM-MRF segmentation of new image  w/o spatial
+%  priors 
+%\item get out segmentation and posteriors ... run some shape stats on
+%  the segmentation 
+%\item separately compute the asymmetry features by mapping to the MV
+%  template 
+%\item input  asymmFeatures + atroposPosteriors + shapeFeatures to the
+%  RF model  ... with or w/o ground truth outcomes ?  same either way?
+%  what do you get out of the RF?
+%\end{enumerate}
+%}
 
 \subsection{\textit{ANTsR}:  An ANTs/R Interface}
-We briefly describe our imaging-specific interface between
-\textit{ANTs} and \textit{R}.
-
-\subsubsection{Overview}
 
@@
 
-\begin{figure}[!htb]
-  \centering
-  \includegraphics[width=120mm]{Figures/challengeResults.pdf}
+\begin{figure*}
+  \vspace{-20mm}
+  \makebox[\textwidth][c]{
+  \includegraphics[width=140mm]{challengeResults.pdf}
+  }
   \caption{Visual results (every tenth slice sampled from the axial center of the
@@
   \label{fig:challengeresults}
-\end{figure}
+\end{figure*}
 
@@
   \begin{tabular}{cc}
-  \includegraphics[width=90mm]{Figures/BRATS_HG_GMM.pdf} &
-  \includegraphics[width=90mm]{Figures/BRATS_HG_MAP_MRF.pdf} \\
+  \includegraphics[width=90mm]{BRATS_HG_GMM.pdf} &
+  \includegraphics[width=90mm]{BRATS_HG_MAP_MRF.pdf} \\
   \end{tabular}
@@
   ranking describing the importance of each feature for predictive accuracy. 
-  Features are listed from most to least important.
   }
@@
   \begin{tabular}{cc}
-  \includegraphics[width=90mm]{Figures/BRATS_LG_GMM.pdf} &
-  \includegraphics[width=90mm]{Figures/BRATS_LG_MAP_MRF.pdf} \\
+  \includegraphics[width=90mm]{BRATS_LG_GMM.pdf} &
+  \includegraphics[width=90mm]{BRATS_LG_MAP_MRF.pdf} \\
   \end{tabular}
@@
   ranking describing the importance of each feature for predictive accuracy.
-  Features are listed from most to least important.
   }
@@
 %\section*{Acknowledgments}
-\clearpage
+%\clearpage
 \section*{References}

### 2014-04-25 paper.tex
BUG: eq 1

@@
 coordinate system $\psi(\mathbf{x})$) to minimize the
-following cost function \textcolor{red}{not sure the 2nd term is correct...}:
+following cost function:
 \begin{align}
@@
   \sum_{n=1}^N \Bigg[ &D \left( \psi(\mathbf{x}),\phi_1^n(\mathbf{x},1)\right) \nonumber  \\
-           &+ \sum_{k=1}^K \lambda_k \Pi_k \left(I_k^n\left(\phi_n(\mathbf{x},0.5)\right),J_k   
-           \left(\phi^{-1}_n(\mathbf{x},0.5)\right)\right)\Bigg]
+           &+ \sum_{k=1}^K \lambda_k \Pi_k \left(I_k^n,J_k   
+           \left(\phi^{-1}_n(\mathbf{x},1)\right)\right)\Bigg]
 \end{align}

### 2014-04-28 paper.tex
WIP: minor comments

@@
 feature images and the concatenated RF construction.  We also
-describe the code and data%
+describe the code and data used for the BRATS challenge so that 
+the interested reader can reproduce our results.  Finally, we 
+illustrate performance with results from the BRATS 2013 challenge.
+
 %\footnote{
@@
 %}
-used for the BRATS challenge so that 
-the interested reader can reproduce our results.  Finally, we 
-illustrate performance with results from the BRATS 2013 challenge.
 
@@
 first RF model (or first stage) which produces a voxelwise 
-probabilistic tissue estimate%
-\footnote{
-More precisely, in the RF framework, each prediction sample (i.e. the feature vector at each voxel),
+probabilistic tissue estimate.  More precisely, in the RF framework, each prediction sample (i.e. the feature vector at each voxel),
 is propagated through each tree of the ensemble where it is labeled as belonging to a specific
-class.  These ``votes'' are converted to voxelwise probabilistic estimates for each class.
-}
- for the seven brain/tumor tissue types:
+class.  These ``votes'' are converted to voxelwise probabilistic
+estimates for each class.
+\textcolor{red}{FIXME --- above comments need more detail ... how/why
+  is this a probability, what does this mean?   it's bootstrapping etc
+right so what can we say about this?}
+The seven brain/tumor tissue types include:
 \begin{itemize}

### 2014-04-29 paper.tex
WIP: more minor edits

@@
 These factors limit the applicability of techniques with established
-performance in related tasks such as normal brain tissue segmentation.
-We introduce a template-based framework for computing geometric and
-asymmetry-driven feature sets from multiple modality neuroimaging.  We
-use these features to drive a supervised whole-brain and tumor segmentation
-approach based on random forests.  The asymmetry-related features 
+performance in related tasks such as normal brain tissue
+segmentation.  To alleviate the limitations of traditional tissue
+classification, we introduce a symmetric template-based framework for supervised
+segmentation based on multiple modality intensity, geometry and asymmetry feature sets.  These features drive a supervised whole-brain and tumor segmentation
+approach based on random forest-derived probabilities.  The
+asymmetry-related features, in particular, 
 demonstrate excellent discriminative properties within this framework. 
-This approach is made possible by symmetric multimodality brain templates which are optimal
-in terms of both shape and multivariate intensity.  Further
-performance improvements are demonstrated by a successive application
-of random forest models for increasing levels of feature refinement to
-which we refer as {\it concatenated random forests}.  Although the
-derived workflow involves many disparate components, the newly
-developed \textit{ANTsR} open source software package significantly facilitates
+We also gain performance by alternately generating
+probability maps from random forest models and feeding these maps 
+into Markov random field regularized probabilistic segmentation.  This
+strategy allows us to interface the supervised learning capabilities of the random forest
+model with regularized probabilistic segmentation.  The newly
+developed \textit{ANTsR} open source software package facilitates
 the proposed methodology.  \textit{ANTsR} is a comprehensive interface between the
@@
 MS lesion segmentation \cite{geremia2011}, 
-multimodal brain tumor segmentationp
+multimodal brain tumor segmentation
 \citep{bauer2012,zikic2012}, brain extraction \citep{iglesias2010}, 

### 2014-04-29 paper.tex
WIP: more careful reading / mods

--- 2014-04-28 paper.tex
@@
 
-\usepackage{color,hyperref}
+\usepackage{color,hyperref,bm}
 \usepackage{multirow,booktabs,ctable,array}
@@
   \item generation of other image-based features,
-  \item training of RF models, and
+  \item training and geometric refinement of RF models, and
   \item prediction using the proposed framework.
@@
 \subsection{Image Features for RF Model Training and Prediction}
+Feature generation is a critical step in machine learning regardless
+of which prediction methods are employed.  Below, we detail the
+feature extraction approach and how these results are fed into the
+random forest technology on which we based our final segmentations.
 
@@
   binary morphological resulting in the final labeled brain.
+  \textcolor{red}{I still find this unclear, especially the arrows
+    that cross between training and testing --- intuitively, this
+    seems impossible w/o circularity.}
   }
@@
 
+We use random forests as supervised learners.  Given ground truth,
+voxel-wise $n$-class labels for a specific image, we can vectorize the
+label image to produce $v_L$, the $p$-length vector of class labels
+where $p$ is the number of voxels in the given brain of interest.
+We also assume a matrix of features for the given brain, which we
+denote $\bm{F}$ where $\bm{F}$ has dimensionality of $p$ by
+$k$, where each of the $k$ columns corresponds a potentially valuable
+feature vector.  Given this representation, the random forest model
+can be trained, in \textit{R} notation, to predict the labels via
+$v_L \sim  \bm{F}$.
 Our segmentation protocol involves training and application
@@
 is propagated through each tree of the ensemble where it is labeled as belonging to a specific
-class.  These ``votes'' are converted to voxelwise probabilistic
-estimates for each class.
-\textcolor{red}{FIXME --- above comments need more detail ... how/why
-  is this a probability, what does this mean?   it's bootstrapping etc
-right so what can we say about this?}
-The seven brain/tumor tissue types include:
+class.  
+These ``votes'' are converted to voxelwise probabilistic
+estimates for each class via standard mechanisms used in random forest
+models, e.g. \cite{liaw2002}.
+The seven labels of brain/tumor tissue type denote:
 \begin{itemize}
@@
 We then use these output tissue map estimates as spatial 
-priors for generating a second set of image features.  These
+priors for generating a second set of geometrically refined image features.  These
 are used as input for a second RF model application (or
@@
 the output of which constitute the final tumor segmentation estimate.
-Note that some feature images are included in both stages such as
+Note that some feature images are included \textbf{directly and
+  without modification(?)} in both stages such as
 the asymmetry features described in the previous section.
@@
 univariate probability map into the feature vector of the training
-data.  As pointed out in \cite{menze2010}, multivariate modeling
+data.  As pointed out in \cite{menze2010}, parametric multivariate modeling
 might obscure the distinct biological information provided by each 
@@
 of each connected component, the volume to surface area ratio, 
-the elongation, and eccentricity.
-
-\paragraph{Stage 1 (Initial voxelwise classification)} 
-Gaussian mixture modeling (GMM) is used to assign voxelwise
-tissue probabilities which comprise an additional set of image features 
-where 
-prior cluster centers for specific tissue types for each
-modality are learned from training data \citep{reynolds2009}.  
-Intensity normalization
-across the training cohort was originally performed using 
-\cite{nyul2000}.  However, despite good performance on normal
-data, application of this intensity normalization algorithm 
-(after a thorough parameter search) on BRATS data led to slightly 
-worse performance than simply winsorizing the intensity values to the quantile
+the elongation, and eccentricity.  Note that random forests are
+known to be robust to the presence of multiple correlated features.
+
+\paragraph{Stage 1 voxelwise classification with Gaussian mixture
+  modeling} 
+The first step in processing a new image is to segment the image into
+the $n$ tissue classes listed above.   We achieve this via an expectation-maximization
+segmentation method based on
+Gaussian mixture modeling (GMM).  GMM results in voxelwise
+tissue probabilities which comprise an additional set of image
+features.  The GMM is initialized with prior cluster centers for
+specific tissue types.  We learn the values for these cluster centers
+for each modality and each tissue from training data
+\citep{reynolds2009}.  We also perform intensity normalization
+across the training cohort by winsorizing the intensity values to the quantile
 range $[0.01, 0.99]$ and then rescaling the resulting intensities to 
-$[0,1]$ so we opted for the latter, more simplistic intensity
-normalization protocol.  The cluster centers are defined as the mean
+$[0,1]$.  The cluster centers are defined as the mean
 normalized intensity value for each tissue type of each modality 
@@
 initializing with these values is optional and that results
-seem to be robust to initialization values.
-
-A GMM formulates the 
+seem to be robust to initialization values.  \textcolor{red}{How else
+  would you initialize? how robust?  how did you test this?}
+
+More formally, the GMM computes the 
 probability distribution at each voxel, $\mathbf{x}$, as the
@@
 where $\sum_{m=1}^M \lambda_m = 1$.  The parameters of the GMM 
-are determined with maximum likelihood estimation using the 
+are determined using the 
 Atropos segmentation tool \citep{avants2011} available in ANTs.
-We include the resulting seven posterior probability tissue images 
+\textcolor{red}{What modalities are input to Atropos?  Should we
+  reference a figure here to make this clear?  Would listing a table of the
+  prior per-class per-modality tissue cluster centers be helpful?}
+
+\paragraph{Stage 1 random forest} 
+The stage 1 random forest training uses the asymmetry features, the seven posterior probability
+tissue images output by the GMM
 along with five additional geometric features based on the connected 
@@
 \label{eq:gmm}
-\mathrm{Tissue}\,\,\mathrm{classes} \sim \sum_{i \in GMM} F_i + \sum_{i \in Asym} F_i + \sum_{i \in Misc} F_i
+ v_L \sim \sum_{i \in Asym} F_i + \sum_{i \in GMM} F_i + \sum_{i \in Misc} F_i
 \end{align}
-given in the standard \textit{R} notation of \cite{wilkinson1973}.  The
-set of $\left\{F_i: i \in Asym\right\}$ are the feature images 
+given in the standard \textit{R} notation of \cite{wilkinson1973}.  
+In other words, $\bm{F}= [ \bm{F}_{Asym}, \bm{F}_{GMM},
+\bm{F}_{Misc} ]$, where we column-wise concatenate the GMM, Asym and
+Misc feature matrices.
+The set of $\left\{F_i: i \in Asym\right\}$ are the feature images 
 generated from the symmetric multivariate template 
@@
 are miscellaneous feature images described in Section \ref{sec:misc}.
+Therefore, in total, the number of features is $k=FIXME$ where there
+are 7 GMM features, FIXME asymmetry features and FIXME miscellaneous
+geometric features.
 
 \paragraph{Stage 2 (Refined spatial coherence classification)}
-
-Application of the Stage 1 RF model to a set of feature images 
-from an individual subjects results in a set of
-images estimating the spatial probabilities for each 
+The first step in stage 2 is to use the per-class probabilities output
+by the stage 1 random forest model as input to the Atropos
+segmentation algorithm.  Application of the Stage 1 RF model to a set of feature images 
+from an individual subject results in a set of $n$ spatial images
+containing the voxel-wise probabilities for each 
 of the seven tissue types.  At each voxel, the corresponding multi-feature 
@@
 classification probabilities are derived by normalizing 
-these votes to $[0,1]$.
-
-At the beginning of the second stage, these spatial probability maps are 
-used as spatial priors in a 
-MAP-MRF segmentation also performed using the Atropos 
-segmentation tool \citep{avants2011} available in ANTs although we couple
-it with N4 bias correction \citep{tustison2010} as contained in the script
-{\tt antsAtroposN4.sh}.  The probabilistic segmentation output and corresponding
-connected-component feature images for each modality replace the analogous 
-GMM-based feature images used 
-during the first stage of training/prediction along with the other feature
-images generated during the feature stage (such as the template-based data).
+these votes to $[0,1]$ as in \cite{liaw2002}.
+
+These spatial probability maps are passed as spatial priors to the
+{\tt antsAtroposN4.sh} script in ANTs which couples the Atropos 
+segmentation tool \citep{avants2011} with N4 bias correction \citep{tustison2010}.
+The posterior probabilities that are output by Atropos and the
+connected-component feature images for each modality (described in
+section~\ref{sec:misc}) are then passed to the second stage random
+forest model, along with the same asymmetry feature set based on our
+symmetric templates.  
 The second stage RF model is then produced similarly as the previous stage: 
@@
 \label{eq:mapmrf}
-\mathrm{Tissue}\,\,\mathrm{classes} \sim \sum_{i \in MAP-MRF} F_i + \sum_{i \in Asym} F_i + \sum_{i \in Misc} F_i.
+\mathrm{Tissue}\,\,\mathrm{classes} \sim \sum_{i \in GMM_2} F_i + \sum_{i \in Asym} F_i + \sum_{i \in Misc} F_i,
 \end{align}
+where $GMM_2$ represents the stage 2 application of Atropos and the
+other features are similar to those used in stage 1.
 The use of these MRF features improves the accuracy of the 
@@
 providing a better initialization with the RF posteriors from 
-the initial estimation stage.
-
-%\textcolor{red}{Is this so much a discovery or is it that the MRF
-%  features ``fix'' the RF output by introducing spatial coherence to
-%  the results .... at least , that's what it seems to me that you are
-%  doing ...}
+the initial estimation stage.  \textcolor{red}{This last sentence
+  confuses me --- what is the difference between the GMM and MRF
+  features?  I made a notation change $GMM_2$, b.c i thought the MRF
+  and GMM were the same ... but maybe not?  did you exclude the MRF
+  prior in the first application of Atropos?}
 
@@
 appropriately.  In principal, this will eliminate weak features and
-retain strong ones even in the presence of substantial noise (cite
-some RF paper here ...).  For each modality, several images
-were generated based on calculation of voxelwise neighborhood 
-first-order statistics including mean, variance, skewness, and entropy
-using neighborhood block radii
-of 1 and 3 voxels.  We also calculated two Euclidean distance
+retain strong ones even in the presence of substantial noise.  Given
+that image feature shape may be a valuable predictor of the presence of a
+tumor \cite{zacharaki2009classification}, we calculate, for each modality, several images of neighborhood 
+first-order statistics including mean, variance, skewness, and
+entropy.  Neighborhoods are defined by block radii
+of 1 and 3 voxels \textcolor{red}{Why these values?}.  We also calculated two Euclidean distance
 transforms \citep{maurer2003} to be used for each subject.  One was
@@
 and generates all feature images.  In Figure \ref{fig:featureImages} we provide sample mid-axial 
-image slices from features generated from the {\tt BRATS\_HG0301} data set.
+image slices from features generated from the {\tt BRATS\_HG0301} data
+set.  Note that these feature images do not require the presence of
+any training data.
 

### 2014-04-29 paper.tex
ENH: More edits.

@@
   \centering
-  \begin{tabular}{c}
-%    \includegraphics[width=88mm]{pipeline2.pdf}
-    \includegraphics[width=110mm]{pipeline2.pdf}
-  \end{tabular}
-  \caption{Diagrammatic workflow for the proposed RF model training and prediction.  The feature
-  images are first generated from the input MRI and symmetric multivariate template. If training,
-  the set of feature images derived from the training data are used to create the RF model
-  for the first stage.  The model is then applied to the training or prediction data to generate 
-  a second set of feature images to create the second stage RF model (if training) and/or generate
-  the second RF model.  
-  For prediction, the final refinement process entails a series of heuristically-derived
-  binary morphological resulting in the final labeled brain.
-  \textcolor{red}{I still find this unclear, especially the arrows
-    that cross between training and testing --- intuitively, this
-    seems impossible w/o circularity.}
+   \vspace{-15mm}
+  \makebox[0.9\textwidth][c]{
+    \includegraphics[width=150mm]{pipelineTestingTraining.pdf}
+    }
+  \caption{Diagrammatic workflow for the proposed RF model training (left) and prediction
+  (right).  For training, the feature images are first generated from the set of input 
+  training multimodal MRI and symmetric multivariate template. The set of features images are 
+  use to create the RF model for the first stage.  This first stage RF model is then applied
+  to the set of input training data to yield spatial priors for seeding the generation of the
+  second stage feature images.  These feature images are then used to create the second
+  stage RF model.  For prediction of a single-subject segmentation, a similar scenario 
+  is applied whereby the generated feature images are used as input to the RF models the
+  second of which produces an initial segmentation estimate.  This estimate is then refined
+  using a series of heuristically-derived binary morphological operations resulting in the 
+  final labeled brain.
   }
@@
 the output of which constitute the final tumor segmentation estimate.
-Note that some feature images are included \textbf{directly and
-  without modification(?)} in both stages such as
+Note that some feature images are included directly and
+  without modification in both stages such as
 the asymmetry features described in the previous section.
@@
 
-\paragraph{Stage 1 voxelwise classification with Gaussian mixture
+\paragraph{Stage 1:  Voxelwise classification with Gaussian mixture
   modeling} 
@@
 range $[0.01, 0.99]$ and then rescaling the resulting intensities to 
-$[0,1]$.  The cluster centers are defined as the mean
+$[0, 1]$.  The cluster centers are defined as the mean
 normalized intensity value for each tissue type of each modality 
-image over all the training data.  It should be noted that 
-initializing with these values is optional and that results
-seem to be robust to initialization values.  \textcolor{red}{How else
-  would you initialize? how robust?  how did you test this?}
+image over all the training data.  We tried a more sophisticated intensity
+normalization scheme using the approach of \cite{nyul2000} but found that
+our simpler normalization approach, in combination with the rest of the
+pipeline, produced slightly better results.
 
@@
 where $\sum_{m=1}^M \lambda_m = 1$.  The parameters of the GMM 
-are determined using the 
-Atropos segmentation tool \citep{avants2011} available in ANTs.
-\textcolor{red}{What modalities are input to Atropos?  Should we
-  reference a figure here to make this clear?  Would listing a table of the
-  prior per-class per-modality tissue cluster centers be helpful?}
-
-\paragraph{Stage 1 random forest} 
+are determined using the Atropos segmentation tool \citep{avants2011} available in ANTs.
+For the BRATS data, GMM modeling was applied to each of the four modalities, i.e.,
+FLAIR, T1, T1 contrast, and T2 (see Figure \ref{fig:stageComparison}), separately.  
+Note that no spatial priors are used during this stage including MRF spatial priors.
+
+%\paragraph{Stage 1 random forest} 
 The stage 1 random forest training uses the asymmetry features, the seven posterior probability
@@
 are miscellaneous feature images described in Section \ref{sec:misc}.
-Therefore, in total, the number of features is $k=FIXME$ where there
-are 7 GMM features, FIXME asymmetry features and FIXME miscellaneous
-geometric features.
-
-\paragraph{Stage 2 (Refined spatial coherence classification)}
+Therefore, in total, the number of features is $k=84$ where there
+are twelve GMM features for each of the four modalities ($n=48$), 
+two asymmetry features for each modality ($n=8$), and six modality-specific 
+miscellaneous geometric features ($n=24$) with an additional four feature
+images.
+
+% 48 + 8 + 24 + 4 = 84
+
+\paragraph{Stage 2:  Refined spatial coherence classification}
 The first step in stage 2 is to use the per-class probabilities output
@@
 \label{eq:mapmrf}
-\mathrm{Tissue}\,\,\mathrm{classes} \sim \sum_{i \in GMM_2} F_i + \sum_{i \in Asym} F_i + \sum_{i \in Misc} F_i,
+\mathrm{Tissue}\,\,\mathrm{classes} \sim \sum_{i \in MAP\_MRF} F_i + \sum_{i \in Asym} F_i + \sum_{i \in Misc} F_i,
 \end{align}
-where $GMM_2$ represents the stage 2 application of Atropos and the
-other features are similar to those used in stage 1.
+where $MAP-MRF$ represents the Stage 2 application of Atropos and the
+other features are similar to those used in Stage 1.
 The use of these MRF features improves the accuracy of the 
 intensity-based modeling over the GMM approach by introducing
-a spatial coherence constraint to those features in addition to 
+a spatial coherence constraint (the MRF prior) to those features 
+in addition to 
 providing a better initialization with the RF posteriors from 
-the initial estimation stage.  \textcolor{red}{This last sentence
-  confuses me --- what is the difference between the GMM and MRF
-  features?  I made a notation change $GMM_2$, b.c i thought the MRF
-  and GMM were the same ... but maybe not?  did you exclude the MRF
-  prior in the first application of Atropos?}
+the initial estimation stage (which lacked the MRF spatial prior).
 
@@
 first-order statistics including mean, variance, skewness, and
-entropy.  Neighborhoods are defined by block radii
-of 1 and 3 voxels \textcolor{red}{Why these values?}.  We also calculated two Euclidean distance
+entropy.  Block neighborhood radii of values $\{1,2,3\}$ were tested.
+Due to overlapping discriminative qualities, we narrowed the feature
+values to 1 and 3 voxels.  We also calculated two Euclidean distance
 transforms \citep{maurer2003} to be used for each subject.  One was
@@
 \begin{enumerate}
-  \item Calculate stage 1 features based on:  
+  \item Calculate Stage 1 features based on:  
   \begin{itemize}
@@
   \end{itemize}
-  \item Apply stage 1 RF model to the above calculated features resulting in seven 
+  \item Apply Stage 1 RF model to the above calculated features resulting in seven 
         probability maps for the seven tissue types.  
-  \item Use the probability maps to calculate stage 2 features based on MAP-MRF intensity modeling.
-  \item Apply stage 2 RF model to MAP-MRF, asymmetry, and miscellaneous features to 
+  \item Use the probability maps to calculate Stage 2 features based on MAP-MRF intensity modeling.
+  \item Apply Stage 2 RF model to MAP-MRF, asymmetry, and miscellaneous features to 
   get final RF classification. 

### 2014-04-29 paper.tex
ENH: More edits.

@@
 denote $\bm{F}$ where $\bm{F}$ has dimensionality of $p$ by
-$k$, where each of the $k$ columns corresponds a potentially valuable
+$k$, where each of the $k$ columns corresponds to a potentially valuable
 feature vector.  Given this representation, the random forest model

### 2014-04-30 paper.tex
ENH:  More Brian comments.

@@
 Therefore, in total, the number of features is $k=84$ where there
-are twelve GMM features for each of the four modalities ($n=48$), 
+are twelve GMM features (7 posteriors plus 5 geometric features) for each 
+of the four modalities ($n=48$), 
 two asymmetry features for each modality ($n=8$), and six modality-specific 
@@
 \label{eq:mapmrf}
-\mathrm{Tissue}\,\,\mathrm{classes} \sim \sum_{i \in MAP\_MRF} F_i + \sum_{i \in Asym} F_i + \sum_{i \in Misc} F_i,
+v_L \sim \sum_{i \in MAP\_MRF} F_i + \sum_{i \in Asym} F_i + \sum_{i \in Misc} F_i,
 \end{align}
@@
 the initial estimation stage (which lacked the MRF spatial prior).  
+In total, the number of features is for this final stage is also $k=84$ 
+where the GMM features are replaced by their MAP-MRF counterparts
+(7 posteriors plus 5 geometric features) for each of the four modalities ($n=48$).
+The rest of the feature images from the first stage
+(the two asymmetry features for each modality, six modality-specific 
+miscellaneous geometric features, and four miscellaneous feature
+images) are also included unaltered for this second stage.
 

### 2014-05-01 paper.tex
ENH: 1st sent in abstract

@@
 \begin{abstract} 
-Automated brain tumor segmentation in MRI is complicated by the lack
+Segmenting and quantifying gliomas from MRI is an important task for diagnosis, planning intervention, 
+and for tracking tumor changes over time.
+However, this task is complicated by the lack
 of prior knowledge concerning tumor location, spatial extent, shape,

### 2014-05-03 paper.tex
ENH:  More edits.

--- 2014-04-29 paper.tex
@@
 \author[label1]{Christopher R.~Durst}
-\author[label1]{James C.~Gee}
+\author[label2]{Benjamin M.~Kandel}
+\author[label2]{James C.~Gee}
 \author[label3]{Murray C.~Grossman}
@@
 possible displacement of normal tissue, and intensity signature.
-These factors limit the applicability of techniques with established
-performance in related tasks such as normal brain tissue
-segmentation.  To alleviate the limitations of traditional tissue
-classification, we introduce a symmetric template-based framework for supervised
-segmentation based on multiple modality intensity, geometry and asymmetry feature sets.  These features drive a supervised whole-brain and tumor segmentation
+%These factors limit the applicability of techniques with established
+%performance in related tasks such as normal brain tissue
+%segmentation.  
+To accommodate such complications, we introduce a framework for supervised
+segmentation based on multiple modality intensity, geometry, and asymmetry feature sets.  These features drive a supervised whole-brain and tumor segmentation
 approach based on random forest-derived probabilities.  The
-asymmetry-related features, in particular, 
-demonstrate excellent discriminative properties within this framework. 
-We also gain performance by alternately generating
-probability maps from random forest models and feeding these maps 
-into Markov random field regularized probabilistic segmentation.  This
-strategy allows us to interface the supervised learning capabilities of the random forest
-model with regularized probabilistic segmentation.  The newly
-developed \textit{ANTsR} open source software package facilitates
-the proposed methodology.  \textit{ANTsR} is a comprehensive interface between the
+asymmetry-related features (based on optimal symmetric multimodal 
+templates) demonstrate excellent discriminative 
+properties within this framework. We also gain performance by  
+generating probability maps from random forest models and using these maps 
+for a refining Markov random field regularized probabilistic segmentation.  
+This strategy allows us to interface the supervised learning capabilities 
+of the random forest model with regularized probabilistic segmentation
+using the recently developed \textit{ANTsR} 
+package---a comprehensive statistical and visualization interface between the
 popular Advanced Normalization Tools (ANTs) and the \textit{R}
-statistical project. \textit{ANTsR} integrates proven image analysis methods
-for image registration, segmentation, bias correction, and
-template building with \textit{R}'s statistical and machine
-learning techniques and state-of-the-art visualization possibilities.  
+statistical project. 
+%\textit{ANTsR} integrates proven image analysis methods
+%for image registration, segmentation, bias correction, and
+%template building with \textit{R}'s statistical and machine
+%learning techniques and state-of-the-art visualization possibilities.  
 The reported algorithmic framework was the
 top-performing entry in the MICCAI 2013 Multimodal Brain Tumor
-Segmentation challenge and, to our knowledge, is the only entry for
-both challenge years (2012 \& 2013) which has been made publicly
-available.
+Segmentation challenge.  The challenge data were widely varying 
+consisting of both high-grade and low-grade glioma tumor 
+four-modality MRI from five different institutions.
+Average Dice overlap measures for the final algorithmic assessment
+were 0.87, 0.78, and 0.74 for ``complete'', ``core'', and ``enhanced'' 
+tumor components, respectively.
+%and, to our knowledge, is the only entry for
+%both challenge years (2012 \& 2013) which has been made publicly
+%available.
 \end{abstract}
@@
 \begin{keyword}
-advanced normalization tools \sep BRATS \sep brain tumor segmentation \sep \textit{R} project for statistical computing
+advanced normalization tools \sep BRATS \sep brain tumor segmentation 
+\sep \textit{R} project for statistical computing
 %% keywords here, in the form: keyword \sep keyword
@@
 for precise volumetric measurements for tumor characterization, 
-much research has been invested in computational approaches for 
+much research has been invested in computational methods for 
 automated segmentation of tumor regions in MR images.  Several 
@@
 After observing the 2012 challenge to gain insight for our own research 
-needs \citep{durst2014}, we found that no teams had availed their software in any 
-form.  This prompted our participation in the challenge which took place
-the following year in Nagoya, Japan.  Heavily influenced by the success 
+needs \citep{durst2014}, we found few implementation resources available.
+  This prompted our participation in the challenge which took place
+the following year in Nagoya, Japan concerning brain tumor segmentation
+of a varied data set from five different institutions involving FLAIR, 
+T1-weighted, T1-weighted contrast (T1C), and T2-weighted MRI modalities.  Heavily influenced by the success 
 of participants in the 2012 challenge (specifically, the work of 
@@
 accuracy.%
-\footnote{
-One of the principal advantages of \textit{R} is the extensive community of
-developers  who have contributed on the order of thousands of packages 
-extending \textit{R} 's capabilities beyond its core functionality.
-Most relevant here
-is the {\tt randomForest} package developed from Breiman's original
-Fortran code by Liaw and Wiener \citep{liaw2002}.
-}
-
-Early adoption \cite{viola2005} and success in the
+%\footnote{
+%One of the principal advantages of \textit{R} is the extensive community of
+%developers  who have contributed on the order of thousands of packages 
+%extending \textit{R} 's capabilities beyond its core functionality.
+%Most relevant here
+%is the {\tt randomForest} package developed from Breiman's original
+%Fortran code by Liaw and Wiener \citep{liaw2002}.
+%}
+
+Early adoption \citep{viola2005} and success in the
 computer vision community
@@
 \citep{bauer2012,zikic2012}, brain extraction \citep{iglesias2010}, 
+classification of Alzheimer's disease \citep{gray2013},
 anatomy detection in computed tomography \citep{criminisi2013}, and
@@
 was the use of image features based on multimodal shape and appearance 
-asymmetries which were generated via the construction of symmetric, 
+asymmetries which were generated via the construction of symmetric
 multivariate brain templates.  Additional improvements in performance
@@
 RF model seeds the construction of a subset of new feature images
-for input into the second stage.  The results of the second RF model
-are then refined using a series of heuristically-derived binary morphological
-operations.  The two-stage approach is motivated by
+for input into the second stage.  
+%The results of the second RF model application
+%are then refined using a series of heuristically-derived binary morphological
+%operations.  
+The two-stage approach is motivated by
 the fact that the RF is a powerful tool for ranking the features
-in a high-dimensional model i.e. can be used for model selection.
-However, the RF outputs are not geometrically constrained, which is
-the motivation for our second refinement step.
+in a high-dimensional model i.e., it can be used for model selection
+in voxel-wise classification.  However, the RF outputs are not geometrically 
+constrained, which is the motivation for our second refinement step.
 
@@
 we have made the code available as open source%
-\footnote{
-https://github.com/ntustison/BRATS2013
-}
+%\footnote{
+%https://github.com/ntustison/BRATS2013
+%}
 and, to our knowledge, we are the only competitors out of the approximately 20 teams 
@@
 on the well-vetted and open source Insight Toolkit (ITK)%
-\footnote{
-http://www.itk.org/
-} 
+%\footnote{
+%http://www.itk.org/
+%} 
 of the National Institutes of Health and is available through our \textit{ANTsR} 
@@
 
+Feature generation for the proposed framework is critical.  The performance
+of the selected machine learning strategy is limited by the input used for 
+training. The image features described below were selected based on previous
+work in the literature and our own observations regarding tumor characteristics
+and what could possibly be encoded algorithmically.  Observations 
+regarding tumor intensity differences from normal tissue has been encoded
+in previously proposed research by calculating normalized neighborhood 
+intensity statistics and intensity modeling \citep[e.g.,][]{bauer2011,bauer2012,geremia2012,zikic2012}.
+In these approaches both the features and classification algorithm are limited to 
+voxelwise considerations.
+
+Our success in the MICCAI 2013 challenge was largely due to the inclusion of
+an extremely important set of features based on the common observation of
+bilateral asymmetrization caused by tumor presence.  Additionally, several
+features were designed to encode spatial coherence even though the RF classification
+is performed on a voxelwise basis. For example, connected component geometry of the
+intensity modeling labels were used to discriminate observations of compact,
+isotropic objects (vs. highly anisotropic) typical of tumor presence and growth 
+\citep{greenspan1972}.  Table \ref{table:features} provides a listing of image
+features used and the motivation for their inclusion.
+
 Certain key elements characterize the proposed supervised brain tumor segmentation
@@
 \begin{itemize}
-  \item construction of symmetric multivariate templates for multimodal 
-        asymmetry-based features,
-  \item generation of other image-based features,
+  \item construction of symmetric multivariate templates,
+  \item generation of image features including those based on:
+    \begin{itemize}
+      \item asymmetry and template normalization, 
+      \item intensity modeling and connected component geometry,      
+      \item neighborhood first-order statistics, and
+      \item brain mask coordinate systems.
+    \end{itemize}
   \item training and geometric refinement of RF models, and
@@
 
-\subsection{Image Features for RF Model Training and Prediction}
-Feature generation is a critical step in machine learning regardless
-of which prediction methods are employed.  Below, we detail the
-feature extraction approach and how these results are fed into the
-random forest technology on which we based our final segmentations.
-
-\subsubsection{Multimodal Asymmetry Features from Multivariate Symmetric Templates}
+\begin{table}
+\caption{Image feature list for the proposed supervised brain tumor segmentation.}
+\label{table:features}
+\centering
+\begin{tabular*}{0.99\textwidth}{@{\extracolsep{\fill}} c c l}
+\toprule
+\multicolumn{3}{c}{\bf Asymmetry and template normalization} \\
+\midrule
+\multicolumn{1}{c}{Feature} & \multicolumn{1}{c}{Number} & \multicolumn{1}{c}{Motivation} \\
+\midrule
+contralateral difference & 1 per modality & contralateral comparison \\
+template difference  & 1 per modality & average normal comprison\\
+log Jacobian &  1 & tumor distortion \\
+warped template distance & 1 & tumor distortion \\
+\midrule
+\multicolumn{3}{c}{\bf Intensity modeling and connected component geometry} \\
+\midrule
+\multicolumn{1}{c}{Feature} & \multicolumn{1}{c}{Number} & \multicolumn{1}{c}{Motivation} \\
+\midrule
+Pr(CSF)  & 1 per modality & normal tissue intensity\\
+Pr(gray matter) & 1 per modality & normal tissue intensity\\
+Pr(white matter) & 1 per modality & normal tissue intensity\\
+Pr(edema) & 1 per modality & normal tissue intensity\\
+Pr(non-enhancing) & 1 per modality & tumor tissue intensity\\
+Pr(enhancing) & 1 per modality & tumor tissue intensity\\
+Pr(necrosis) & 1 per modality & tumor tissue intensity\\
+elongation & 1 per modality & anisotropic components \\
+eccentricity & 1 per modality & anisotropic components \\
+volume & 1 per modality & small, isolated components \\
+distance to tumor core & 1 per modality & proximity to tumor core \\
+$\frac{\mathrm{volume}}{\mathrm{surface\,\,area}}$ & 1 per modality & anisotropic components \\
+\midrule
+\multicolumn{3}{c}{\bf Neighborhood first-order statistics} \\
+\midrule
+\multicolumn{1}{c}{Feature} & \multicolumn{1}{c}{Number} & \multicolumn{1}{c}{Motivation} \\
+\midrule
+mean (radius = 1)  & 1 per modality &  tissue \& tumor intensity \\
+std. dev. (radius = 1)  & 1 per modality & tissue \& tumor  intensity \\
+skewness (radius = 1)  & 1 per modality & tissue \& tumor  intensity \\
+mean (radius = 3)  & 1 per modality & tissue \& tumor  intensity \\
+std. dev. (radius = 3)  & 1 per modality & tissue \& tumor  intensity \\
+skewness (radius = 3)  & 1 per modality & tissue \& tumor  intensity \\
+T1, T1C difference & 1 & isolate tumor tissue \\
+\midrule
+\multicolumn{3}{c}{\bf Brain mask coordinate system} \\
+\midrule
+\multicolumn{1}{c}{Feature} & \multicolumn{1}{c}{Number} & \multicolumn{1}{c}{Motivation} \\
+\midrule
+subject distance & 1 & peripheral vs. subcortical \\
+\bottomrule
+\end{tabular*}
+\end{table}
+
+%FeatureSet ,  Number of Features ,   Motivation  ,  Modalities involved
+%
+%Asymmetry ,       2                        ,    ....            ,
+
+\subsection{Optimal Symmetric Multivariate Templates}
 
@@
            subject images result in the asymmetric template represented at 
-           the top of the middle column.  A horizontal reflection 
-           perpendicular to the mid-sagittal
-           plane, $T_R$, resulted in the contralateral counterpart represented
+           the top of the middle column.  A horizontal reflection, $T_R$,
+           perpendicular to the mid-sagittal plane  resulted in the 
+           contralateral counterpart represented
            at the bottom of the second column.  The final template seen on
@@
 
-To take into account these potential asymmetries, we 
+To take into account these potential asymmetries and the distortion of 
+the normal plane of symmetry in the brain, we 
 require a data set with the same modalities as dictated
 by the subject image acquisition protocol.  Although it
-would be preferable to build population-specific multivariate
+is preferable to build population-specific multivariate
 templates from normal data using the same acquisition 
-parameters \citep{avants2010}, we substituted well-known,
-publicly available data since such normal data was not made available
-for the BRATS 2013 challenge. 
-A recent neuroimaging reproducibility study
-by Landman et al. resulted in an open data cohort of 21
+parameters \citep{avants2010}, such data were not available
+from the BRATS challenge.  Therefore, we substituted well-known,
+publicly available data from a recent neuroimaging reproducibility study
+by Landman et al. which resulted in an open data cohort of 21
 normal individuals, each imaged twice, comprising several
@@
 a multivariate template due to its public availability and
-inclusion of several modalities (even permitting future 
+inclusion of several modalities even permitting future 
 incorporation of modalities 
 not currently included with the BRATS challenge into our 
-segmentation framework).
+segmentation framework.
 
@@
   \caption{ Given the mapping between the template, $T$, and subject, $S$, domains ($\phi_b: S  \leftrightarrow \underset{b}{\leftrightsquigarrow} T$), various features can be calculated which demonstrate good discriminative qualities.
-Feature images used are outlined in a dashed line.
+$T_R$ denotes a horizontal reflection
+  perpendicular to the mid-sagittal plane.
+  Feature images used are specified by the dashed box.
 Top:  Difference images with the symmetric multivariate template are created by warping the template to the subject space and performing a voxelwise subtraction from the original modality image.
@@
 optimal multivariate template appearance 
-$\mathbf{J}=\{J_1,J_2,\ldots, J_K\}$ (with corresponding 
-coordinate system $\psi(\mathbf{x})$) to minimize the
+$\mathbf{J}=\{J_1,J_2,\ldots, J_K\}$, with corresponding 
+coordinate system $\psi(\mathbf{x})$, to minimize the
 following cost function:
@@
            &+ \sum_{k=1}^K \lambda_k \Pi_k \left(I_k^n,J_k   
-           \left(\phi^{-1}_n(\mathbf{x},1)\right)\right)\Bigg]
+           \left(\phi^{-1}_n(\mathbf{x},1)\right)\right)\Bigg].
 \end{align}
@@
 
+\subsection{Image Features Based on Asymmetry and Symmetric Template Normalization}
+
 After constructing the template offline, each data set is processed
@@
 which has demonstrated good performance in normal brain registration \citep{tustison2013a}.
-Good performance also extends to this pathological data scenario as 
-indicated by visual inspection and the fact that the derived features
-were amongst the most informative in our winning entry.
-
+%Good performance also extends to this pathological data scenario as 
+%indicated by visual inspection and the fact that the derived features
+%were amongst the most informative in our winning entry.
 We denote the mapping from the subject, $S$, to the template, $T$,
 space as $\phi_b: S \leftrightarrow \underset{b}{\leftrightsquigarrow}  T$
-which consists of both affine and diffeomorphic components.  Transform
+which consists of both affine and diffeomorphic components.  Note that transform
 invertibility is essential for the template-based features.  The first 
@@
 to the subject space and calculating the difference image (see the 
-top portion of Figure \ref{fig:asymmetryFeatures}).  For example,
+top portion of Figure \ref{fig:asymmetryFeatures}).  
+Note that the T1-weighted contrast and non-contrast images are paired with
+the T1-weighted component of the symmetric template.
+For example,
 the T2 symmetric template voxelwise difference image is calculated from
@@
 \end{align}
-Note that the T1-weighted contrast and non-contrast images are paired with
-the T1-weighted component of the symmetric template.  The second
+
+The second
 set of template-based feature images is generated per modality 
@@
 \end{align}
-This process is illustrated in the bottom portion of Figure \ref{fig:asymmetryFeatures}. A final, related feature is the Jacobian
-of the transform.  The basis for this final feature is that 
-relatively larger Jacobian values might be indicative of tumor
-expansion.
+The corresponding feature image is calculated as the voxelwise difference $S_{contralateral} - S$.
+This process is illustrated in the bottom portion of Figure \ref{fig:asymmetryFeatures}. Two 
+related features are the Jacobian determinant image 
+of the transform (calculated voxelwise) from the template to the subject, i.e., $J(\phi_b^{-1})$
+and the warped signed distance template mask \citep{maurer2003}.
+The motivation for inclusion of these two feature is that 
+relatively larger Jacobian values are a potential indicator of tumor
+expansion similar to the distorted distance map.
 
@@
   \caption{Representative feature images derived from the 
-           BRATS\_HG0301  ``challenge'' data set.
+           BRATS\_HG0301  final assessment data set.
            Neighborhood statistical images for each modality were 
            generated by calculating a given statistic within
-           a specified neighborhood radius.  Also calculated for each modality were feature
+           a specified neighborhood radius.  Similarly, we 
+           calculate the voxelwise difference image between the
+           T1 and T1C image.  
+           of 
+           Also calculated for each modality were feature
            images based on either the GMM or the MAP-MRF segmentation.  For the former, we
@@
            corresponding symmetric template itself and with respect to the 
-           contralateral side.   
-           This mapping is also used to produce the log Jacobian image.  Finally, the 
-           (T1 - TC) image is calculated and, from the
+           contralateral side.   This mapping is also used to produce the log Jacobian image
+           and warped template distance mask.  Additionally, the 
+           (T1 - T1C) image is calculated and, from the
            cerebral mask, we calculate the normalized distance image.  
@@
 
-\subsubsection{Image Features for Concatenated Random Forests}
+\subsection{Voxelwise Image Features for Random Forest Supervised Segmentation}
 
@@
 models, e.g. \cite{liaw2002}.
-The seven labels of brain/tumor tissue type denote:
+We then use these output tissue map estimates as spatial 
+priors for generating a second set of geometrically refined image features.  These
+are used as input for a second RF model application (or
+second stage),
+the output of which constitute the final tumor segmentation estimate.
+Note that some feature images are included directly and
+  without modification in both stages such as
+the asymmetry features described in the previous section.
+
+\subsection{Image Features Based on Intensity Modeling and Regional Geometry}
+
+Intensity modeling of the constituent tissue types has proven
+useful in previous tumor segmentation protocols.  For example,
+Gaussian mixture modeling (GMM) was used in \citep{bauer2012,zikic2012} to model
+tumor tissue components.  Similarly, we use GMM to model 
+the seven brain/tumor tissue types:
 \begin{itemize}
@@
 \end{itemize}
-We then use these output tissue map estimates as spatial 
-priors for generating a second set of geometrically refined image features.  These
-are used as input for a second RF model application (or
-second stage),
-the output of which constitute the final tumor segmentation estimate.
-Note that some feature images are included directly and
-  without modification in both stages such as
-the asymmetry features described in the previous section.
-
-\begin{figure}
-  \centering
-  \begin{tabular}{c}
-    \includegraphics[width=88mm]{gmm_map_mrf.pdf}
-  \end{tabular}
-  \caption{Visual comparison between the first and second RF stages
-  on the evaluation {\tt BRATS\_HG0001} data set.
-  Although the GMM intensity modeling does provide certain 
-  discriminative benefit (see Figures \ref{fig:hgimportance}
-  and \ref{fig:lgimportance}), the use of the RF-derived posteriors 
-  coupled with the described MAP-MRF segmentation framework enhances 
-  the accuracy of the intensity-based modeling features.    
-  }
-  \label{fig:stageComparison}
-\end{figure}
-
-%\textcolor{red}{Nick, I think the motivation for this is still unclear
-%  .... doesnt it have something to do with the MRF of atropos
-%  smoothing out the importance features resulting from the RF step?
-%  Figure 4 tells us a little bit about the procedure of the code but
-%  the idea is not conveyed ... would it be possible to show exactly
-%  what is happening in a separate figure?  e.g. just conceptually in a small 
-%  region of voxels with 2 modalities ....?}
-
-For both stages, in contrast to previous generative
+However, since GMM is performed without any spatial 
+considerations, we augment the initial supervised segmentation 
+classification result by employing a second classification round
+using an MAP-MRF prior on the same seven brain/tumor tissue types
+(detailed below).
+
+In contrast to previous generative
 modeling approaches for multimodal tumor segmentation 
@@
 segmentation method based on
-Gaussian mixture modeling (GMM).  GMM results in voxelwise
+GMM.  GMM results in voxelwise
 tissue probabilities which comprise an additional set of image
@@
 along with five additional geometric features based on the connected 
-components of the segmentation image (distance image to tumor core,
-physical volume image, volume to surface area ratio, eccentricity,
-and elongation).
+components of the segmentation image.  These are created by first isolating
+each of the seven tissue labels.  For each label we determine the connected
+components then, for each connected component, we calculate its physical
+volume and volume-to-surface ratio. Additionally, we calculate the component's 
+second-order central image moments assembled into the image covariance matrix 
+\citep{padfield2008}.  
+The ordered eigenvalues are denoted as $\{\lambda_1,\lambda_2,\lambda_3\}$ and 
+are used to calculate a 3-D estimate of the eccentricity,
+\begin{align}
+  eccentricity = \frac{\lambda_3 - \lambda_1}{\lambda_3},
+\end{align}
+and elongation,
+\begin{align}
+  elongation = \frac{\lambda_3}{\lambda_1},
+\end{align}
+meant to discriminate isotropic vs. anisotropic objects.
+Note that each measure applies to a single connected component but
+the value is assigned to each voxel comprising that labeled object.
+For each image, we also calculate the Euclidean distance
+to the tumor core (label 7).
+
 During the training phase (or RF model construction),
 these feature images, $\left\{F_i: i \in GMM\right\}$, generated during Stage 1 are used to create
-the Stage 1 RF model using the following relationship
+the Stage 1 RF model with the following relationship
 \begin{align}
 \label{eq:gmm}
- v_L \sim \sum_{i \in Asym} F_i + \sum_{i \in GMM} F_i + \sum_{i \in Misc} F_i
+ v_L \sim \sum_{i \in Asym} F_i + \sum_{i \in GMM} F_i + \sum_{i \in Neigh} F_i + \sum_{i \in Dist} F_i
 \end{align}
@@
 In other words, $\bm{F}= [ \bm{F}_{Asym}, \bm{F}_{GMM},
-\bm{F}_{Misc} ]$, where we column-wise concatenate the GMM, Asym and
-Misc feature matrices.
-The set of $\left\{F_i: i \in Asym\right\}$ are the feature images 
-generated from the symmetric multivariate template 
-described in the previous section and $\left\{F_i: i \in Misc\right\}$
-are miscellaneous feature images described in Section \ref{sec:misc}.
-Therefore, in total, the number of features is $k=84$ where there
-are twelve GMM features for each of the four modalities ($n=48$), 
-two asymmetry features for each modality ($n=8$), and six modality-specific 
-miscellaneous geometric features ($n=24$) with an additional four feature
-images.
-
-% 48 + 8 + 24 + 4 = 84
-
-\paragraph{Stage 2:  Refined spatial coherence classification}
-The first step in stage 2 is to use the per-class probabilities output
-by the stage 1 random forest model as input to the Atropos
-segmentation algorithm.  Application of the Stage 1 RF model to a set of feature images 
+\bm{F}_{Neigh}, \bm{F}_{Dist} ]$, where we column-wise concatenate the asymmetry ($Asym$),
+intensity modeling and connected component geometry ($GMM$), neighborhood statistics ($Neigh$),
+ and brain mask signed distances ($Dist$) feature matrices.
+
+Application of the Stage 1 RF model to a set of feature images 
 from an individual subject results in a set of $n$ spatial images
@@
 classification probabilities are derived by normalizing 
-these votes to $[0,1]$ as in \cite{liaw2002}.
-
-These spatial probability maps are passed as spatial priors to the
+these votes to $[0,1]$ \citep{liaw2002}.
+
+%Therefore, in total, the number of features is $k=84$ where there
+%are twelve GMM features (7 posteriors plus 5 geometric features) for each 
+%of the four modalities ($n=48$), 
+%two asymmetry features for each modality ($n=8$), and six modality-specific 
+%miscellaneous geometric features ($n=24$) with an additional four feature
+%images.
+
+% 48 + 8 + 24 + 4 = 84
+
+\paragraph{Stage 2:  Refined spatial coherence classification}
+Intialization of stage 2 employs the per-class probabilities output
+by the stage 1 RF model as input to an expectation-maximization MAP-MRF 
+segmentation algorithm.  
+ Specifically, these spatial probability maps are passed as priors to the
 {\tt antsAtroposN4.sh} script in ANTs which couples the Atropos 
@@
 The posterior probabilities that are output by Atropos and the
-connected-component feature images for each modality (described in
-section~\ref{sec:misc}) are then passed to the second stage random
-forest model, along with the same asymmetry feature set based on our
-symmetric templates.  
-The second stage RF model is then produced similarly as the previous stage: 
+connected-component feature images for each modality  are denoted as $\bm{F}_{MRF}$.    
+The second stage RF model is then produced similarly as the previous stage
+using the concatenated feature set $\bm{F}= [ \bm{F}_{Asym}, \bm{F}_{MRF},
+\bm{F}_{Neigh}, \bm{F}_{Dist} ]$: 
 \begin{align}
 \label{eq:mapmrf}
-\mathrm{Tissue}\,\,\mathrm{classes} \sim \sum_{i \in MAP\_MRF} F_i + \sum_{i \in Asym} F_i + \sum_{i \in Misc} F_i,
+ v_L \sim \sum_{i \in Asym} F_i + \sum_{i \in MRF} F_i + \sum_{i \in Neigh} F_i + \sum_{i \in Dist} F_i.
 \end{align}
-where $MAP-MRF$ represents the Stage 2 application of Atropos and the
-other features are similar to those used in Stage 1.
 The use of these MRF features improves the accuracy of the 
@@
 the initial estimation stage (which lacked the MRF spatial prior).  
-
-\subsubsection{Miscellaneous Multimodal Feature Image Generation}
-\label{sec:misc}
-Our strategy for feature selection was to, first, generate an ample feature
-set and, second, to use the training phase to prioritize the features
-appropriately.  In principal, this will eliminate weak features and
-retain strong ones even in the presence of substantial noise.  Given
-that image feature shape may be a valuable predictor of the presence of a
-tumor \cite{zacharaki2009classification}, we calculate, for each modality, several images of neighborhood 
-first-order statistics including mean, variance, skewness, and
-entropy.  Block neighborhood radii of values $\{1,2,3\}$ were tested.
-Due to overlapping discriminative qualities, we narrowed the feature
-values to 1 and 3 voxels.  We also calculated two Euclidean distance
-transforms \citep{maurer2003} to be used for each subject.  One was
-calculated from the subject's own cerebral mask and the other was the 
-distance transform of the symmetric template cerebral mask warped
-to the subject.  Finally, we also generated the  (T1 - T1C) difference 
-image \citep{prastawa2003}.
-The script {\tt createFeatureImages.sh} performs all preprocessing
-and generates all feature images.  In Figure \ref{fig:featureImages} we provide sample mid-axial 
-image slices from features generated from the {\tt BRATS\_HG0301} data
-set.  Note that these feature images do not require the presence of
-any training data.
-
-\subsection{Overview}
-
-In summary, the proposed RF classification approach per subject consists of the 
-following steps:
-
-\begin{enumerate}
-  \item Calculate Stage 1 features based on:  
-  \begin{itemize}
-    \item GMM intensity modeling,
-    \item asymmetry, and
-    \item miscellaneous.
-  \end{itemize}
-  \item Apply Stage 1 RF model to the above calculated features resulting in seven 
-        probability maps for the seven tissue types.  
-  \item Use the probability maps to calculate Stage 2 features based on MAP-MRF intensity modeling.
-  \item Apply Stage 2 RF model to MAP-MRF, asymmetry, and miscellaneous features to 
-  get final RF classification. 
-\end{enumerate}
+In total, the number of features is for this final stage is the same
+where the GMM features are replaced by their MRF counterparts.
+The rest of the feature images from the first stage are also included unaltered for this second stage.
+
+%\subsubsection{Miscellaneous Multimodal Feature Image Generation}
+%\label{sec:misc}
+%Our strategy for feature selection was to, first, generate an ample feature
+%set and, second, to use the training phase to prioritize the features
+%appropriately.  In principle, this will eliminate weak features and
+%retain strong ones even in the presence of substantial noise.  Given
+%that image feature shape may be a valuable predictor of the presence of a
+%tumor \cite{zacharaki2009classification}, we calculate, for each modality, several images of neighborhood 
+%first-order statistics including mean, variance, skewness, and
+%entropy.  Block neighborhood radii of values $\{1,2,3\}$ were tested.
+%Due to overlapping discriminative qualities, we narrowed the feature
+%values to 1 and 3 voxels.  We also calculated two Euclidean distance
+%transforms \citep{maurer2003} to be used for each subject.  One was
+%calculated from the subject's own cerebral mask and the other was the 
+%distance transform of the symmetric template cerebral mask warped
+%to the subject.  Finally, we also generated the  (T1 - T1C) difference 
+%image \citep{prastawa2003}.
+%The script {\tt createFeatureImages.sh} performs all preprocessing
+%and generates all feature images.  In Figure \ref{fig:featureImages} we provide sample mid-axial 
+%image slices from features generated from the {\tt BRATS\_HG0301} data
+%set.  Note that these feature images do not require the presence of
+%any training data.
+
+\begin{figure}
+  \centering
+  \begin{tabular}{c}
+    \includegraphics[width=88mm]{gmm_map_mrf.pdf}
+  \end{tabular}
+  \caption{Visual comparison between the first and second RF stages
+  on the evaluation {\tt BRATS\_HG0001} data set.
+  Although the GMM intensity modeling does provide certain 
+  discriminative benefit (see Figures \ref{fig:hgimportance}
+  and \ref{fig:lgimportance}), the use of the RF-derived posteriors 
+  coupled with the described MAP-MRF segmentation framework enhances 
+  the accuracy of the intensity-based modeling features.    
+  }
+  \label{fig:stageComparison}
+\end{figure}
+
+%\textcolor{red}{Nick, I think the motivation for this is still unclear
+%  .... doesnt it have something to do with the MRF of atropos
+%  smoothing out the importance features resulting from the RF step?
+%  Figure 4 tells us a little bit about the procedure of the code but
+%  the idea is not conveyed ... would it be possible to show exactly
+%  what is happening in a separate figure?  e.g. just conceptually in a small 
+%  region of voxels with 2 modalities ....?}
+
+%\subsection{Overview}
+%
+%In summary, the proposed RF classification approach per subject consists of the 
+%following steps:
+%
+%\begin{enumerate}
+%  \item Calculate Stage 1 features based on:  
+%  \begin{itemize}
+%    \item GMM intensity modeling,
+%    \item asymmetry, and
+%    \item miscellaneous.
+%  \end{itemize}
+%  \item Apply Stage 1 RF model to the above calculated features resulting in seven 
+%        probability maps for the seven tissue types.  
+%  \item Use the probability maps to calculate Stage 2 features based on MAP-MRF intensity modeling.
+%  \item Apply Stage 2 RF model to MAP-MRF, asymmetry, and miscellaneous features to 
+%  get final RF classification. 
+%\end{enumerate}
 
@@
 
-\subsection{\textit{ANTsR}:  An ANTs/R Interface}
+\subsection{\textit{ANTsR}:  An ANTs/R Interface for Random Forest Training and Prediction}
 
@@
 numerical and visualization software packages.
-The well-known neuroimaging package SPM%
-\footnote{
-http://www.fil.ion.ucl.ac.uk/spm/
-}
+The well-known neuroimaging package SPM
+%\footnote{
+%http://www.fil.ion.ucl.ac.uk/spm/
+%}
 is a significant extension of the commercial computing and visualization environment
-Matlab.  Open source neuroimaging packages, such as NIPY (neuroimaging in Python),%
-\footnote{
-http://nipy.org
-} 
+Matlab.  Open source neuroimaging packages, such as NIPY (neuroimaging in Python),
+%\footnote{
+%http://nipy.org
+%} 
 rely on other open source packages for numerical/statistical analysis.  NIPY,
@@
 optimization.%
-\footnote{
-http://www.numpy.org
-}
+%\footnote{
+%http://www.numpy.org
+%}
 
@@
 transformations, and the \textit{R} project%
-\footnote{
-http://www.r-project.org
-}
+%\footnote{
+%http://www.r-project.org
+%}
 for statistical computing and visualization thus providing a complete
@@
 \verb#rgl#, and \verb#misc3d#.%  
-\footnote{
-See \href{http://stnava.github.io/software/2014/01/08/antsr/}{http://stnava.github.io/software/2014/01/08/antsr/} for current status.}
+%\footnote{
+%See \href{http://stnava.github.io/software/2014/01/08/antsr/}{http://stnava.github.io/software/2014/01/08/antsr/} for current status.}
 Additionally, in order
@@
 and \verb#ggplot2#.%
-\footnote{
-Packages are easily installed using the {\tt install.packages()} \textit{R} mechanism.
-}
+%\footnote{
+%Packages are easily installed using the {\tt install.packages()} \textit{R} mechanism.
+%}
 
@@
 required.  CMake%
-\footnote{
-http://www.cmake.org/
-}
+%\footnote{
+%http://www.cmake.org/
+%}
 is an open source tool for the management and building of 
@@
 such as the Insight Toolkit (ITK)%
-\footnote{
-http://www.itk.org/
-}
+%\footnote{
+%http://www.itk.org/
+%}
 and ANTs.  Further instructions for download and
@@
 operating systems.%  
-\footnote{Windows installation should be possible
-but, to our knowledge, has not been attempted.}
+%\footnote{Windows installation should be possible
+%but, to our knowledge, has not been attempted.}
 
@@
 both testing and training on sample challenge data.%
-\footnote{
-https://github.com/ntustison/ANTsAndArboles/tree/master/SimpleExample
-}
+%\footnote{
+%https://github.com/ntustison/ANTsAndArboles/tree/master/SimpleExample
+%}
 After pulling the repository, one can run the scripts {\tt exampleTrain.sh}
@@
 Challenge on Multimodal Brain Tumor Segmentation%
-\footnote{
-http://martinos.org/qtim/miccai2013/index.html
-}
+%\footnote{
+%http://martinos.org/qtim/miccai2013/index.html
+%}
 organized by K. Farahani, M. Reyes, B. Menze, E. Gerstner, J. Kirby and J. Kalpathy-Cramer. 
@@
 protocol.%
-\footnote{
-http://martinos.org/qtim/miccai2013/data.html
-}
+%\footnote{
+%http://martinos.org/qtim/miccai2013/data.html
+%}
 The labeling of pathology was categorized into four regions:
@@
 earlier, we provide the \textit{R} script {\tt plotVariableImportance.R} to render
-one such quantity denoted as {\tt MeanDecreaseAccuracy}.  During model construction
+one such quantity use to assess RF feature importance denoted as {\tt MeanDecreaseAccuracy}.  
+During model construction
 (specifically the out-of-bag error calculation stage), the decrease in prediction accuracy
@@
 to be the most discriminative.%
-\footnote{
-There is an issue with consistently ranking correlated predictors as described at {\tt http://www.r-bloggers.com/random-forest-variable-importance/} since the permutation testing performed for predictive accuracy assessment assumes predictor independence.
+%\footnote{
+%There is an issue with consistently ranking correlated predictors as described at {\tt http://www.r-bloggers.com/random-forest-variable-importance/} since the permutation testing performed for predictive accuracy assessment assumes predictor independence.
 Correctives have been proposed but we ignore these issues for this particular application.
-}
+%}
 In this work, we do not use these measurements for feature pruning.  However,
@@
 processing on the computational cluster at the University of Virginia.%
-\footnote{
-http://www.uvacse.edu
-}
+%\footnote{
+%http://www.uvacse.edu
+%}
 
@@
 
-Although these competitions are extremely useful as they provide objective
-assessment of algorithmic performance given the prevalence of selection bias
-in reporting results in conventional publication venues, even more important
-is the availing of the actual algorithmic instantiation (i.e. code) so that
+Objective assessment  of algorithmic performance (given the prevalence of selection bias
+in reporting results in conventional publication venues) is certainly a motivating
+factor for these competitions. Equally, if not more, important
+is the availability of the actual algorithmic instantiation (i.e. code) so that
 others can more easily build upon and utilize what has proven effective. 

### 2014-05-03 paper.tex
ENH:  Getting ready to submit.

@@
 In line with our philosophy regarding open science \citep{tustison2013,ince2012}, 
-we have made the code available as open source%
+we have made the code available as open source
 %\footnote{
@@
 that participated in either years' challenges to do this.  All code is based
-on the well-vetted and open source Insight Toolkit (ITK)%
+on the well-vetted and open source Insight Toolkit (ITK)
 %\footnote{
@@
 for example, uses the more generic packages NumPy and SciPy for numerical analysis and 
-optimization.%
+optimization.
 %\footnote{
@@
 powerful neuroimaging toolkit for producing reliable imaging data 
-transformations, and the \textit{R} project%
+transformations, and the \textit{R} project
 %\footnote{
@@
 \verb#mFilter#, \verb#doParallel#, \verb#robust#, \verb#magic#, \verb#knitr#, \verb#pixmap#, 
-\verb#rgl#, and \verb#misc3d#.%  
+\verb#rgl#, and \verb#misc3d#.
 %\footnote{
@@
 \verb#randomForest#, \verb#snowfall#, \verb#rlecuyer#,
-and \verb#ggplot2#.%
+and \verb#ggplot2#.
 %\footnote{
@@
 In addition to \textit{R} and the add-on packages previously mentioned, CMake is also 
-required.  CMake%
+required.  CMake
 %\footnote{
@@
 to coordinate the downloading of external packages,
-such as the Insight Toolkit (ITK)%
+such as the Insight Toolkit (ITK)
 %\footnote{
@@
 that \textit{ANTsR} is currently only tested on UNIX-alikes such OSX and Ubuntu
-operating systems.%  
+operating systems.
 %\footnote{Windows installation should be possible
@@
 We also include a fully functional 2-D example which performs
-both testing and training on sample challenge data.%
+both testing and training on sample challenge data.
 %\footnote{
@@
 Brain tumor image data used in this work were obtained from the NCI-MICCAI 2013 
-Challenge on Multimodal Brain Tumor Segmentation%
+Challenge on Multimodal Brain Tumor Segmentation
 %\footnote{
@@
 Manual labeling was performed in the axial plane following a detailed
-protocol.%
+protocol.
 %\footnote{
@@
 those features which have the greatest decrease in mean accuracy are considered
-to be the most discriminative.%
+to be the most discriminative.
 %\footnote{
@@
 {\tt applyTumorSegmentationModelForCohort.pl} designed for parallel subject 
-processing on the computational cluster at the University of Virginia.%
+processing on the computational cluster at the University of Virginia.
 %\footnote{

### 2014-05-10 Manuscript/template.tex
ENH:  Submit to Neuroinformatics.

--- 2014-05-03 paper.tex
@@
-%% This is file `elsarticle-template-1-num.tex',
-%%
-%% Copyright 2009 Elsevier Ltd
-%%
-%% This file is part of the 'Elsarticle Bundle'.
-%% ---------------------------------------------
-%%
-%% It may be distributed under the conditions of the LaTeX Project Public
-%% License, either version 1.2 of this license or (at your option) any
-%% later version.  The latest version of this license is in
-%%    http://www.latex-project.org/lppl.txt
-%% and version 1.2 or later is part of all distributions of LaTeX
-%% version 1999/12/01 or later.
-%%
-%% The list of all files belonging to the 'Elsarticle Bundle' is
-%% given in the file `manifest.txt'.
-%%
-%% Template article for Elsevier's document class `elsarticle'
-%% with numbered style bibliographic references
-%%
-%% $Id: elsarticle-template-1-num.tex 149 2009-10-08 05:01:15Z rishi $
-%% $URL: http://lenova.river-valley.com/svn/elsbst/trunk/elsarticle-template-1-num.tex $
-%%
-
-\documentclass[preprint,authoryear,review,12pt]{elsarticle}
-%\documentclass[final,5p,times,twocolumn]{elsarticle}
-
-%% Use the option review to obtain double line spacing
-%% \documentclass[preprint,review,12pt]{elsarticle}
-
-%% Use the options 1p,two column; 3p; 3p,twocolumn; 5p; or 5p,twocolumn
-%% for a journal layout:
-%% \documentclass[final,1p,times]{elsarticle}
-%% \documentclass[final,1p,times,twocolumn]{elsarticle}
-%% \documentclass[final,3p,times]{elsarticle}
-%% \documentclass[final,3p,times,twocolumn]{elsarticle}
-%% \documentclass[final,5p,times]{elsarticle}
-%% \documentclass[final,5p,times,twocolumn]{elsarticle}
-
-\usepackage{color,hyperref,bm}
-\usepackage{multirow,booktabs,ctable,array}
-\usepackage{lscape}
-\usepackage{amsmath}
-\usepackage{lineno}
-\usepackage{ulem}
-\usepackage{setspace}
-\usepackage{listings}
-\usepackage{float}
-\usepackage{listings}
-\usepackage{color,colortbl}
-\usepackage{rccol}
-\usepackage{pbox}
-%\usepackage[table]{xcolor}
-
-    \definecolor{listcomment}{rgb}{0.0,0.5,0.0}
-    \definecolor{listkeyword}{rgb}{0.0,0.0,0.5}
-    \definecolor{listnumbers}{gray}{0.65}
-    \definecolor{listlightgray}{gray}{0.955}
-    \definecolor{listwhite}{gray}{1.0}
-    \definecolor{lightcyan}{rgb}{0.88,1,1}
-
-\newcommand{\lstsetcpplong}
-{
-\lstset{frame = tb,
-        framerule = 0.25pt,
-        float,
-        fontadjust,
-        backgroundcolor={\color{listlightgray}},
-        basicstyle = {\ttfamily\scriptsize},
-        keywordstyle = {\ttfamily\color{listkeyword}\textbf},
-        identifierstyle = {\ttfamily},
-        commentstyle = {\ttfamily\color{listcomment}\textit},
-        stringstyle = {\ttfamily},
-        showstringspaces = false,
-        showtabs = false,
-        numbers = none,
-        numbersep = 6pt,
-        numberstyle={\ttfamily\color{listnumbers}},
-        tabsize = 2,
-        language=,
-        floatplacement=!h,
-        caption={\small \baselineskip 12pt DiReCT long command line menu which is invoked using the `{\ttfamily {-}{-}help}' option.
-The short command line menu is obtained by typing `{\ttfamily {-}h}'},
-        captionpos=b,
-        label=listing:long
-        }
-}
-
-\floatstyle{plain}
-\newfloat{command}{thp}{lop}
-\floatname{command}{Command}
-
-%\usepackage[nomarkers,notablist]{endfloat}
-
-%% if you use PostScript figures in your article
-%% use the graphics package for simple commands
-%% \usepackage{graphics}
-%% or use the graphicx package for more complicated commands
-%% \usepackage{graphicx}
-%% or use the epsfig package if you prefer to use the old commands
-%% \usepackage{epsfig}
-
-%% The amssymb package provides various useful mathematical symbols
-\usepackage{amssymb}
-%% The amsthm package provides extended theorem environments
-% \usepackage{amsthm}
-
- \usepackage{makecell}
-
-%% The lineno packages adds line numbers. Start line numbering with
-%% \begin{linenumbers}, end it with \end{linenumbers}. Or switch it on
-%% for the whole article with \linenumbers after \end{frontmatter}.
-%% \usepackage{lineno}
-
-%% natbib.sty is loaded by default. However, natbib options can be
-%% provided with \biboptions{...} command. Following options are
-%% valid:
-
-%%   round  -  round parentheses are used (default)
-%%   square -  square brackets are used   [option]
-%%   curly  -  curly braces are used      {option}
-%%   angle  -  angle brackets are used    <option>
-%%   semicolon  -  multiple citations separated by semi-colon
-%%   colon  - same as semicolon, an earlier confusion
-%%   comma  -  separated by comma
-%%   numbers-  selects numerical citations
-%%   super  -  numerical citations as superscripts
-%%   sort   -  sorts multiple citations according to order in ref. list
-%%   sort&compress   -  like sort, but also compresses numerical citations
-%%   compress - compresses without sorting
-%%
-%% \biboptions{comma,round}
-
-% \biboptions{}
-
-\providecommand{\OO}[1]{\operatorname{O}\bigl(#1\bigr)}
+%%%%%%%%%%%%%%%%%%%%%%% file template.tex %%%%%%%%%%%%%%%%%%%%%%%%%
+%
+% This is a general template file for the LaTeX package SVJour3
+% for Springer journals.          Springer Heidelberg 2010/09/16
+%
+% Copy it to a new file with a new name and use it as the basis
+% for your article. Delete % signs as needed.
+%
+% This template includes a few options for different layouts and
+% content for various journals. Please consult a previous issue of
+% your journal as needed.
+%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%
+% First comes an example EPS file -- just ignore it and
+% proceed on the \documentclass line
+% your LaTeX will extract the file if required
+\begin{filecontents*}{example.eps}
+%!PS-Adobe-3.0 EPSF-3.0
+%%BoundingBox: 19 19 221 221
+%%CreationDate: Mon Sep 29 1997
+%%Creator: programmed by hand (JK)
+%%EndComments
+gsave
+newpath
+  20 20 moveto
+  20 220 lineto
+  220 220 lineto
+  220 20 lineto
+closepath
+2 setlinewidth
+gsave
+  .4 setgray fill
+grestore
+stroke
+grestore
+\end{filecontents*}
+%
+\RequirePackage{fix-cm}
+%
+%\documentclass{svjour3}                     % onecolumn (standard format)
+%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
+\documentclass[smallextended]{svjour3}       % onecolumn (second format)
+%\documentclass[twocolumn]{svjour3}          % twocolumn
+%
+\smartqed  % flush right qed marks, e.g. at end of proof
+%
+\usepackage{graphicx}
 
@@
 
-\long\def\symbolfootnote[#1]#2{\begingroup%
-\def\thefootnote{\fnsymbol{footnote}}\footnote[#1]{#2}\endgroup}
-
-\journal{NeuroImage}
-
+%
+% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
+%
+% insert here the call for the packages your document requires
+%\usepackage{latexsym}
+
+\usepackage{color,bm,natbib}
+\usepackage[colorlinks=true,breaklinks]{hyperref}
+\usepackage{multirow,booktabs,ctable,array}
+\usepackage{amsmath}
+\usepackage{amssymb}
+% etc.
+%
+% please place your own definitions here and don't use \def but
+% \newcommand{}{}
+%
+% Insert the name of "your journal" with
+% \journalname{myjournal}
+%
 \begin{document}
 
-\begin{frontmatter}
-
-\title{Optimal Symmetric Multimodal Templates and Concatenated Random Forests for Supervised Brain Tumor Segmentation (Simplified) with \textit{ANTsR}}
-
-\author[label1]{Nicholas J.~Tustison
-  \fnref{label0}}
-  \fntext[label0]{\scriptsize Corresponding author:  PO Box 801339, Charlottesville, VA 22908; T:  434-924-7730; email address:  ntustison@virginia.edu.
-  }
-\author[label2]{K.~L.~Shrinidhi}
-\author[label1]{Max Wintermark}
-\author[label1]{Christopher R.~Durst}
-\author[label2]{Benjamin M.~Kandel}
-\author[label2]{James C.~Gee}
-\author[label3]{Murray C.~Grossman}
-\author[label2]{Brian B.~Avants}
-\address[label1]{Department of Radiology and Medical Imaging, University of Virginia, Charlottesville, VA}
-\address[label2]{Penn Image Computing and Science Laboratory, 
-                 Department of Radiology, University of Pennsylvania,
-                Philadelphia, PA}
-\address[label3]{Department of Neurology, University of Pennsylvania,
-                Philadelphia, PA}
-
-%\maketitle
+\title{Optimal Symmetric Multimodal Templates and Concatenated Random Forests for Supervised Brain Tumor Segmentation (Simplified) with \textit{ANTsR}%\thanks{Grants or other notes
+%about the article that should go on the front page should be
+%placed here. General acknowledgments should be placed at the end of the article.}
+}
+%\subtitle{Do you have a subtitle?\\ If so, write it here}
+
+%\titlerunning{Short form of title}        % if too long for running head
+
+\author{Nicholas J.~Tustison \and 
+        K.~L.~Shrinidhi \and
+        Max Wintermark \and
+        Christopher R.~Durst \and
+        Benjamin M.~Kandel \and
+        James C.~Gee \and
+        Murray C.~Grossman \and
+        Brian B.~Avants
+}
+
+%\authorrunning{Short form of author list} % if too long for running head
+
+\institute{N. Tustison \at
+              Department of Radiology and Medical Imaging, University of Virginia, Charlottesville, VA \\
+%              Tel.: +123-45-678910\\
+%              Fax: +123-45-678910\\
+              \email{ntustison@virginia.edu}           %  \\
+%             \emph{Present address:} of F. Author  %  if needed
+              \and
+              K. Shrinidhi, B. Kandel, J. Gee, and M. Grossman \at
+              Penn Image Computing and Science Laboratory, Department of Radiology, University of Pennsylvania, Philadelphia, PA
+              \and
+              M. Wintermark, C. Durst \at
+              Department of Radiology and Medical Imaging, University of Virginia, Charlottesville, VA \\
+              \and
+}
+
+\date{Received: date / Accepted: date}
+% The correct dates will be entered by the editor
+
+\maketitle
 
@@
 %available.
+\keywords{advanced normalization tools \and BRATS \and glioma \and \textit{R} project
+}
+% \PACS{PACS code1 \and PACS code2 \and more}
+% \subclass{MSC code1 \and MSC code2 \and more}
 \end{abstract}
 
-\begin{keyword}
-advanced normalization tools \sep BRATS \sep brain tumor segmentation 
-\sep \textit{R} project for statistical computing
-%% keywords here, in the form: keyword \sep keyword
-\end{keyword}
-
-\end{frontmatter}
-%\linenumbers
-
-%
-%
-\newpage
-
-%% MSC codes here, in the form: \MSC code \sep code
-%% or \MSC[2008] code \sep code (2000 is the default)
-
-%%
-%% Start line numbering here if you want
-%%
-% \linenumbers
-
-%% The Appendices part is started with the command \appendix;
-%% appendix sections are then done as normal sections
-%% \appendix
-
-%% \section{}
-%% \label{}
-
-%% References
-%%
-%% Following citation commands can be used in the body text:
-%% Usage of \cite is as follows:
-%%   \citep{key}          ==>>  [#]
-%%   \cite[chap. 2]{key} ==>>  [#, chap. 2]
-%%   \citet{key}         ==>>  Author [#]
-
-%% main text
-
-%\input{intro}
-%
-%\input{methods}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-% Introduction
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 \section{Introduction}
+
 Given the complexity of tumor growth and appearance and the need
@@
 %}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-% Materials and Methods
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
   \begin{tabular}{cccc}
-  \includegraphics[width=30mm]{BRATS_HG0001/BRATS_HG0001_FLAIR_slice93.png} &
-  \includegraphics[width=30mm]{BRATS_HG0001/BRATS_HG0001_T1_slice93.png} &
-  \includegraphics[width=30mm]{BRATS_HG0001/BRATS_HG0001_T1C_slice93.png} &
-  \includegraphics[width=30mm]{BRATS_HG0001/BRATS_HG0001_T2_slice93.png} \\
-%  \includegraphics[width=20mm]{BRATS_HG0001/BRATS_HG0001_FLAIR_slice93.png} &
-%  \includegraphics[width=20mm]{BRATS_HG0001/BRATS_HG0001_T1_slice93.png} &
-%  \includegraphics[width=20mm]{BRATS_HG0001/BRATS_HG0001_T1C_slice93.png} &
-%  \includegraphics[width=20mm]{BRATS_HG0001/BRATS_HG0001_T2_slice93.png} \\
+  \includegraphics[width=28mm]{Fig1a.eps} &
+  \includegraphics[width=28mm]{Fig1b.eps} &
+  \includegraphics[width=28mm]{Fig1c.eps} &
+  \includegraphics[width=28mm]{Fig1d.eps} \\
   (a) & (b) &
@@
 \begin{figure*}
-    \centerline{\includegraphics[width=130mm]{templateKirby.pdf}}
+    \centerline{\includegraphics[width=120mm]{Fig2.eps}}
   \caption{Multivariate symmetric template created from the MMRR 
@@
   \centering
-%    \includegraphics[width=80mm]{asymmetryFeatureImages.pdf}
-    \includegraphics[width=120mm]{asymmetryFeatureImages.pdf}
+    \includegraphics[width=100mm]{Fig3.eps}
   \caption{ Given the mapping between the template, $T$, and subject, $S$, domains ($\phi_b: S  \leftrightarrow \underset{b}{\leftrightsquigarrow} T$), various features can be calculated which demonstrate good discriminative qualities.
@@
   \makebox[0.9\textwidth][c]{
-  \includegraphics[width=150mm]{featureImages.pdf}
+  \includegraphics[width=120mm]{Fig4.eps}
   }
@@
   \makebox[0.9\textwidth][c]{
-    \includegraphics[width=150mm]{pipelineTestingTraining.pdf}
+    \includegraphics[width=150mm]{Fig5.eps}
     }
@@
   \begin{tabular}{c}
-    \includegraphics[width=88mm]{gmm_map_mrf.pdf}
+    \includegraphics[width=88mm]{Fig6.eps}
   \end{tabular}
@@
 earlier, we provide the \textit{R} script {\tt plotVariableImportance.R} to render
-one such quantity use to assess RF feature importance denoted as {\tt MeanDecreaseAccuracy}.  
+one such quantity use to assess RF importance using {\tt MeanDecreaseAccuracy}.  
 During model construction
@@
   \makebox[\textwidth][c]{
-  \includegraphics[width=140mm]{challengeResults.pdf}
+  \includegraphics[width=140mm]{Fig7.eps}
   }
@@
 \bottomrule
-\multicolumn{5}{l}{\scriptsize$^\dagger$Indicates performance
-for all measures over all regions.  Component rankings are 
-given in parentheses.
+\multicolumn{5}{l}{\scriptsize$^\dagger$Indicates total performance
+ over all regions.  Component rankings are 
+in parentheses.
 }
@@
   \begin{tabular}{cc}
-  \includegraphics[width=90mm]{BRATS_HG_GMM.pdf} &
-  \includegraphics[width=90mm]{BRATS_HG_MAP_MRF.pdf} \\
+  \includegraphics[width=80mm]{Fig8a.eps} &
+  \includegraphics[width=80mm]{Fig8b.eps} \\
   \end{tabular}
@@
   \begin{tabular}{cc}
-  \includegraphics[width=90mm]{BRATS_LG_GMM.pdf} &
-  \includegraphics[width=90mm]{BRATS_LG_MAP_MRF.pdf} \\
+  \includegraphics[width=80mm]{Fig9a.eps} &
+  \includegraphics[width=80mm]{Fig9b.eps} \\
   \end{tabular}
@@
 
-%% References with bibTeX database:
-
-%\section*{Acknowledgments}
-%\clearpage
-\section*{References}
-
-\bibliographystyle{elsarticle-harv}
+\section*{Disclosures}
+The authors declare that they have no conflict of interest.
+
+%\begin{acknowledgements}
+%If you'd like to thank anyone, place your comments here
+%and remove the percent signs.
+%\end{acknowledgements}
+
+% BibTeX users please use one of
+%\bibliographystyle{spmpsci}      % mathematics and physical sciences
+%\bibliographystyle{spphys}       % APS-like style for physics
+%\bibliography{}   % name your BibTeX data base
+
+\bibliographystyle{spbasic}      % basic style, author-year citations
 \bibliography{references}
+
 \end{document}
+% end of file template.tex

### 2014-09-08 Manuscript/template.tex
ENH:  Minor edit.
ENH:  running title.

@@
 
-%\titlerunning{Short form of title}        % if too long for running head
+\titlerunning{Templates, Random Forests, and ANTsR for Brain Tumor Segmentation}        % if too long for running head
 
@@
 free to contact the authors if installation trouble occurs.  We note
-that \textit{ANTsR} is currently only tested on UNIX-alikes such OSX and Ubuntu
+that \textit{ANTsR} is currently only tested on OSX and certain Linux
 operating systems.
@@
 
+\section*{Information Sharing Statement}
+
+The open source ANTsR package is available as compilable source
+code or as a set of binaries from its dedicated github
+repository (https://github.com/stnava/ANTsR).  
+ANTsR is compatible with R versions 3 and above.
+Directions for compilation can be found in the README.md file associated 
+with the repository.  Installation from source assumes one has a compiler 
+and build environment that can be managed with
+the cross platform build tool CMake which is available at http://www.cmake.org.  
+The associated scripts written specifically
+for the BRATS 2013 challenge are found at https://github.com/ntustison/BRATS2013.
+A simple 2-D example is also provided at this location illustrating
+both training an RF model and prediction using the previously 
+described framework.
+
 %\begin{acknowledgements}
